---
title: "Comparison of Results from Zymo D6300; D6331 and D6323 FR : V1-V9 EMU on EMU
  database"
author: "Balázs Kakuk"
date: "2025-01-09"
output:
  html_document:
    code_folding: hide
    theme: cosmo
    toc: true
    toc_float:
      collapsed: false
  pdf_document:
    toc: true
---

```{css, echo=FALSE}
    body .main-container {
      max-width: 90% !important;
      width: 90% !important;
    }
    body {
      max-width: 90% !important;
      margin-left: auto;
      margin-right: auto;
    }
```

# Start Workflow

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, fig.align = 'center', fig.height = 10, fig.width = 30, warning = F, message = F, error = F)
library(tidyverse)
library(ggpubr)
library(factoextra)
library(FactoMineR)
library(ggrepel)
library(phyloseq, quietly = T)
library(DESeq2)
library(data.table)
library(ggsci)
library(hrbrthemes)
library(ggh4x)
library(RColorBrewer)
library(stringr); library(stringi)
library("ape")
library("ggtree")
library(Biostrings)
library(dplyr)
library(UpSetR)
library(ComplexHeatmap)
library(genefilter)
library(viridis)
library(knitr)
library(Maaslin2)
library(vegan)


## Own functions
misc.dir    <- 'C:/GitHub/Rlyeh/R'
minitax.dir <- 'C:/GitHub/minitax/R'

if (.Platform$OS.type!="windows") {
  
  misc.dir  <- paste0('/mnt/', gsub(':', '/', tolower(gsub('/.*', '', misc.dir))),
                      stri_replace_first_regex(misc.dir, '.*:\\/', '')
  )
  
  minitax.dir  <- paste0('/mnt/', gsub(':', '/', tolower(gsub('/.*', '', minitax.dir))),
                         stri_replace_first_regex(minitax.dir, '.*:\\/', '')
  )
  
}

for(f in list.files(misc.dir,    '*.R', full.names = T)) { try({message(f); source(f)}) }
for(f in list.files(minitax.dir, '*.R', full.names = T)) { try({message(f); source(f)}) }

bam.flags           <- fread(paste0(misc.dir, '/bam.flags.tsv'))
gff_compare_classes <- fread(paste0(misc.dir, '/gff_compare.txt'))
nproc <- 32


### General settings

filt.version <- ''
filtering    <- ''
writetables  <- T
save.Figs    <- F
res.dir      <- 'Zymo_ONT_V1.V9_EMU_EMUdb_compare'; dir.create(res.dir)
#platforms <- c('illumina', 'PacBIO')

## color palette
alpha   <- 0.9
pal.man <- c(brewer.pal(7, 'Set1'), brewer.pal(8, 'Accent')[c(1,2,6)], brewer.pal(8, 'Set2')[c(1,3,5,6,7)], brewer.pal(8, 'Dark2') )
pal.man <- c(colorvec[c(8,32,12:14,17,24,1:2,4,7,5,6,3,10,11,27:28,9,29)], 'grey', 'black')
pal.man <- alpha(c(pal.man, pal.man), alpha)

pal.gigasci <- c(
  #"giga_green"  = 
    "#71BF44",
  #"medium_gray" = 
   "#7E7E7E",
  #"deep_green"  = 
   "#005B2A",
  #"dark_gray"   = 
   "#333333",
  #"light_green" = 
  "#9AD07C"
)

## taxonomic ranks
ranks <- c("superkingdom", "phylum", "class", "order", "family", "genus", "species")

### minitax settings
## Database
db <- 'EMUdb'
db.dir <- '../databases/EMUdb'
## Genome Size normalization?
NormToGenomeSize <- F
NormFunction     <- 'mean'
## MAPQ filtering ?
mapq.filt <- NA #  1:59 # 
## Filtering samples
filter.samples.pattern <- 'LCA|Undetermined|SpeciesEstimate|BestAln'  ### filter out LCA method


## Exclude unclassified reads?
exclude.unclass <- T

## Omit very low abundance samples
abund.threshold <- 10

## Other settings
consider.euk <- T
count_multiplier <- 100000
count_colname    <- c('Zymo_D6323')


### Gold Standard Settings
cami.lineage <- 'db'
add.GS.taxa <- F ## add all GS taxa to the barplots and the correlation plots?
zymo.GS <- read_delim('../standards/Zymo.D6323.DNA.report/zymo_D6323_specdata.tsv')
GS_name <- 'Zymo_6323_WGS_kraken2'
GS_nr   <- 1
GS_version <- 'D6323'
just_GS <- T
vars <- c('platform') # 'GS_name' ('project') # c('DNA_isolation_method', 'db', 'workflow', 'method', 
#vars <- c('Vregion', 'platform', 'db', 'workflow', 'method','DNA_isolation_method') # 'GS_version'
glom_spec <- T

## Gram strain info
phyla.gram <- read.delim('../databases/phyla.gram.tsv')[,1:2]
colnames(phyla.gram)[2] <- 'Gram.Stain'


## Exclude Bifidobacteria?
exclude.Bifido <- F
if(exclude.Bifido) {
  zymo.GS <- zymo.GS[!grepl('Bifodobac', zymo.GS[,1]) & !grepl('Bifodobac', zymo.GS[,2]), ]
}

## Keep and plot GS taxa only?
keep.GS.taxa.only <- F

mark.GS.Taxa <- F
merge.non.GS.Taxa <- F


## calculate correlation and maaslin2 and PERMANOVA from topN (20) or from all taxa?
correl <- 'topN' ## 'topN' ## or: 

## calculate MAASLIN2 and PERMANOVA stats from topN (20) or from all taxa?
stats  <- 'all' ## 'topN' ## or: 

## Fix taxa
crop_taxa <- NULL # 'Veillonella rogosae|Prevotella corporis'

genera_to_glom <- data.frame(glom= c('Veillonella', 'Faecalibacterium', 'Prevotella', 'Roseburia'),
                           to  = c('Veillonella rogosae','Faecalibacterium prausnitzii', 'Prevotella corporis', 'Roseburia hominis')) ## NULL
species_to_glom <- data.frame(glom= c('Bacteroides fragilis', 'Bacteroides fragilis CAG:47', 'Bacteroides fragilis_A'),
                           to  = c('Bacteroides fragilis')) ## NULL



abundance_threshold <- 1
filter_func <- filterfun(function(x) sum(x) > abundance_threshold)

```



# Notes

- Unclassified reads *were* excluded
- Taxa with total reads number of 10 or lower were excluded 
- Eukaryotes *were* excluded from the analyis

```{r, Database import}
# altough *Lactobacillus fermentum* was changed to *Limosilactobacillus fermentum* 
  
#### Database import ####

if (db == 'proGcontigs_2') {
  prog.db <- fread(paste0(db.dir, '/proGenomes2.1_specI_lineageNCBI.tab'), header = F)
  colnames(prog.db) <- c("genome", "superkingdom", "phylum", "class", "order", "family", "genus", "species")
  prog.db <- data.frame(taxid=gsub('\\..*', '', prog.db$genome  ), prog.db)
  ranks <- c("superkingdom", "phylum", "class", "order", "family", "genus", "species")
  prog.db$taxid <- as.integer(prog.db$taxid)
  
  prog.db[,ranks]  <- as.data.frame(apply(prog.db[,ranks], 2, function(x) str_replace(as.character(unlist(x)),  pattern = "^\\d+\\s", replacement = "")))
  
  prog.gt <- gather(prog.db, rank, taxon, -c(1:2))
  prog.db.spec.uni <- unique.data.frame(prog.db[,ranks])
  
  db.data <- prog.db
  db.name <- db
  
  prog.db.uni     <- db.data %>% distinct(across(all_of(c('taxid', ranks)))) #unique.data.frame(db.data[,c('taxid', ranks)])
  db.uni.data     <- prog.db.uni

   if (NormToGenomeSize) {
    genome_sizes <- as.data.frame(fread(paste0(db.dir, '/proGenomes2.1_specI_lineageNCBI.genomesize.tab')))
    genome_sizes[genome_sizes == ''] <- NA
    genome_sizes[,ranks]  <- 
      as.data.frame(apply(genome_sizes[,ranks], 2, 
                          function(x) str_replace(as.character(unlist(x)),  pattern = "^\\d+\\s", replacement = "")))
  }
  
  db.uni.ranks <- unique.data.frame(db.uni.data[,ranks])
  
  #### Fix Lactobacillus in proGenomes_2
  db.uni.ranks$genus  [db.uni.ranks$species ==  'Lactobacillus fermentum' ] <- 'Limosilactobacillus'
  db.uni.ranks$species[db.uni.ranks$species ==  'Lactobacillus fermentum' ] <- 'Limosilactobacillus fermentum'

} else if (db=="all_NCBI_genomes") {
  
  #db.data     <- fread(paste0(db.dir, "/NCBI.db.tsv"), header = T)
  db.uni.data <- fread(paste0(db.dir, "/NCBI.db.uni.tsv"), header = T)
  db.uni.data <- unique(db.uni.data[,..ranks])
  db.uni.data[db.uni.data == ''] <- NA

  if (NormToGenomeSize) {
    genome_sizes <- as.data.frame(fread(paste0(db.dir, '/NCBI.db.genomesize.tsv')))
    genome_sizes[genome_sizes == ''] <- NA
  }
  
  
  #### Fix Akkermansia municiphila
  db.uni.data <- data.table(plyr::rbind.fill(db.uni.data,
    setNames(data.frame(
      rbind(c('Bacteria', 'Verrucomicrobia', 'Verrucomicrobiae', 'Verrucomicrobiales', 'Akkermansiaceae', 'Akkermansia', 'Akkermansia muciniphila'))),
             ranks)
  ))
} else if (db == 'EMUdb') {

  ranks <- c("superkingdom", "phylum", "class", "order", "family", "genus", "species")
  
  try({ db.uni.data <- fread(file.path(db.dir, 'db.uni.data.tsv')) })
  if(!exists('db.uni.data')) {
    emu.db    <- read.delim(paste0(db.dir, '/taxonomy.tsv'))
    colnames(emu.db)[1] <- 'taxid'
    emu.db <- emu.db[,c('taxid', ranks)]
    emu.db[emu.db == ''] <- NA
  
    emu.fasta <- seqinr::read.fasta(paste0(db.dir, '/species_taxid.fasta'), whole.header = T)
    emu.fasta <- data.frame(names=names(emu.fasta), taxid=gsub(':.*', '', names(emu.fasta)))
    emu.fasta$seq_id <- gsub(' \\[.*', '', emu.fasta$names)
  
    emu.idx    <- merge(emu.fasta, emu.db, by='taxid')
  
    db.data <- data.table(emu.db)
    db.name <- db #'EMUdb'
    db.uni.data <- db.data
  }

}


```


# Import Exp data

## MCM D6300
```{r}
load('MCM_D6300_EMU_EMUdb.RData')

ps.all   <- prune_samples(grep('Zymo_', sample_names(ps.all), value = T, invert = T), ps.all)

metadata <- data.frame(sample_data(ps.all))
otutab   <- data.frame(otu_table(ps.all))
taxtab   <- data.frame(tax_table(ps.all))


mark.GS.Taxa      <- F
merge.non.GS.Taxa <- F

ps.d6300 <- ps.all
```

### Species-level

```{r}
ps <- ps.all

n <- 7

t <- rank_names(ps)[n]

ps.glom.norm.top <- ps.glom.norm.top.spec

fig.width.vr   <- 12
fig.height.vr  <- 6
multiV         <- c(T,T)
```

```{r, eval=T}
## ALREADY DONE

## Mark GS Taxa with Asterisks
if (mark.GS.Taxa) {
  ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]] <-
    paste0(ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]], "*")
}

```

```{r, eval=T}
## ALREADY DONE

if (merge.non.GS.Taxa) {
  gs.taxa          <- unique(taxtab.gs[,t])
  ps.glom.norm.top.merged <- merge.other.Taxa(ps.glom.norm.top, gs.taxa )
} else {
  ps.glom.norm.top.merged <- ps.glom.norm.top
}

```

```{r, fig.height = fig.height.vr, fig.width = fig.width.vr, eval=T}
my.pal <- colors_for_PS(ps.glom.norm.top.merged, rank=t, pal.man)

ggtop20    <- plotfun(
  ps.glom.norm.top.merged,
                t=t, fill=t) + 
  theme(legend.position = 'right') +
  ggtitle(paste0('A')) +
  guides(fill = guide_legend(ncol = 2))

ggtop20

##
ps.glom.norm.top.spec.d6300  <- ps.glom.norm.top.merged
gg.glom.norm.top.spec.d6300  <- ggtop20

```

### Detection Metrics 
```{r}

det.stats <- fread('MCM_D6300_EMU_EMUdb/Detection statistics.tsv')
det.sum   <- det.stats[statistic %in% c('Precision', 'Recall'), .(mean = mean(value), sd = sd(value)), by = .(statistic, Detection_threshold_percent, Detection_threshold_ratio, level, primer, GS_version)]
det.sum.d6300 <- det.sum

gs.taxa          <- unique(taxtab.gs[,t])

d6300.dt <- dt.from.ps(ps.all, glom_taxa = T, t = t)
d6300.dt[,.N,sample]

d6300.dt[, norm_count := count / sum(count), by = sample]

n_distinct(d6300.dt[sample == 'Zymo_6300_ONT' & count > 0, species]) == n_distinct(gs.taxa)

d6300.cast <- dcast.data.table(d6300.dt, species ~ sample, value.var = 'norm_count' )

fwrite(d6300.cast, 'MCM_D6300_EMU_EMUdb/MCM_D6300_EMU_EMUdb.tsv', sep = '\t')


```


#### Presence/absence Heatmap
```{r, eval=F}
library(data.table)
library(ggplot2)

# ---- load
dt <- fread("FR_D6323_EMU_EMUdb/FR_D6323_EMU_EMUdb.tsv")  # adjust path if needed

# ---- parameters
top_n <- 250          # number of dominant reference taxa to show
detect_thr <- 1e-6   # detection threshold on relative abundance

# ---- identify library columns (everything except 'species')
lib_cols <- setdiff(names(dt), "species")

# ---- pick top taxa from the "gold/reference" column
stopifnot("Zymo_6323_ONT" %in% names(dt))
top_taxa <- dt[order(-Zymo_6323_ONT)][1:top_n, species]

# ---- long format
long <- melt(
  dt[species %in% top_taxa],
  id.vars = "species",
  measure.vars = lib_cols,
  variable.name = "library",
  value.name = "rel_ab"
)

# ---- detection (binary)
long[, detected := as.integer(rel_ab >= detect_thr)]

# ---- simple platform annotation from column names
long[, platform := fifelse(grepl("_PB_", library), "PacBio",
                    fifelse(grepl("_ONT_", library), "ONT",
                    fifelse(grepl("_MONT_", library), "ONT (mod/deg)", 
                    fifelse(library == "Zymo_6323_ONT", "Reference", "Other"))))]

# ---- order columns: Reference first, then groups
setorder(long, platform, library)
lib_levels <- unique(long$library)
long[, library := factor(library, levels = lib_levels)]

# ---- order rows: by reference abundance descending
tax_levels <- dt[species %in% top_taxa][order(-Zymo_6323_ONT), species]
long[, species := factor(species, levels = rev(tax_levels))]  # reverse so top at top

# ---- plot
p <- ggplot(long, aes(x = library, y = species, fill = detected)) +
  geom_tile(color = "grey90", linewidth = 0.2) +
  facet_grid(. ~ platform, scales = "free_x", space = "free_x") +
  scale_fill_gradient(low = "white", high = "black") +
  theme_classic(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title = element_blank(),
    legend.position = "right",
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  ) +
  labs(fill = paste0("Detected\n(≥ ", detect_thr, ")"),
       title = paste0("Mock standard D6323: detection of top ", top_n, " reference taxa across libraries"))

ggsave("D6323_detection_heatmap.pdf", p, width = 220, height = top_n*4, units = "mm", device = cairo_pdf)
ggsave("D6323_detection_heatmap.png", p, width = 220, height = top_n*4, units = "mm", dpi = 600)

p
```

#### Recall vs Precision
```{r, fig.width=22, fig.height=10}

ggplot(det.stats[statistic %in% c('Precision', 'Recall')]) +
  geom_point(aes(statistic, value, fill = primer), shape = 21, alpha = 0.85, size = 3) +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
  theme_ipsum() +
  facet_wrap(~statistic + Detection_threshold_percent + level, scales = 'free', nrow = 2)


ggplot(det.stats[statistic %in% c('Precision', 'Recall')]) +
  geom_col(aes(statistic, value, fill = primer), position = 'dodge2', color = 'black') +
  scale_fill_npg() +
  coord_cartesian( ylim = c(0,1)) +
  theme_ipsum() +
  facet_wrap(~statistic + Detection_threshold_percent + level, scales = 'free', nrow = 2)
```


```{r, fig.width=12, fig.height=10}
prec.rec <- dcast.data.table(det.stats[statistic %in% c('Precision', 'Recall')], ... ~ statistic, value.var = 'value')

ggplot(prec.rec) +
  geom_point(aes(Precision, Recall, fill = primer), shape = 21, alpha = 0.85, size = 3) +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
  theme_ipsum() +
  facet_wrap(~level + Detection_threshold_percent, scales = 'free')


prec.rec <- dcast.data.table(det.stats[statistic %in% c('False_negatives', 'False_positives')], ... ~ statistic, value.var = 'value')

ggplot(prec.rec) +
  geom_point(aes(False_negatives, False_positives, fill = primer), shape = 21, alpha = 0.85, size = 3) +
  #coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
  theme_ipsum() +
  facet_wrap(~level + Detection_threshold_percent, scales = 'free')


```

#### Fig 3
```{r, fig.width=14, fig.height=8}

ggp <- ggplot(det.sum,
       aes(Detection_threshold_ratio, mean, color = primer, shape = statistic, fill = primer)) +
  geom_point(size = 4, color = 'black') +
  scale_shape_manual(values = c(
    Precision = 21,   # filled circle
    Recall    = 24    # filled triangle
  )) +
  geom_line(aes(group = interaction(primer, statistic))) +
  scale_fill_npg() +
  scale_color_npg() +
  scale_x_continuous(labels = scales::label_percent()) +
  coord_cartesian( ylim = c(0,1)) +
  labs(y = 'Mean value', x = 'Detection threshold (% of reads') +
  theme_ipsum() +
  theme(panel.border = element_rect(linewidth = 0.5)) +
  facet_wrap(~ level, scales = 'free', nrow = 1)


ggp

#ggsave('Article_Figures/FR_D6323_EMU_EMUdb_prec.vs.recall.png', width = 14, height = 8, bg = 'white') 
```




## MCM D6331
```{r}
load('MCM_D6331_EMU_EMUdb.RData')

ps.all   <- prune_samples(grep('Zymo_', sample_names(ps.all), value = T, invert = T), ps.all)

metadata <- data.frame(sample_data(ps.all))
otutab   <- data.frame(otu_table(ps.all))
taxtab   <- data.frame(tax_table(ps.all))

mark.GS.Taxa      <- F
merge.non.GS.Taxa <- F
   
ps.d6331 <- ps.all

```

### Species-level

```{r}
ps <- ps.all

n <- 7

t <- rank_names(ps)[n]

ps.glom.norm.top <- ps.glom.norm.top.spec

fig.width.vr   <- 12
fig.height.vr  <- 6
multiV         <- c(T,T)
```

```{r, eval=T}
## Already done!

## Mark GS Taxa with Asterisks
if (mark.GS.Taxa) {
  ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]] <-
    paste0(ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]], "*")
}

```

```{r, eval=T}
## Already done!

if (merge.non.GS.Taxa) {
  gs.taxa          <- unique(taxtab.gs[,t])
  ps.glom.norm.top.merged <- merge.other.Taxa(ps.glom.norm.top, gs.taxa )
} else {
  ps.glom.norm.top.merged <- ps.glom.norm.top
}

```

```{r, fig.height = fig.height.vr, fig.width = fig.width.vr, eval=T}
my.pal <- colors_for_PS(ps.glom.norm.top.merged, rank=t, pal.man)

ggtop20    <- plotfun(
  ps.glom.norm.top.merged,
                t=t, fill=t) + 
  theme(legend.position = 'right') +
  ggtitle(paste0('B')) +
  guides(fill = guide_legend(ncol = 2))

ggtop20

##
ps.glom.norm.top.spec.d6331  <- ps.glom.norm.top.merged
gg.glom.norm.top.spec.d6331  <- ggtop20

```

### Detection Metrics 
```{r}

det.stats <- fread('MCM_D6331_EMU_EMUdb/Detection statistics.tsv')
det.sum   <- det.stats[statistic %in% c('Precision', 'Recall'), .(mean = mean(value), sd = sd(value)), by = .(statistic, Detection_threshold_percent, Detection_threshold_ratio, level, primer, GS_version)]
det.sum.d6331 <- det.sum

gs.taxa          <- unique(taxtab.gs[,t])

d6331.dt <- dt.from.ps(ps.all, glom_taxa = T, t = t)
d6331.dt[,.N,sample]

d6331.dt[, norm_count := count / sum(count), by = sample]

n_distinct(d6331.dt[sample == 'Zymo_6331_ONT' & count > 0, species]) == n_distinct(gs.taxa)

d6331.cast <- dcast.data.table(d6331.dt, species ~ sample, value.var = 'norm_count' )

fwrite(d6331.cast, 'MCM_D6331_EMU_EMUdb/MCM_D6331_EMU_EMUdb.tsv', sep = '\t')


```


#### Presence/absence Heatmap
```{r, eval=F}
library(data.table)
library(ggplot2)

# ---- load
dt <- fread("FR_D6323_EMU_EMUdb/FR_D6323_EMU_EMUdb.tsv")  # adjust path if needed

# ---- parameters
top_n <- 250          # number of dominant reference taxa to show
detect_thr <- 1e-6   # detection threshold on relative abundance

# ---- identify library columns (everything except 'species')
lib_cols <- setdiff(names(dt), "species")

# ---- pick top taxa from the "gold/reference" column
stopifnot("Zymo_6323_ONT" %in% names(dt))
top_taxa <- dt[order(-Zymo_6323_ONT)][1:top_n, species]

# ---- long format
long <- melt(
  dt[species %in% top_taxa],
  id.vars = "species",
  measure.vars = lib_cols,
  variable.name = "library",
  value.name = "rel_ab"
)

# ---- detection (binary)
long[, detected := as.integer(rel_ab >= detect_thr)]

# ---- simple platform annotation from column names
long[, platform := fifelse(grepl("_PB_", library), "PacBio",
                    fifelse(grepl("_ONT_", library), "ONT",
                    fifelse(grepl("_MONT_", library), "ONT (mod/deg)", 
                    fifelse(library == "Zymo_6323_ONT", "Reference", "Other"))))]

# ---- order columns: Reference first, then groups
setorder(long, platform, library)
lib_levels <- unique(long$library)
long[, library := factor(library, levels = lib_levels)]

# ---- order rows: by reference abundance descending
tax_levels <- dt[species %in% top_taxa][order(-Zymo_6323_ONT), species]
long[, species := factor(species, levels = rev(tax_levels))]  # reverse so top at top

# ---- plot
p <- ggplot(long, aes(x = library, y = species, fill = detected)) +
  geom_tile(color = "grey90", linewidth = 0.2) +
  facet_grid(. ~ platform, scales = "free_x", space = "free_x") +
  scale_fill_gradient(low = "white", high = "black") +
  theme_classic(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title = element_blank(),
    legend.position = "right",
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  ) +
  labs(fill = paste0("Detected\n(≥ ", detect_thr, ")"),
       title = paste0("Mock standard D6323: detection of top ", top_n, " reference taxa across libraries"))

ggsave("D6323_detection_heatmap.pdf", p, width = 220, height = top_n*4, units = "mm", device = cairo_pdf)
ggsave("D6323_detection_heatmap.png", p, width = 220, height = top_n*4, units = "mm", dpi = 600)

p
```

#### Recall vs Precision
```{r, fig.width=22, fig.height=10}

ggplot(det.stats[statistic %in% c('Precision', 'Recall')]) +
  geom_point(aes(statistic, value, fill = primer), shape = 21, alpha = 0.85, size = 3) +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
  theme_ipsum() +
  facet_wrap(~statistic + Detection_threshold_percent + level, scales = 'free', nrow = 2)


ggplot(det.stats[statistic %in% c('Precision', 'Recall')]) +
  geom_col(aes(statistic, value, fill = primer), position = 'dodge2', color = 'black') +
  scale_fill_npg() +
  coord_cartesian( ylim = c(0,1)) +
  theme_ipsum() +
  facet_wrap(~statistic + Detection_threshold_percent + level, scales = 'free', nrow = 2)
```


```{r, fig.width=12, fig.height=10}
prec.rec <- dcast.data.table(det.stats[statistic %in% c('Precision', 'Recall')], ... ~ statistic, value.var = 'value')

ggplot(prec.rec) +
  geom_point(aes(Precision, Recall, fill = primer), shape = 21, alpha = 0.85, size = 3) +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
  theme_ipsum() +
  facet_wrap(~level + Detection_threshold_percent, scales = 'free')


prec.rec <- dcast.data.table(det.stats[statistic %in% c('False_negatives', 'False_positives')], ... ~ statistic, value.var = 'value')

ggplot(prec.rec) +
  geom_point(aes(False_negatives, False_positives, fill = primer), shape = 21, alpha = 0.85, size = 3) +
  #coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
  theme_ipsum() +
  facet_wrap(~level + Detection_threshold_percent, scales = 'free')


```

#### Fig 3
```{r, fig.width=14, fig.height=8}

ggp <- ggplot(det.sum,
       aes(Detection_threshold_ratio, mean, color = primer, shape = statistic, fill = primer)) +
  geom_point(size = 4, color = 'black') +
  scale_shape_manual(values = c(
    Precision = 21,   # filled circle
    Recall    = 24    # filled triangle
  )) +
  geom_line(aes(group = interaction(primer, statistic))) +
  scale_fill_npg() +
  scale_color_npg() +
  scale_x_continuous(labels = scales::label_percent()) +
  coord_cartesian( ylim = c(0,1)) +
  labs(y = 'Mean value', x = 'Detection threshold (% of reads') +
  theme_ipsum() +
  theme(panel.border = element_rect(linewidth = 0.5)) +
  facet_wrap(~ level, scales = 'free', nrow = 1)


ggp

#ggsave('Article_Figures/FR_D6323_EMU_EMUdb_prec.vs.recall.png', width = 14, height = 8, bg = 'white') 
```


## FR D6323
```{r}
load('FR_D6323_EMU_EMUdb.RData')

ps.all   <- prune_samples(grep('Zymo_', sample_names(ps.all), value = T, invert = T), ps.all)

metadata <- data.frame(sample_data(ps.all))
otutab   <- data.frame(otu_table(ps.all))
taxtab   <- data.frame(tax_table(ps.all))


mark.GS.Taxa      <- F
merge.non.GS.Taxa <- F


ps.d6323 <- ps.all

```

### Species-level

```{r}
ps <- ps.all

n <- 7

t <- rank_names(ps)[n]

ps.glom.norm.top <- ps.glom.norm.top.spec

fig.width.vr   <- 12
fig.height.vr  <- 6
multiV         <- c(T,T)
```

```{r, eval=T}
## Already done!

## Mark GS Taxa with Asterisks
if (mark.GS.Taxa) {
  ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]] <-
    paste0(ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]], "*")
}

```

```{r, eval=T}
## Already done!

if (merge.non.GS.Taxa) {
  gs.taxa          <- unique(taxtab.gs[,t])
  ps.glom.norm.top.merged <- merge.other.Taxa(ps.glom.norm.top, gs.taxa )
} else {
  ps.glom.norm.top.merged <- ps.glom.norm.top
}

```

```{r, fig.height = fig.height.vr, fig.width = fig.width.vr, eval=T}
my.pal <- colors_for_PS(ps.glom.norm.top.merged, rank=t, pal.man)


ggtop20    <- plotfun(
  ps.glom.norm.top.merged,
                t=t, fill=t) + 
  theme(legend.position = 'right') +
  ggtitle(paste0('A')) +
  guides(fill = guide_legend(ncol = 2))
  

ggtop20

##
ps.glom.norm.top.spec.d6323  <- ps.glom.norm.top.merged
gg.glom.norm.top.spec.d6323  <- ggtop20

```

### Detection Metrics 
```{r}

det.stats <- fread('FR_D6323_EMU_EMUdb/Detection statistics.tsv')
det.sum   <- det.stats[statistic %in% c('Precision', 'Recall'), .(mean = mean(value), sd = sd(value)), by = .(statistic, Detection_threshold_percent, Detection_threshold_ratio, level, primer, GS_version)]
det.sum.d6323 <- det.sum

gs.taxa          <- unique(taxtab.gs[,t])

d6323.dt <- dt.from.ps(ps.all, glom_taxa = T, t = t)
d6323.dt[,.N,sample]

d6323.dt[, norm_count := count / sum(count), by = sample]

n_distinct(d6323.dt[sample == 'Zymo_6323_ONT' & count > 0, species]) == n_distinct(gs.taxa)

d6323.cast <- dcast.data.table(d6323.dt, species ~ sample, value.var = 'norm_count' )

fwrite(d6323.cast, 'FR_D6323_EMU_EMUdb/FR_D6323_EMU_EMUdb.tsv', sep = '\t')


```


#### Presence/absence Heatmap
```{r, eval=F}
library(data.table)
library(ggplot2)

# ---- load
dt <- fread("FR_D6323_EMU_EMUdb/FR_D6323_EMU_EMUdb.tsv")  # adjust path if needed

# ---- parameters
top_n <- 250          # number of dominant reference taxa to show
detect_thr <- 1e-6   # detection threshold on relative abundance

# ---- identify library columns (everything except 'species')
lib_cols <- setdiff(names(dt), "species")

# ---- pick top taxa from the "gold/reference" column
stopifnot("Zymo_6323_ONT" %in% names(dt))
top_taxa <- dt[order(-Zymo_6323_ONT)][1:top_n, species]

# ---- long format
long <- melt(
  dt[species %in% top_taxa],
  id.vars = "species",
  measure.vars = lib_cols,
  variable.name = "library",
  value.name = "rel_ab"
)

# ---- detection (binary)
long[, detected := as.integer(rel_ab >= detect_thr)]

# ---- simple platform annotation from column names
long[, platform := fifelse(grepl("_PB_", library), "PacBio",
                    fifelse(grepl("_ONT_", library), "ONT",
                    fifelse(grepl("_MONT_", library), "ONT (mod/deg)", 
                    fifelse(library == "Zymo_6323_ONT", "Reference", "Other"))))]

# ---- order columns: Reference first, then groups
setorder(long, platform, library)
lib_levels <- unique(long$library)
long[, library := factor(library, levels = lib_levels)]

# ---- order rows: by reference abundance descending
tax_levels <- dt[species %in% top_taxa][order(-Zymo_6323_ONT), species]
long[, species := factor(species, levels = rev(tax_levels))]  # reverse so top at top

# ---- plot
p <- ggplot(long, aes(x = library, y = species, fill = detected)) +
  geom_tile(color = "grey90", linewidth = 0.2) +
  facet_grid(. ~ platform, scales = "free_x", space = "free_x") +
  scale_fill_gradient(low = "white", high = "black") +
  theme_classic(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title = element_blank(),
    legend.position = "right",
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  ) +
  labs(fill = paste0("Detected\n(≥ ", detect_thr, ")"),
       title = paste0("Mock standard D6323: detection of top ", top_n, " reference taxa across libraries"))

ggsave("D6323_detection_heatmap.pdf", p, width = 220, height = top_n*4, units = "mm", device = cairo_pdf)
ggsave("D6323_detection_heatmap.png", p, width = 220, height = top_n*4, units = "mm", dpi = 600)

p
```

#### Recall vs Precision
```{r, fig.width=22, fig.height=10}

ggplot(det.stats[statistic %in% c('Precision', 'Recall')]) +
  geom_point(aes(statistic, value, fill = primer), shape = 21, alpha = 0.85, size = 3) +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
  theme_ipsum() +
  facet_wrap(~statistic + Detection_threshold_percent + level, scales = 'free', nrow = 2)


ggplot(det.stats[statistic %in% c('Precision', 'Recall')]) +
  geom_col(aes(statistic, value, fill = primer), position = 'dodge2', color = 'black') +
  scale_fill_npg() +
  coord_cartesian( ylim = c(0,1)) +
  theme_ipsum() +
  facet_wrap(~statistic + Detection_threshold_percent + level, scales = 'free', nrow = 2)
```


```{r, fig.width=12, fig.height=10}
prec.rec <- dcast.data.table(det.stats[statistic %in% c('Precision', 'Recall')], ... ~ statistic, value.var = 'value')

ggplot(prec.rec) +
  geom_point(aes(Precision, Recall, fill = primer), shape = 21, alpha = 0.85, size = 3) +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
  theme_ipsum() +
  facet_wrap(~level + Detection_threshold_percent, scales = 'free')


prec.rec <- dcast.data.table(det.stats[statistic %in% c('False_negatives', 'False_positives')], ... ~ statistic, value.var = 'value')

ggplot(prec.rec) +
  geom_point(aes(False_negatives, False_positives, fill = primer), shape = 21, alpha = 0.85, size = 3) +
  #coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
  theme_ipsum() +
  facet_wrap(~level + Detection_threshold_percent, scales = 'free')


```

#### Fig 3
```{r, fig.width=14, fig.height=8}

ggp <- ggplot(det.sum,
       aes(Detection_threshold_ratio, mean, color = primer, shape = statistic, fill = primer)) +
  geom_point(size = 4, color = 'black') +
  scale_shape_manual(values = c(
    Precision = 21,   # filled circle
    Recall    = 24    # filled triangle
  )) +
  geom_line(aes(group = interaction(primer, statistic))) +
  scale_fill_npg() +
  scale_color_npg() +
  scale_x_continuous(labels = scales::label_percent()) +
  coord_cartesian( ylim = c(0,1)) +
  labs(y = 'Mean value', x = 'Detection threshold (% of reads') +
  theme_ipsum() +
  theme(panel.border = element_rect(linewidth = 0.5)) +
  facet_wrap(~ level, scales = 'free', nrow = 1)


ggp

#ggsave('Article_Figures/FR_D6323_EMU_EMUdb_prec.vs.recall.png', width = 14, height = 8, bg = 'white') 
```


## *Canis lupus*
```{r, eval=T}
load('Clupus_EMU_EMUdb.RData')

ps.all   <- prune_samples(grep('Zymo_', sample_names(ps.all), value = T, invert = T), ps.all)

metadata <- data.frame(sample_data(ps.all))
otutab   <- data.frame(otu_table(ps.all))
taxtab   <- data.frame(tax_table(ps.all))

ps.all@sam_data$source <- 'Canis_lupus'

mark.GS.Taxa      <- F
merge.non.GS.Taxa <- F

ps.clupus <- ps.all

```

### Species-level

```{r}
ps <- ps.all

n <- 7

t <- rank_names(ps)[n]

ps.glom.norm.top <- ps.glom.norm.top.spec

fig.width.vr   <- 12
fig.height.vr  <- 6
multiV         <- c(T,T)
```

```{r, eval=T}
## Already done!

## Mark GS Taxa with Asterisks
if (mark.GS.Taxa) {
  ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]] <-
    paste0(ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]], "*")
}

```

```{r, eval=T}
## Already done!

if (merge.non.GS.Taxa) {
  gs.taxa          <- unique(taxtab.gs[,t])
  ps.glom.norm.top.merged <- merge.other.Taxa(ps.glom.norm.top, gs.taxa )
} else {
  ps.glom.norm.top.merged <- ps.glom.norm.top
}

```

```{r, fig.height = fig.height.vr, fig.width = fig.width.vr, eval=T}
my.pal <- colors_for_PS(ps.glom.norm.top.merged, rank=t, pal.man)


ggtop20    <- plotfun(
  ps.glom.norm.top.merged,
                t=t, fill=t) + 
  theme(legend.position = 'right') +
  ggtitle(paste0('B')) +
  guides(fill = guide_legend(ncol = 2))
  

ggtop20

##
ps.glom.norm.top.spec.clupus  <- ps.glom.norm.top.merged
gg.glom.norm.top.spec.clupus  <- ggtop20

```


## *Homo sapiens*
```{r, eval=T}
load('Hsapiens_EMU_EMUdb.RData')

ps.all   <- prune_samples(grep('Zymo_', sample_names(ps.all), value = T, invert = T), ps.all)

metadata <- data.frame(sample_data(ps.all))
otutab   <- data.frame(otu_table(ps.all))
taxtab   <- data.frame(tax_table(ps.all))

ps.all@sam_data$source <- 'Homo_sapiens'

mark.GS.Taxa      <- F
merge.non.GS.Taxa <- F

ps.hsapiens <- ps.all

```

### Species-level

```{r}
ps <- ps.all

n <- 7

t <- rank_names(ps)[n]

ps.glom.norm.top <- ps.glom.norm.top.spec

fig.width.vr   <- 12
fig.height.vr  <- 6
multiV         <- c(T,T)
```

```{r, eval=T}
## Already done!

## Mark GS Taxa with Asterisks
if (mark.GS.Taxa) {
  ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]] <-
    paste0(ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]], "*")
}

```

```{r, eval=T}
## Already done!

if (merge.non.GS.Taxa) {
  gs.taxa          <- unique(taxtab.gs[,t])
  ps.glom.norm.top.merged <- merge.other.Taxa(ps.glom.norm.top, gs.taxa )
} else {
  ps.glom.norm.top.merged <- ps.glom.norm.top
}

```

```{r, fig.height = fig.height.vr, fig.width = fig.width.vr, eval=T}
my.pal <- colors_for_PS(ps.glom.norm.top.merged, rank=t, pal.man)


ggtop20    <- plotfun(
  ps.glom.norm.top.merged,
                t=t, fill=t) + 
  ggtitle(paste0('C')) +
  guides(fill = guide_legend(ncol = 2)) +
  theme(legend.position = 'right', title = element_text(face = 'italic'))
  

ggtop20

##
ps.glom.norm.top.spec.hsapiens <- ps.glom.norm.top.merged
gg.glom.norm.top.spec.hsapiens <- ggtop20

```


## Figure 3

### Figure 3A
```{r, fig.width = 18, fig.height = 14}
det.sum <- rbind(det.sum.d6300,
                 det.sum.d6331,
                 det.sum.d6323)

ggp <- ggplot(det.sum,
       aes(Detection_threshold_ratio, mean, color = primer, shape = statistic, fill = primer)) +
  geom_point(size = 4, color = 'black') +
  scale_shape_manual(values = c(
    Precision = 21,   # filled circle
    Recall    = 24    # filled triangle
  )) +
  geom_line(aes(group = interaction(primer, statistic))) +
  scale_fill_npg() +
  scale_color_npg() +
  scale_x_continuous(labels = scales::label_percent()) +
  coord_cartesian( ylim = c(0,1)) +
  labs(y = 'Mean value', x = 'Detection threshold (% of reads') +
  theme_ipsum() +
  theme(panel.border = element_rect(linewidth = 0.5)) +
  facet_wrap(~ level + GS_version, scales = 'free', nrow = 2)


ggp

ggsave('Article_Figures/Fig3A_Prec.vs.recall.png', width = 18, height = 14, bg = 'white') 

Fig3A <- ggp
```


### Figure 3B
```{r}
library(eulerr)
library(ComplexUpset)
```

```{r, fig.height=8, fig.width=12}
i <- 7
t <- rank_names(ps.all)[i]

comb.dt <- rbind(dt.from.ps(ps.d6323),
                 dt.from.ps(ps.hsapiens),
                 dt.from.ps(ps.clupus),
                 fill = T)

comb.dt <- comb.dt[count > 0, ]

cols <- c(t, 'source')
taxa.pres <- comb.dt[, .(count = sum(count)), by = cols]

taxa.pres <- dcast.data.table(taxa.pres, ... ~ source, value.var = 'count')

## 6  Overlap UpSet plot ------------------------------------------------------

up_dt <- copy(taxa.pres)
up_dt[, (names(up_dt)[-1]) := lapply(.SD, Negate(is.na)),
          .SDcols = -1]
intersect <- colnames(up_dt)[-1]

### UpSet
p1 <- ComplexUpset::upset(up_dt,
    intersect = intersect,
    min_size  = 1,
    sort_sets = FALSE,
    themes = theme_ipsum(),
    sort_intersections_by = "cardinality",
    base_annotations = list("Intersection size" = intersection_size(counts = TRUE)))

p1

ggsave(paste0('Article_Figures/', t, '_overlap_UpSet.png'), p1,
       width = 12, height = 9)
#saveRDS(p1, paste0(res.dir, '/', t, '_overlap_UpSet.rds'))
```


```{r, fig.height=6, fig.width=8}
### Euler
fit1 <- euler(list(
    `Fecal Reference`= unique( up_dt[`FR`           == T,  get(t)] ),
    `Homo sapiens`   = unique( up_dt[`Homo_sapiens` == T,  get(t)] ),
    `Canis lupus`    = unique( up_dt[`Canis_lupus`  == T,  get(t)] )
))

p1 <- plot(
  fit1,
  quantities = TRUE,
  fills = list(fill = c("Fecal Reference" = "#1f77b4", "Homo sapiens" = "#ff7f0e", "Canis lupus" = "#005B2A"), alpha=0.6),
  edges = list(col  = c("Fecal Reference" = "black",   "Homo sapiens" = "black",   "Canis lupus" = "black"  ), lwd = 1)
)

p1

ggsave(paste0('Article_Figures/', t, '_overlap_Euler.png'), p1,
       width = 12, height = 9)
saveRDS(p1, paste0('Article_Figures/', t, '_overlap_Euler.rds'))

Fig3B <- p1
```



```{r}

library(cowplot)

Fig3 <- plot_grid(Fig3A, Fig3B, nrow = 1, rel_widths = c(2.5,1), labels = c('a', 'b'))

dir.create('Article_Figures')
ggsave(paste0('Article_Figures', '/', 'Figure3_Final.png'), Fig3, width = 21, height = 12, bg = 'white')


Fig3

```


## Figure 6
```{r, fig.height = fig.height.vr*3, fig.width = fig.width.vr+8, eval=T}

res.dir <- 'Article_Figures'; dir.create(res.dir)


gg.glom.norm.top.spec.all <- cowplot::plot_grid(
  gg.glom.norm.top.spec.d6300,
  gg.glom.norm.top.spec.d6331,
  ncol = 1, axis = 'b', align = 'vh'
)

gg.glom.norm.top.spec.all

ggsave(file.path(res.dir, 'Figure_6.png'), gg.glom.norm.top.spec.all, height = 18, width = 20)
```


## Figure 7
```{r, fig.height = fig.height.vr*3, fig.width = fig.width.vr+8, eval=T}

res.dir <- 'Article_Figures'; dir.create(res.dir)


gg.glom.norm.top.spec.all <- cowplot::plot_grid(
  gg.glom.norm.top.spec.d6323,
  gg.glom.norm.top.spec.clupus,
  gg.glom.norm.top.spec.hsapiens,
  ncol = 1, axis = 'b', align = 'vh'
)

gg.glom.norm.top.spec.all

ggsave(file.path(res.dir, 'Figure_7.png'), gg.glom.norm.top.spec.all, height = 22, width = 14)
```



```{r}

fwrite(df.from.ps(ps.glom.norm.top.spec.d6300, comb=F),    'd6300.norm.top.spec.tsv',    sep ='\t')
fwrite(df.from.ps(ps.glom.norm.top.spec.d6331, comb=F),    'd6331.norm.top.spec.tsv',    sep ='\t')
fwrite(df.from.ps(ps.glom.norm.top.spec.d6323, comb=F),    'd6323.norm.top.spec.tsv',    sep ='\t')
fwrite(df.from.ps(ps.glom.norm.top.spec.hsapiens, comb=F), 'hsapiens.norm.top.spec.tsv', sep ='\t')
fwrite(df.from.ps(ps.glom.norm.top.spec.clupus, comb=F),   'clupus.norm.top.spec.tsv',   sep ='\t')


save.image(paste0(res.dir, '.RData'))
```


