---
title: "Zymo FR D6323 V1-V9 ONT EMU on EMU database"
author: "Balázs Kakuk"
date: "2024-10-20"
output:
  html_document:
    df_print: paged
    code_folding: hide
    theme: cosmo
    toc: true
    toc_float:
      collapsed: false
  pdf_document:
    toc: true

---

```{css, echo=FALSE}
    body .main-container {
      max-width: 90% !important;
      width: 90% !important;
    }
    body {
      max-width: 90% !important;
      margin-left: auto;
      margin-right: auto;
    }
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, fig.align = 'center', fig.height = 10, fig.width = 30, warning = F, message = F, error = F)
library(tidyverse)
library(ggpubr)
library(factoextra)
library(FactoMineR)
library(ggrepel)
library(phyloseq, quietly = T)
library(DESeq2)
library(data.table)
library(ggsci)
library(hrbrthemes)
library(ggh4x)
library(RColorBrewer)
library(stringr); library(stringi)
library("ape")
library("ggtree")
library(Biostrings)
library(dplyr)
library(UpSetR)
library(ComplexHeatmap)
library(genefilter)
library(viridis)

## Own functions <<- SET PATH !!!
misc.dir    <- 'C:/GitHub/Rlyeh/R'
minitax.dir <- 'C:/GitHub/minitax/R'

if (.Platform$OS.type!="windows") {
  
  misc.dir  <- paste0('/mnt/', gsub(':', '/', tolower(gsub('/.*', '', misc.dir))),
                      stri_replace_first_regex(misc.dir, '.*:\\/', '')
  )
  
  minitax.dir  <- paste0('/mnt/', gsub(':', '/', tolower(gsub('/.*', '', minitax.dir))),
                         stri_replace_first_regex(minitax.dir, '.*:\\/', '')
  )
  
}

for(f in list.files(misc.dir,    '*.R', full.names = T)) { try({message(f); source(f)}) }
for(f in list.files(minitax.dir, '*.R', full.names = T)) { try({message(f); source(f)}) }

bam.flags           <- fread(paste0(misc.dir, '/bam.flags.tsv'))
gff_compare_classes <- fread(paste0(misc.dir, '/gff_compare.txt'))
nproc <- 32


```

# Settings
```{r}
# -----------------------------
# 2) Publication theme
# -----------------------------
theme_pub <- function(base_size = 10) {
  theme_classic(base_size = base_size) +
    theme(
      plot.title.position = "plot",
      plot.title = element_text(face = "bold"),
      axis.title = element_text(face = "plain"),
      legend.position = "none",
      axis.text.x = element_text(face = "bold")
    )
}

source('SeqSum_QC.R')

```


# Toti 

```{r}
out_dir <- '1B-C_Toti_WGS'
```

## Illumina WGS

```{r}

```



## ONT WGS : Toti MinION LSK-114 WGS (Zymo)
```{r}

```


## Comb
```{r, message=T, eval=F}
files   <- list.files('D:/ENA/PRJEB75753/.fastq_read_stats', recursive = T, full.names = T)

i <- 1

file <- files[i]

mapfun <- function(i, files) { 
  file <- files[i]
  sample_name <- gsub('.meanQ.tsv', '', gsub('D:/ENA/PRJEB75753/.fastq_read_stats/', '', file) )
  dt <- fread(file)
  dt[, sample := sample_name]
  return(dt)
}

proj_stats_dt <- rbindlist( purrr::map(1:length(files), mapfun, files) )

fwrite(proj_stats_dt[,.(sample, read_id, length, mean_Q)], 'ENA/PRJEB75753/PRJEB75753_stats.tsv.gz', sep = '\t')
```


```{r, message=T, eval=F}
proj_stats_dt <- fread('ENA/PRJEB75753/PRJEB75753_stats.tsv.gz')

proj_stats_dt[, project := 'PRJEB75753']

ena_mt <- fread('ENA/PRJEB75753/filereport_read_run_PRJEB75753.tsv')

PRJEB75753_st <- merge(proj_stats_dt, 
                       ena_mt[,.(run_accession, study_accession, sample_accession, secondary_sample_accession, sample_title, instrument_platform)], 
                       by.x = c('sample', 'project'), by.y = c('run_accession', 'study_accession')
                       #, all = T
                       )

if(nrow(proj_stats_dt) == nrow(PRJEB75753_st)) {
  message('ok! every sample matches with the metadata!')
  PRJEB75753_stats <- PRJEB75753_st; rm(PRJEB75753_st)
} else {
  stop('ERROR! Some samples have not been downloaded and/or imported!')
}

PRJEB75753_stats[,.N, sample_title]


all_stats_uni  <- PRJEB75753_stats[,.(sample, read_id, length, mean_Q)]
proj_meta      <- unique(PRJEB75753_stats[,.(sample, sample_title, project, secondary_sample_accession, instrument_platform)])




fwrite(proj_meta, '1B-C_Toti_WGS/proj_meta.tsv.gz', sep = '\t')

fwrite(all_stats_uni, '1B-C_Toti_WGS/all_stats_uni.tsv.gz', sep = '\t')

```


```{r, eval=F}
cmd <- 'split -b 95M 1B-C_Toti_WGS/all_stats_uni.tsv.gz 1B-C_Toti_WGS/all_stats_uni.tsv.gz.part_'
system(cmd)

cmd <- 'rm 1B-C_Toti_WGS/all_stats_uni.tsv.gz'
system(cmd)

```

```{bash}
#cmd <- '
cat 1B-C_Toti_WGS/all_stats_uni.tsv.gz.part_* > 1B-C_Toti_WGS/all_stats_uni.tsv.gz
#'
#system(cmd)


```

```{r}
try({ all_stats_uni <- fread('1B-C_Toti_WGS/all_stats_uni.tsv.gz') })


proj_meta <- fread('1B-C_Toti_WGS/proj_meta.tsv.gz')


PRJEB75753_stats <- merge(all_stats_uni, proj_meta, by = 'sample')
```




## Summary
```{r}
library(data.table)

# all_stats_uni <- fread('all_stats_uni.tsv.gz')

# Make sure it's data.table
setDT(PRJEB75753_stats)

# Keep only rows with defined isolation method
dt <- PRJEB75753_stats[!is.na(instrument_platform)]

# Ensure numeric
dt[, length := as.numeric(length)]
dt[, mean_Q := as.numeric(mean_Q)]

# ---- N50 function
n50 <- function(x) {
  x <- x[is.finite(x) & x > 0]
  x <- sort(x, decreasing = TRUE)
  x[which(cumsum(x) >= sum(x)/2)[1]]
}

# =========================
# 1️⃣ Per-sample summary
# =========================

sample_summary <- dt[, .(
  n_reads = .N,
  total_bases = sum(length, na.rm = TRUE),
  mean_len = mean(length, na.rm = TRUE),
  median_len = median(length, na.rm = TRUE),
  n50_len = n50(length),
  mean_Q   = mean(mean_Q, na.rm = TRUE),
  median_Q = median(mean_Q, na.rm = TRUE),
  max_Q    = max(mean_Q, na.rm = TRUE),
  min_Q    = min(mean_Q, na.rm = TRUE)
), by = .(sample, instrument_platform)]


# =========================
# 2️⃣ Per-source summary (across samples)
# =========================

platform_summary <- sample_summary[, .(
  n_samples = .N,
  mean_reads = mean(n_reads),
  sd_reads = sd(n_reads),
  mean_total_bases = mean(total_bases),
  sd_total_bases = sd(total_bases),
  mean_read_length = mean(mean_len),
  sd_read_length = sd(mean_len),
  mean_Q = mean(mean_Q),
  sd_Q = sd(mean_Q)
), by = .(instrument_platform)]



# Save
fwrite(sample_summary,   "1B-C_Toti_WGS/Toti_sample_QC_summary.tsv", sep = '\t')
fwrite(platform_summary, "1B-C_Toti_WGS/Toti_platform_QC_summary.tsv", sep = '\t')

sample_summary
platform_summary

```


## Plots

```{r}
out_dir <- '1B-C_Toti_WGS'

dt_plot <- copy(PRJEB75753_stats)

dt_plot[,log_len := log10(length)]

dt_plot <- dt_plot[sample(.N)]  # randomize draw order if you keep any point layers

library(ggpubr)
library(cowplot)
library(ggrastr)
library(scales)

my_pal <- pal_npg()(10)[]
gc()
```

### Fig 1A

```{r}
# Marginal density plot of x (top panel) and y (right panel)
xplot <- ggdensity(dt_plot, "log_len", fill = "instrument_platform",
                   palette = "npg") 
 #+ ggpubr::rotate()

yplot <- ggdensity(dt_plot, "mean_Q", fill = "instrument_platform", 
                   palette = "npg") +
  ggpubr::rotate()



sp <- ggplot(dt_plot, aes(log_len, mean_Q)) +
  geom_point_rast(aes(fill = instrument_platform), alpha = 0.8, shape = 21, color = 'black', size = 3) +
  scale_fill_manual(values = my_pal) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    #title = "Read quality comparison between extraction methods",
    x = "Read count per sample",
    y = "Mean Q-score per sample"
  ) +
  theme_pub()



# Cleaning the plots
sp <- sp + theme(legend.position = 'bottom') #+ rremove("legend")
yplot <- yplot + clean_theme() + rremove("legend")
xplot <- xplot + clean_theme() + rremove("legend")

# Arranging the plot using cowplot
p_count_q <- plot_grid(xplot, NULL, 
                       sp, yplot, 
                       ncol = 2, align = "hv",
                       rel_widths = c(2, 1), rel_heights = c(1, 2))



out_prefix <- "Platform_comparison_QC_Fig1A"

ggsave(paste0(out_dir, '/', out_prefix, '.png'), p_count_q, width = 30, height = 30, bg = 'white')

p_count_q

#save_plot(p_count_q, out_dir, out_prefix, w = 300, h = 300, bg = 'white')

```


### Fig 1B
```{r}

sp2 <- ggplot(dt_plot,
             aes(x = log_len, y = mean_Q )
             ) +
  geom_point_rast(aes(fill = instrument_platform),
             shape = 21, color = "grey20", #color = "transparent", 
             size = 0.35, alpha = 0.4, stroke = 0.4) +
  scale_fill_npg() +
  theme_pubr()


sp <- sp2 +  
  geom_vline(data = platform_summary, aes(xintercept = mean_read_length, colour = instrument_platform), linewidth = 1) +
  #geom_hline(yintercept = mean(platform_summary$min_pass_Qscore), colour = 'grey30', linewidth = 1 ) +
  geom_hline(data = platform_summary, aes(yintercept = mean_Q, colour = instrument_platform), linewidth = 1) +
  
  geom_label_repel(data = platform_summary, aes(label = paste0('Overall mean Q = ', round(mean_Q, 3)), fill = instrument_platform, y = mean_Q, x = 1), color = "grey20") +
  #geom_label_repel(data = platform_summary, aes(label = paste0('Overall mean read length = ', round(mean_read_length, 3)), fill = sample, y = 18, x = log10(n50_len)), color = "grey20") +
  
  labs(#title = paste0(sample_name, ": read length distribution (capped)"),
       x = "Read length ( log10(bp) )", y = "Read mean Q-score", fill = NULL) +
  
  guides(colour = "none") +
  
  theme(legend.position = 'bottom') +
  
  ggpubr::border()                                       


xplot <- ggplot(dt_plot, aes(log_len)) +
  geom_density(aes(fill = instrument_platform), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  theme_pubr() 


yplot <- ggplot(dt_plot, aes(mean_Q)) +
  geom_density(aes(fill = instrument_platform), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  theme_pubr() +
  ggpubr::rotate()

# Arranging the plot using cowplot
p_count_q <- plot_grid(xplot, NULL, 
                       sp, 
                       yplot, 
                       ncol = 2, align = "hv",
                       rel_widths = c(2, 1), rel_heights = c(1, 2))


out_prefix <- "Platform_comparison_QC_Fig1B"

ggsave(paste0(out_dir, '/', out_prefix, '.png'), p_count_q, width = 30, height = 30, bg = 'white')

p_count_q

#save_plot(p_count_q, out_dir, out_prefix, w = 300, h = 300, bg = 'white')
```


### Fig 1C
```{r}

sp2 <- ggplot(dt_plot,
             aes(x = length, y = mean_Q )
             ) +
  geom_point_rast(aes(fill = instrument_platform),
             shape = 21, color = "grey20", #color = "transparent", 
             size = 0.35, alpha = 0.4, stroke = 0.4) +
  scale_fill_npg() +
  scale_x_continuous(limits = c(0, 2000)) +
  theme_pubr()


sp <- sp2 +  
  
  ## Mean length lines and labels
  geom_vline(data = platform_summary, aes(xintercept = mean_read_length, colour = instrument_platform), linewidth = 1) +
  geom_label_repel(data = platform_summary, 
                   aes(label = paste0('Mean read length = ', round(mean_read_length, 0)), fill = instrument_platform, y = 38, x = mean_read_length), color = "black") +
  
  ## Mean Q lines and labels
  geom_hline(data = platform_summary, aes(yintercept = mean_Q, colour = instrument_platform), linewidth = 1) +
  geom_label_repel(data = platform_summary, 
                   aes(label = paste0('Overall mean Q = ', round(mean_Q, 3)), fill = instrument_platform, y = mean_Q, x = 1800), color = "black") +
  
  labs(#title = paste0(sample_name, ": read length distribution (capped)"),
       x = "Read length (bp)", y = "Read mean Q-score", fill = NULL) +
  
  guides(colour = "none") +
  
  theme(legend.position = 'bottom') +
  
  ggpubr::border()                                       


xplot <- ggplot(dt_plot, aes(length)) +
  geom_density(aes(fill = instrument_platform, y = after_stat(scaled)), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  scale_x_continuous(limits = c(0, 2000)) +
  theme_pubr() 


yplot <- ggplot(dt_plot, aes(mean_Q)) +
  geom_density(aes(fill = instrument_platform, y = after_stat(scaled)), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  theme_pubr() +
  ggpubr::rotate()

# Arranging the plot using cowplot
p_count_q <- plot_grid(xplot, NULL, 
                       sp, 
                       yplot, 
                       ncol = 2, align = "hv",
                       rel_widths = c(2, 1), rel_heights = c(1, 2))


out_prefix <- "Platform_comparison_QC_Fig1C"

ggsave(paste0(out_dir, '/', out_prefix, '.png'), p_count_q, width = 30, height = 30, bg = 'white')

p_count_q

saveRDS(p_count_q, paste0(out_dir, '/', out_prefix, '.rds'))

#save_plot(p_count_q, out_dir, out_prefix, w = 300, h = 300, bg = 'white')
# xplot
```


### Fig 1 Final
```{r}
Fig1B <- readRDS(paste0('Article_Figures', '/', 'Figure_3B_ONT_vs_Illumina_WGS_sourmash_Euler.rds') )

# Arranging the plot using cowplot
Fig1 <- plot_grid(xplot, Fig1B, 
                  sp, 
                  yplot, 
                  ncol = 2, align = "hv",
                  rel_widths = c(2, 1), rel_heights = c(1, 2),
                  labels = c('a', 'b', '', ''))

ggsave(ppaste0('Article_Figures', '/', 'Figure_1_Final.png'), Fig1, width = 30, height = 30, bg = 'white')

```



### Fig 1D

```{r}

oplot <- ggplot(dt_plot[
  instrument_platform == 'OXFORD_NANOPORE'
  ], aes(length)) +
  #geom_col(aes(fill = instrument_platform), alpha = 0.5, position = 'dodge') +
  geom_density(aes(fill = instrument_platform), alpha = 0.5) +
  #geom_smooth(aes(color = instrument_platform, group = instrument_platform), alpha = 0.5) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal[-1]) +
  scale_x_continuous(limits = c(0, 5000)) +
  theme_pubr() 


iplot <- ggplot(dt_plot[
  instrument_platform == 'ILLUMINA'
  ], aes(length)) +
  #geom_col(aes(fill = instrument_platform), alpha = 0.5, position = 'dodge') +
  geom_density(aes(fill = instrument_platform), alpha = 0.5) +
  #geom_smooth(aes(color = instrument_platform, group = instrument_platform), alpha = 0.5) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  scale_x_continuous(limits = c(100, 250)) +
  theme_pubr() 


xplot <- cowplot::plot_grid(oplot, iplot, ncol = 2)
xplot


out_prefix <- "Platform_comparison_QC_Fig1E"

ggsave(paste0(out_dir, '/', out_prefix, '.png'), xplot, width = 20, height = 12, bg = 'white')

```



# SertePerti MinION V1-V9 Sequencing Summary

## Import and Process Data
```{r, message=T, eval=F}
files   <- list.files('ENA/PRJEB82097/.fastq_read_stats', recursive = T, full.names = T)

i <- 1

file <- files[i]

mapfun <- function(i, files) { 
  file <- files[i]
  sample_name <- gsub('.meanQ.tsv', '', gsub('ENA/PRJEB82097/.fastq_read_stats/', '', file) )
  dt <- fread(file)
  dt[, sample := sample_name]
  return(dt)
}

proj_stats_dt <- rbindlist( purrr::map(1:length(files), mapfun, files) )

fwrite(proj_stats_dt[,.(sample, read_id, length, mean_Q)], 'ENA/PRJEB82097/PRJEB82097_stats.tsv.gz', sep = '\t')
```


```{r, message=T, eval=F}
proj_stats_dt <- fread('ENA/PRJEB82097/PRJEB82097_stats.tsv.gz')

proj_stats_dt[, project := 'PRJEB82097']

ena_mt <- fread('ENA/PRJEB82097/filereport_read_run_PRJEB82097.tsv')

PRJEB82097_st <- merge(proj_stats_dt, 
                       ena_mt[,.(run_accession, study_accession, sample_accession, secondary_sample_accession, sample_title)], 
                       by.x = c('sample', 'project'), by.y = c('run_accession', 'study_accession')
                       , all.x = T
                       )

if(nrow(proj_stats_dt) == nrow(PRJEB82097_st)) {
  message('ok! every sample matches with the metadata!')
  PRJEB82097_stats <- PRJEB82097_st; rm(PRJEB82097_st)
} else {
  stop('ERROR! Some samples have not been downloaded and/or imported!')
}

```

```{r, message=T, eval=F}
# the ENA files contain samples that are not published in this article

# so we filter them to the actual 40 dogs sample

SuppTab_S4 <- fread('2_40dogs_ZHMW_vs_ZMB/Supplementary Table S4.txt')

SuppTab_S4 <- SuppTab_S4[!is.na(Sample) & Sample != '']

PRJEB82097_st  <- merge(PRJEB82097_stats, 
                        SuppTab_S4, 
                        by.x = 'secondary_sample_accession', 
                        by.y = 'Accession'
                        , all=T)

setnames(PRJEB82097_st, c('Sample', 'sample'), c('sample', 'biosample'))

PRJEB82097_meta <- PRJEB82097_st[,.(read_count=.N), by = .(sample, biosample,  DNA_isolation_method, project, sample_accession, secondary_sample_accession, Sex, Date, Sample_name, Dog, `Read count`)]

PRJEB82097_meta[,.N,is.na(sample)]

PRJEB82097_st[!is.na(sample),.N, sample]

PRJEB82097_st  <- PRJEB82097_st[!is.na(sample)]

## ok!
if(all(is.element(PRJEB82097_st$secondary_sample_accession, SuppTab_S4$Accession))) {
  message('ok! every sample matches with the metadata!')
  PRJEB82097_stats <- PRJEB82097_st; rm(PRJEB82097_st)
} else {
  stop('ERROR! Some samples have not been downloaded and/or imported!')
}

PRJEB82097_meta <- PRJEB82097_stats[,.(read_count=.N), by = .(sample, biosample,  DNA_isolation_method, project, sample_accession, secondary_sample_accession, Sex, Date, Sample_name, Dog, `Read count`)]

```


```{r, eval=F}
### Copy these files to another folder for EMU re-run
dir.create('D:/ENA/PRJEB82097/.fastq_other')

files   <- list.files('D:/ENA/PRJEB82097/.fastq', recursive = T, full.names = T)

## v1
# from    <- grep(paste0(PRJEB82097_meta$biosample, collapse = '|'), files, value = T, invert = T)


## v2
from    <- paste0('D:/ENA/PRJEB82097/.fastq/', PRJEB82097_meta$biosample, '.fastq.gz')

to      <- gsub(".fastq", ".fastq_other", from)

file.copy(from, to)

to2 <- gsub(".fastq_other.gz", ".fastq.gz", to)

file.rename(to, to2 )

file.rename('D:/ENA/PRJEB82097/.fastq_other', 'D:/ENA/PRJEB82097/.fastq_keep')



##
to      <- list.files('D:/ENA/PRJEB82097/.fastq_emu', recursive = T, full.names = T)

to      <- grep('_rel-abundance.tsv', to, value = T)

to      <- gsub('.*\\/', '', to)

to      <- gsub('_rel-abundance.tsv', '', to)

to3     <- grep(paste0(to, collapse = '|'), to2, value = T)

file.remove(to3)



from     <- grep(paste0(to, collapse = '|'), files, invert = T, value = T)
to4      <- gsub('/.fastq/', '/.fastq_other/', from)

file.copy(from, to4)
```


```{r, eval=F}
miss <- PRJEB85714_stats[is.na(read_id), .N, sample][,sample]

cmd <- fread('D:/ENA/PRJEB82097/ena-file-download-read_run-PRJEB82097-fastq_ftp-20260224-1621.sh', header = F)

cmd[, sample_name := gsub('.*\\/', '', gsub('.fastq.gz', '', V3) ) ]
  
cmd <- cmd[sample_name %in% miss, ]

fwrite(cmd, eol = '\n', col.names = F,
       'D:/ENA/PRJEB82097/ena-file-download-read_run-PRJEB82097-fastq_ftp-20260224-1621_miss.sh', 
       sep = ' ')

```


```{r}
PRJEB82097_stats_uni <- PRJEB82097_stats[,.(sample, read_id, length, mean_Q)]

fwrite(PRJEB82097_meta, '2_40dogs_ZHMW_vs_ZMB/PRJEB82097_meta.tsv', sep = '\t')

fwrite(PRJEB82097_stats_uni, '2_40dogs_ZHMW_vs_ZMB/PRJEB82097_stats_uni.tsv.gz', sep = '\t')
```

## Import Prepared Data
```{r}

PRJEB82097_meta      <- fread('2_40dogs_ZHMW_vs_ZMB/PRJEB82097_meta.tsv')

PRJEB82097_stats_uni <- fread('2_40dogs_ZHMW_vs_ZMB/PRJEB82097_stats_uni.tsv.gz')


PRJEB82097_stats <- merge(PRJEB82097_stats_uni, PRJEB82097_meta, by = 'sample')

if(nrow(PRJEB82097_stats) == nrow(PRJEB82097_stats_uni)) {
  rm(PRJEB82097_stats_uni)
} else {
  stop()
}
```



## Summary
```{r}
library(data.table)

# Make sure it's data.table
setDT(PRJEB82097_stats)

# Keep only rows with defined isolation method
dt <- PRJEB82097_stats[!is.na(DNA_isolation_method)]

# Ensure numeric
dt[, length := as.numeric(length)]
dt[, mean_Q := as.numeric(mean_Q)]

# ---- N50 function
n50 <- function(x) {
  x <- x[is.finite(x) & x > 0]
  x <- sort(x, decreasing = TRUE)
  x[which(cumsum(x) >= sum(x)/2)[1]]
}

# =========================
# 1️⃣ Per-sample summary
# =========================

sample_summary <- dt[, .(
  n_reads = .N,
  total_bases = sum(length, na.rm = TRUE),
  mean_len = mean(length, na.rm = TRUE),
  median_len = median(length, na.rm = TRUE),
  n50_len = n50(length),
  max_len = max(length),
  min_len = min(length),
  mean_Q = mean(mean_Q, na.rm = TRUE),
  median_Q = median(mean_Q, na.rm = TRUE)
), by = .(sample, DNA_isolation_method)]

# Save
fwrite(sample_summary, "2_40dogs_ZHMW_vs_ZMB/Serteperti_sample_QC_summary.tsv", sep = '\t')

# =========================
# 2️⃣ Per-method summary (across samples)
# =========================

method_summary <- sample_summary[, .(
  n_samples = .N,
  mean_reads = mean(n_reads),
  sd_reads = sd(n_reads),
  mean_total_bases = mean(total_bases),
  sd_total_bases = sd(total_bases),
  mean_read_length = mean(mean_len),
  sd_read_length = sd(mean_len),
  mean_max_read_len = mean(max_len),
  mean_min_read_len = mean(min_len),
  mean_Q = mean(mean_Q),
  sd_Q = sd(mean_Q)
), by = DNA_isolation_method]

fwrite(method_summary, "2_40dogs_ZHMW_vs_ZMB/Serteperti_method_QC_summary.tsv", sep = '\t')

sample_summary
method_summary

```

```{r, eval=F}
# Publication-ready multi-panel QC figure:
# Compare read length + read quality between ZHMW vs ZMB
#  (A) group-level distribution of per-sample mean read length
#  (B) group-level distribution of per-sample mean Q
#  (C) paired comparison (same biological sample extracted with both methods) using ggpaired
#
# Inputs:
#  - PRJEB82097_stats (per-read table) OR, preferably,
#  - sample_summary (per-sample summary table you already computed)
#
# Output:
#  - Serteperti_ZHMW_vs_ZMB_QC_Fig.pdf + .png

suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(ggpubr)     # ggpaired
  library(patchwork)  # combine panels
  library(scales)
})

# -----------------------------
# 0) Use sample_summary if it exists; otherwise build it quickly
# -----------------------------
if (!exists("sample_summary")) {
  stop("sample_summary not found. Please run your aggregation step first (per-sample QC).")
}

setDT(sample_summary)

# Expect columns:
# sample, dna.isolation.methods, n_reads, total_bases, mean_len, median_len, n50_len, mean_Q, median_Q

# -----------------------------
# 1) Create a pairing ID (strip the method suffix from sample name)
# -----------------------------
sample_summary[, method := dna.isolation.methods]
sample_summary[, pair_id := gsub("_(ZHMW|ZMB)$", "", sample)]

# Keep only pairs that exist in BOTH methods
pair_counts <- sample_summary[, .N, by = .(pair_id)]
paired_ids <- pair_counts[N == 2, pair_id]
ss <- sample_summary[pair_id %in% paired_ids]

# Ensure consistent ordering
ss[, method := factor(method, levels = c("ZHMW", "ZMB"))]


# -----------------------------
# 3) Panel A: group-level mean read length distribution (per sample)
# -----------------------------
pA <- ggplot(ss, aes(x = method, y = mean_len)) +
  geom_violin(trim = TRUE, alpha = 0.6, width = 0.85) +
  geom_boxplot(width = 0.18, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.08, height = 0), size = 1.2, alpha = 0.8) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Read length (per-sample means)",
    x = NULL,
    y = "Mean read length (bp)"
  ) +
  theme_pub()

# -----------------------------
# 4) Panel B: group-level mean Q distribution (per sample)
# -----------------------------
pB <- ggplot(ss, aes(x = method, y = mean_Q)) +
  geom_violin(trim = TRUE, alpha = 0.6, width = 0.85) +
  geom_boxplot(width = 0.18, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.08, height = 0), size = 1.2, alpha = 0.8) +
  labs(
    title = "Read quality (per-sample means)",
    x = NULL,
    y = "Mean Q-score"
  ) +
  theme_pub()

# -----------------------------
# 5) Panel C: paired comparison using ggpaired
#    We'll do BOTH metrics in one panel via long-format + facet
# -----------------------------
long <- melt(
  ss,
  id.vars = c("pair_id", "method"),
  measure.vars = c("mean_len", "mean_Q"),
  variable.name = "metric",
  value.name = "value"
)

long[, metric := factor(metric,
                        levels = c("mean_len", "mean_Q"),
                        labels = c("Mean read length (bp)", "Mean Q-score"))]

pC <- ggpaired(
  long,
  x = "method",
  y = "value",
  id = "pair_id",
  line.color = "grey60",
  line.size = 0.4,
  point.size = 1.4
) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +
  labs(
    title = "Paired comparison (same sample, two extraction methods)",
    x = NULL,
    y = NULL
  ) +
  theme_pub() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold"),
    legend.position = "none"
  )

# -----------------------------
# 6) Assemble into a 3-panel figure
# -----------------------------
# Layout: A | B on top, C across bottom
fig <- (pA | pB) / pC +
  plot_annotation(tag_levels = "A")

# -----------------------------
# 7) Save
# -----------------------------
out_prefix <- "Serteperti_ZHMW_vs_ZMB_QC_Fig"
ggsave(paste0(out_prefix, ".pdf"), fig, width = 180, height = 160, units = "mm", device = cairo_pdf)
ggsave(paste0(out_prefix, ".png"), fig, width = 180, height = 160, units = "mm", dpi = 600)

# Print to viewer
fig

```





```{r}
out_dir <- '2_40dogs_ZHMW_vs_ZMB'
dt_plot <- copy(dt)

library(ggpubr)
library(cowplot)
library(ggrastr)
library(scales)

my_pal <- pal_npg()(10)[-c(1:2)]

dt_plot[, log_len := log10(length)]

dt_plot <- dt_plot[sample(.N)]  # randomize draw order

gc()
```


## Number of reads below and above threshold lengths
```{r, fig.height=25, fig.width=9}
min_len <- 1000
max_len <- 2000

#dt_plot[length >= min_len & length <= max_len, .N, by = .(sample)]

dt_summary <- dt_plot[, .N, by = .(
  sample, DNA_isolation_method,
  category = fifelse(length < min_len, "below_min",
              fifelse(length > max_len, "above_max", "normal"))
)]

dt_summary[, ratio := N / sum(N), by = .(sample, DNA_isolation_method)]

dcast(dt_summary, sample ~ category, value.var = "N", fill = 0)

dcast(dt_summary, sample ~ category, value.var = "ratio", fill = 0)

fwrite(dt_summary, paste0(out_dir, '/', out_dir, '_length_summary.tsv'), sep = '\t')
```


## Plots

### Distribution of Read Lengths (0-2500 bp)
```{r, fig.width = 18, fig.height = 9}

  
xplot <- ggplot(dt_plot, aes(length)) +
  geom_density(aes(fill = DNA_isolation_method), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  scale_x_continuous(labels = label_comma(), limits = c(0, 2500)) +
  theme_pubr() # + facet_nested_wrap(~sample_ID, ncol=10)


xplot2 <- ggplot(dt_plot, aes(length)) +
  geom_density(aes(fill = DNA_isolation_method), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  scale_x_continuous(labels = label_comma()) + #, limits = c(0, 2500)) +
  theme_pubr() # + facet_nested_wrap(~sample_ID, ncol=10)


plots     <- cowplot::plot_grid(xplot, xplot2, align = 'hv', axis = 'tlbr', nrow = 1)

ggsave(paste0(out_dir, '/', out_dir, '_length_distribution.png'), plots, width = 18, height = 9, bg = 'white' )

plots

```


```{r, fig.width = 20, fig.height = 15}

plotfun <- function(sample_to_plot, dt_plot) {
  
  dt_sub <- dt_plot[Dog == sample_to_plot, ]
  
  xplot <- ggplot(dt_sub, aes(length)) +
    geom_density(aes(fill = DNA_isolation_method), alpha = 0.5, n = 128) +
    #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
    scale_fill_manual(values = my_pal) +
    scale_x_continuous(labels = label_comma(), limits = c(0, 2500)) +
    theme_pubr() # + facet_nested_wrap(~sample_ID, ncol=10)
  
}



samples_to_plot <- unique(dt_plot$Dog)
sample_to_plot  <- samples_to_plot[1]

plot_list <- purrr::map(samples_to_plot[], plotfun, dt_plot)

plots     <- cowplot::plot_grid(plotlist = plot_list, align = 'hv', axis = 'tlbr', ncol = 4)

ggsave(paste0(out_dir, '/', out_dir, '_length_distribution_perDog.png'), plots, width = 24, height = 16, bg = 'white' )

plots


```


### Distribution of Sample's Values
```{r}

setDT(sample_summary)

# Create pairing ID
sample_summary[, method := DNA_isolation_method]
sample_summary[, pair_id := gsub("_(ZHMW|ZMB)$", "", sample)]

# Keep only fully paired samples
paired_ids <- sample_summary[, .N, by = pair_id][N == 2, pair_id]
ss <- sample_summary[pair_id %in% paired_ids]

# Ensure order
ss[, method := factor(method, levels = c("ZHMW", "ZMB"))]

theme_pub <- function() {
  theme_pubr(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold"),
      axis.text.x = element_text(face = "bold"),
      legend.position = "none"
    )
}

# Plot
p_reads <- ggplot(ss, aes(x = method, y = n_reads, fill = method)) +
  
  # Violin
  #geom_violin(trim = TRUE, alpha = 0.4, width = 0.85) +
  scale_fill_manual(values = my_pal[]) +
  # Boxplot
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.5) +
  
  # Paired lines
  geom_line(aes(group = pair_id),
            color = "grey60",
            linewidth = 0.4,
            alpha = 0.7) +
  
  # Points
  geom_point(size = 2.5, shape=21, color = 'grey20') +
  
  scale_y_continuous(labels = label_comma()) +
  
  labs(
    #title = "Read count comparison between extraction methods",
    x = NULL,
    y = "Read count per sample"
  ) +
  
  theme_pub()

p_reads


p_q <- ggplot(ss, aes(x = method, y = mean_Q, fill = method)) +
  # Violin
  #geom_violin(trim = TRUE, alpha = 0.4, width = 0.85) +
  scale_fill_manual(values = my_pal[]) +
  # Box
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.5) +
  # Paired lines
  geom_line(aes(group = pair_id),
            color = "grey60",
            linewidth = 0.4,
            alpha = 0.7) +
  # Points
  geom_point(size = 2.5, shape=21, color = 'grey20') +
  labs(
    #title = "Read quality comparison between extraction methods",
    x = NULL,
    y = "Mean Q-score per sample"
  ) +
  theme_pub()

p_q
```


```{r, eval=T}
# Layout: A | B on top, C across bottom
fig <- cowplot::plot_grid(p_reads, p_q, rel_widths = c(1,1), labels = c('a', 'b'))

out_dir    <- "2_40dogs_ZHMW_vs_ZMB"
out_prefix <- "Serteperti_ZHMW_vs_ZMB_QC_Fig"

save_plot(fig, out_dir, out_prefix, w = 350, h = 250)

# Print to viewer
fig

saveRDS(fig, '2_40dogs_ZHMW_vs_ZMB/Article Figures/Figure_2A-B.rds')
```



```{r}
sp <- ggplot(sample_summary, aes(n_reads, mean_Q)) +
  geom_point(aes(fill = DNA_isolation_method), alpha = 0.8, shape = 21, color = 'black', size = 3) +
  scale_fill_manual(values = my_pal[-c(1:2)]) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    #title = "Read quality comparison between extraction methods",
    x = "Read count per sample",
    y = "Mean Q-score per sample"
  ) +
  theme_pub()


xplot <- ggplot(sample_summary, aes(n_reads)) +
  geom_density(aes(fill = DNA_isolation_method), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal[-c(1:2)]) +
  theme_pubr() 


yplot <- ggplot(dt_plot, aes(mean_Q)) +
  geom_density(aes(fill = DNA_isolation_method), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal[-c(1:2)]) +
  theme_pubr() +
  ggpubr::rotate()


# Cleaning the plots
#sp <- sp #+ rremove("legend")
yplot <- yplot + clean_theme() + rremove("legend")
xplot <- xplot + clean_theme() + rremove("legend")



# Arranging the plot using cowplot
fig <- plot_grid(plot_grid(p_reads + ggtitle(NULL), xplot + ggpubr::rotate(), nrow = 1, align = "hv", rel_widths = c(2,1)), 
                 NULL,
                 plot_grid(p_q     + ggtitle(NULL), yplot, nrow = 1, align = "hv", rel_widths = c(2,1)), 
                 nrow = 1, align = "hv",
                 rel_widths = c(3,1,3), 
                 labels = c('A', '', 'B'))

out_dir    <- "2_40dogs_ZHMW_vs_ZMB"
out_prefix <- "Serteperti_ZHMW_vs_ZMB_QC_Fig1"

save_plot(fig, out_dir, out_prefix, w = 500, h = 200, bg = 'white')

# Print to viewer
fig




# Cleaning the plots
sp <- sp + theme(legend.position = 'bottom') #+ rremove("legend")
yplot <- yplot + clean_theme() + rremove("legend")
xplot <- xplot + clean_theme() + rremove("legend")

# Arranging the plot using cowplot
p_count_q <- plot_grid(xplot, NULL, 
                       sp, yplot, 
                       ncol = 2, align = "hv",
                       rel_widths = c(2, 1), rel_heights = c(1, 2))


p_count_q

out_prefix <- "Serteperti_ZHMW_vs_ZMB_QC_Fig2A"

save_plot(p_count_q, out_dir, out_prefix, w = 300, h = 300, bg = 'white')

```

### Distribution of all reads
```{r}

sp <- ggplot(dt_plot, aes(length, mean_Q)) +
  geom_point_rast(aes(fill = DNA_isolation_method), alpha = 0.8, shape = 21, color = 'black', size = 1) +
  scale_fill_manual(values = my_pal) +
  scale_x_continuous(labels = label_comma(), limits = c(0,2000)) +
  labs(
    #title = "Read quality comparison between extraction methods",
    x = "Read Length (bp)",
    y = "Mean Q-score"
  ) +
  theme_pub()


xplot <- ggplot(dt_plot, aes(length)) +
  geom_density(aes(fill = DNA_isolation_method), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  scale_x_continuous(labels = label_comma(), limits = c(0,2000)) +
  theme_pubr() 


yplot <- ggplot(dt_plot, aes(mean_Q)) +
  geom_density(aes(fill = DNA_isolation_method), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  theme_pubr() +
  ggpubr::rotate()



# Cleaning the plots
sp <- sp + theme(legend.position = 'bottom') #+ rremove("legend")
yplot <- yplot + clean_theme() + rremove("legend")
xplot <- xplot + clean_theme() + rremove("legend")

# Arranging the plot using cowplot
p_count_q <- plot_grid(xplot, NULL, 
                       sp, yplot, 
                       ncol = 2, align = "hv",
                       rel_widths = c(2, 1), rel_heights = c(1, 2))


p_count_q

out_prefix <- "2_40dogs_ZHMW_vs_ZMB_QC_Fig2B"

save_plot(p_count_q, out_dir, out_prefix, w = 900, h = 600, bg = 'white')

```
```{r}

sp <- ggplot(dt_plot, aes(length, mean_Q)) +
  geom_point_rast(aes(fill = DNA_isolation_method), alpha = 0.8, shape = 21, color = 'black', size = 1) +
  scale_fill_manual(values = my_pal) +
  #scale_x_continuous(labels = label_comma(), limits = c(0,2000)) +
  labs(
    #title = "Read quality comparison between extraction methods",
    x = "Read Length (bp)",
    y = "Mean Q-score"
  ) +
  theme_pub()


xplot <- ggplot(dt_plot, aes(length)) +
  geom_density(aes(fill = DNA_isolation_method), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  #scale_x_continuous(labels = label_comma(), limits = c(0,2000)) +
  theme_pubr() 


yplot <- ggplot(dt_plot, aes(mean_Q)) +
  geom_density(aes(fill = DNA_isolation_method), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  theme_pubr() +
  ggpubr::rotate()



# Cleaning the plots
sp <- sp + theme(legend.position = 'bottom') #+ rremove("legend")
yplot <- yplot + clean_theme() + rremove("legend")
xplot <- xplot + clean_theme() + rremove("legend")

# Arranging the plot using cowplot
p_count_q <- plot_grid(xplot, NULL, 
                       sp, yplot, 
                       ncol = 2, align = "hv",
                       rel_widths = c(2, 1), rel_heights = c(1, 2))


p_count_q

out_prefix <- "2_40dogs_ZHMW_vs_ZMB_QC_Fig2D"

save_plot(p_count_q, out_dir, out_prefix, w = 900, h = 600, bg = 'white')

```


### Distribution of all reads faceted
```{r, fig.height = 40, fig.width = 40}
dir.create('2_40dogs_ZHMW_vs_ZMB/Article Figures/suppfig')

sample_to_plot <- 'M1_210317'

plotfun <- function(sample_to_plot, dt_plot) {
    
  dt_sub <- dt_plot[Sample_name == sample_to_plot, ]
  
  sp <- ggplot(dt_sub, aes(length, mean_Q)) +
    geom_point_rast(aes(fill = DNA_isolation_method), alpha = 0.8, shape = 21, color = 'black', size = 1) +
    scale_fill_manual(values = my_pal) +
    scale_x_continuous(labels = label_comma(), limits = c(0,2000)) +
    labs(
      #title = "Read quality comparison between extraction methods",
      x = "Read Length (bp)",
      y = "Mean Q-score"
    ) +
    theme_pub()
  
  
  xplot <- ggplot(dt_sub, aes(length)) +
    geom_density(aes(fill = DNA_isolation_method), alpha = 0.5, n = 128) +
    #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
    scale_fill_manual(values = my_pal) +
    scale_x_continuous(labels = label_comma(), limits = c(0,2000)) +
    theme_pubr() 
  
  
  yplot <- ggplot(dt_sub, aes(mean_Q)) +
    geom_density(aes(fill = DNA_isolation_method), alpha = 0.5, n = 128) +
    #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
    scale_fill_manual(values = my_pal) +
    theme_pubr() +
    ggpubr::rotate()
  
  
  
  # Cleaning the plots
  sp <- sp + theme(legend.position = 'bottom') #+ rremove("legend")
  yplot <- yplot + clean_theme() + rremove("legend")
  xplot <- xplot + clean_theme() + rremove("legend")
  
  # Arranging the plot using cowplot
  p_count_q <- plot_grid(xplot, NULL, 
                         sp, yplot, 
                         ncol = 2, align = "hv",
                         rel_widths = c(2, 1), rel_heights = c(1, 2))
  
  ggsave(paste0('2_40dogs_ZHMW_vs_ZMB/Article Figures/suppfig/', sample_to_plot, '.png'),
         p_count_q, width = 9, height = 6, bg = 'white')
  
  return(p_count_q)
  
  
}

samples_to_plot <- unique(PRJEB82097_meta$Sample_name)

plot_list <- purrr::map(samples_to_plot[], plotfun, dt_plot, .progress = T)

plots <- cowplot::plot_grid(plotlist = plot_list, align = 'hv', axis = 'tlbr', ncol = 8)

out_prefix <- "2_40dogs_ZHMW_vs_ZMB_QC_Fig2C"

save_plot(plots, out_dir, out_prefix, w = 1500, h = 1800, bg = 'white')

plots
```



# Primeres 


## PRJEB85714 (human samples)

```{r}

files   <- list.files('ENA/PRJEB85714/.fastq_read_stats', recursive = T, full.names = T)

i <- 1

file <- files[i]

mapfun <- function(i, files) { 
  file <- files[i]
  sample_name <- gsub('.meanQ.tsv', '', gsub('ENA/PRJEB85714/.fastq_read_stats/', '', file) )
  dt <- fread(file)
  dt[, sample := sample_name]
  return(dt)
}

proj_stats_dt <- rbindlist( purrr::map(1:length(files), mapfun, files) )

fwrite(proj_stats_dt[,.(sample, read_id, length, mean_Q)], 'ENA/PRJEB85714/PRJEB85714_stats.tsv.gz', sep = '\t')
```


```{r, message=T}
proj_stats_dt <- fread('ENA/PRJEB85714/PRJEB85714_stats.tsv.gz')

proj_stats_dt[, project := 'PRJEB85714']

ena_mt <- fread('ENA/PRJEB85714/filereport_read_run_PRJEB85714.tsv')

PRJEB85714_st <- merge(proj_stats_dt, 
                       ena_mt[,.(run_accession, study_accession, sample_accession, secondary_sample_accession, sample_title)], 
                       by.x = c('sample', 'project'), by.y = c('run_accession', 'study_accession')
                       #, all = T
                       )

if(nrow(proj_stats_dt) == nrow(PRJEB85714_st)) {
  message('ok! every sample matches with the metadata!')
  PRJEB85714_stats <- PRJEB85714_st; rm(PRJEB85420_st)
} else {
  stop('ERROR! Some samples have not been downloaded and/or imported!')
}

```

```{r, message=T}
PRJEB85714_meta <- fread('3_Primer_comparison/metadata.tsv')

## trim off date and barcode
PRJEB85714_meta[, sample  := gsub('_bc.*', '', filename)]
PRJEB85714_meta[, sample  := gsub('_20.*', '', sample  )]

PRJEB85714_stats[, sample_title  := gsub('_20.*', '', sample_title  )]

## merge with metadata
PRJEB85714_st <- merge(PRJEB85714_stats, 
                       PRJEB85714_meta, 
                       by.x = 'sample_title', by.y = 'sample',
                       all.x=T)


if(nrow(PRJEB85714_stats) == nrow(PRJEB85714_st)) {
  message('ok! every sample matches with the metadata!')
  PRJEB85714_stats <- PRJEB85714_st; rm(PRJEB85714_st)
} else {
  stop('ERROR! Some samples have not been downloaded and/or imported!')
}

setnames(PRJEB85714_stats, c('sample_title', 'sample'), c('sample', 'biosample'))

```



## PRJEB85420 (synthetic and canine samples) 

```{r}

files   <- list.files('ENA/PRJEB85420/.fastq_read_stats', recursive = T, full.names = T)

i <- 1

file <- files[i]

mapfun <- function(i, files) { 
  file <- files[i]
  sample_name <- gsub('.meanQ.tsv', '', gsub('ENA/PRJEB85420/.fastq_read_stats/', '', file) )
  dt <- fread(file)
  dt[, sample := sample_name]
  return(dt)
}

proj_stats_dt <- rbindlist( purrr::map(1:length(files), mapfun, files) )

fwrite(proj_stats_dt[,.(sample, read_id, length, mean_Q)], 'ENA/PRJEB85420/PRJEB85420_stats.tsv.gz', sep = '\t')
```


```{r, message=T}
proj_stats_dt <- fread('ENA/PRJEB85420/PRJEB85420_stats.tsv.gz')

proj_stats_dt[, project := 'PRJEB85420']

ena_mt <- fread('ENA/PRJEB85420/filereport_read_run_PRJEB85420.tsv')

PRJEB85420_st <- merge(proj_stats_dt, 
                       ena_mt[,.(run_accession, study_accession, sample_accession, secondary_sample_accession, sample_title)], 
                       by.x = c('sample', 'project'), by.y = c('run_accession', 'study_accession')
                       #, all = T
                       )

if(nrow(proj_stats_dt) == nrow(PRJEB85420_st)) {
  message('ok! every sample matches with the metadata!')
  PRJEB85420_stats <- PRJEB85420_st; rm(PRJEB85420_st)
} else {
  stop('ERROR! Some samples have not been downloaded and/or imported!')
}

```

```{r, message=T}
PRJEB85420_meta <- fread('3_Primer_comparison/metadata.tsv')

## trim off date and barcode
PRJEB85420_meta[, sample  := gsub('_bc.*', '', filename)]
PRJEB85420_meta[, sample  := gsub('_20.*', '', sample  )]

PRJEB85420_stats[, sample_title  := gsub('_20.*', '', sample_title  )]

## merge with metadata
PRJEB85420_st <- merge(PRJEB85420_stats, 
                       PRJEB85420_meta, 
                       by.x = 'sample_title', by.y = 'sample',
                       all.x=T)


if(nrow(PRJEB85420_stats) == nrow(PRJEB85420_st)) {
  message('ok! every sample is downloaded and imported!')
  PRJEB85420_stats <- PRJEB85420_st; rm(PRJEB85420_st)
} else {
  stop('ERROR! Some samples have not been downloaded and/or imported!')
}


setnames(PRJEB85420_stats, c('sample_title', 'sample'), c('sample', 'biosample'))


```


## Combine

```{r}
all_stats_dt <- rbind( PRJEB85420_stats, PRJEB85714_stats )

all_stats_dt[,.N, project]

all_stats_uni <- all_stats_dt[,.(sample, source, primer, rep, name, project, read_id, length, mean_Q)]

rm(all_stats_dt, PRJEB85420_stats, PRJEB85714_stats )

gc()


proj_meta <- unique(all_stats_uni[,.(sample, source, primer, rep, name, project)])

all_stats_uni <- all_stats_uni[,.(sample, read_id, length, mean_Q)]

fwrite(proj_meta, '3_Primer_comparison/proj_meta.tsv.gz', sep = '\t')

fwrite(all_stats_uni, '3_Primer_comparison/all_stats_uni.tsv.gz', sep = '\t')

```




```{r, eval=F}
cmd <- 'split -b 95M 3_Primer_comparison/all_stats_uni.tsv.gz 3_Primer_comparison/all_stats_uni.tsv.gz.part_'
system(cmd)

cmd <- 'rm 3_Primer_comparison/all_stats_uni.tsv.gz'
system(cmd)

```

```{bash}
#cmd <- '
cat 3_Primer_comparison/all_stats_uni.tsv.gz.part_* > 3_Primer_comparison/all_stats_uni.tsv.gz
#'
#system(cmd)


```

```{r}
try({ all_stats_uni <- fread('3_Primer_comparison/all_stats_uni.tsv.gz') })


proj_meta <- fread('3_Primer_comparison/proj_meta.tsv.gz')


all_stats_uni <- merge(all_stats_uni, proj_meta, by = 'sample')
```



## Summary
```{r}
# Make sure it's data.table
setDT(all_stats_uni)

# Keep only rows with defined isolation method
dt <- all_stats_uni[!is.na(source) & !is.na(primer)]

# Ensure numeric
dt[, length := as.numeric(length)]
dt[, mean_Q := as.numeric(mean_Q)]

# ---- N50 function
n50 <- function(x) {
  x <- x[is.finite(x) & x > 0]
  x <- sort(x, decreasing = TRUE)
  x[which(cumsum(x) >= sum(x)/2)[1]]
}

# =========================
# 1️⃣ Per-sample summary
# =========================

sample_summary <- dt[, .(
  n_reads = .N,
  total_bases = sum(length, na.rm = TRUE),
  mean_len = mean(length, na.rm = TRUE),
  median_len = median(length, na.rm = TRUE),
  n50_len = n50(length),
  max_len = max(length),
  min_len = min(length),
  mean_Q = mean(mean_Q, na.rm = TRUE),
  median_Q = median(mean_Q, na.rm = TRUE)
), by = .(sample, source, primer)]


# =========================
# 2️⃣ Per-source summary (across samples)
# =========================

source_summary <- sample_summary[, .(
  n_samples = .N,
  mean_reads = mean(n_reads),
  sd_reads = sd(n_reads),
  mean_total_bases = mean(total_bases),
  sd_total_bases = sd(total_bases),
  mean_read_length = mean(mean_len),
  sd_read_length = sd(mean_len),
  mean_max_read_len = mean(max_len),
  mean_min_read_len = mean(min_len),
  mean_Q = mean(mean_Q),
  sd_Q = sd(mean_Q)
), by = .(source, primer)]

# =========================
# 3️⃣ Per-length and sample summay
# =========================

length_summary <- dt[, .(
  n_reads = .N
), by = .(length ,sample, source, primer)]

# Save
fwrite(sample_summary, "3_Primer_comparison/Primer_sample_QC_summary.tsv", sep = '\t')

fwrite(source_summary, "3_Primer_comparison/Primer_source_QC_summary.tsv", sep = '\t')

fwrite(length_summary, "3_Primer_comparison/Primer_length_QC_summary.tsv", sep = '\t')


sample_summary
source_summary
length_summary
```

### Finding the valley for length filtering

EMU filters for 0-2000 bp by default, so the reads around the lower peak (~900 bp ) are *not filtered out*  
-->> we must change the EMU's parameter or *filter manually*

```{r}
min_search <- 1200
max_search <- 1550

min_len_by_sample <- length_summary[
  length >= min_search & length <= max_search,
  .SD[which.min(n_reads)],
  by = sample
][, .(sample, best_min_len = length, reads_at_valley = n_reads)]

min(min_len_by_sample$best_min_len)
max(min_len_by_sample$best_min_len)
mean(min_len_by_sample$best_min_len)

min_search <- 1700
max_search <- 2000

max_len_by_sample <- length_summary[
  length >= min_search & length <= max_search,
  .SD[which.min(n_reads)],
  by = sample
][, .(sample, best_min_len = length, reads_at_valley = n_reads)]

min( max_len_by_sample$best_min_len)
max( max_len_by_sample$best_min_len)
mean(max_len_by_sample$best_min_len)


```

Filtering : 1500 - 1800 bp


```{r}
out_dir <- '3_Primer_comparison'
dt_plot <- copy(all_stats_uni)

dt_plot[, log_len := log10(length)]

dt_plot <- dt_plot[sample(.N)]  # randomize draw order if you keep any point layers

library(ggpubr)
library(cowplot)
library(ggrastr)
library(scales)

my_pal <- pal_npg()(10)[-c(1:4)]
gc()

```


## Number of reads below and above threshold lengths
```{r, fig.height=25, fig.width=9}
min_len <- 1000
max_len <- 2000

#dt_plot[length >= min_len & length <= max_len, .N, by = .(sample)]

dt_summary <- dt_plot[, .N, by = .(
  sample,
  category = fifelse(length < min_len, "below_min",
              fifelse(length > max_len, "above_max", "normal"))
)]

dt_summary[, ratio := N / sum(N), by = sample]

dcast(dt_summary, sample ~ category, value.var = "N", fill = 0)

dcast(dt_summary, sample ~ category, value.var = "ratio", fill = 0)

fwrite(dt_summary, paste0(out_dir, '/', out_dir, '_length_summary.tsv'), sep = '\t')
```




## Plots

```{r, fig.height=25, fig.width=12}
ggp <- cowplot::plot_grid(
  ggplot(dt_summary, aes(x = sample, y = N, fill = category)) +
    geom_col(position = 'identity', color = 'black') +
    coord_flip() +
    scale_fill_npg() +
    theme_ipsum() +
    theme(legend.position = 'bottom')
  ,
  ggplot(dt_summary, aes(x = sample, y = ratio, fill = category)) +
    geom_col(position = 'identity', color = 'black') +
    coord_flip() +
    scale_fill_npg() +
    theme_ipsum() +
    theme(legend.position = 'bottom')
  ,
  nrow = 1, align = 'vh', axis = 'tlbr'
)

ggsave(paste0(out_dir, '/', out_dir, '_length_thresholds.png'), ggp, width = 20, height = 15, bg = 'white' )

ggp
```


### Distribution of Read Lengths (0-2500 bp)
```{r, fig.width = 20, fig.height = 15}

plotfun <- function(sample_to_plot, dt_plot) {
  
  dt_sub <- dt_plot[source == sample_to_plot, ]
  
  xplot <- ggplot(dt_sub, aes(length)) +
    geom_density(aes(fill = primer), alpha = 0.5, n = 128) +
    #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
    scale_fill_manual(values = my_pal) +
    scale_x_continuous(labels = label_comma(), limits = c(0, 2500)) +
    theme_pubr() + # + facet_nested_wrap(~sample_ID, ncol=10)
    ggtitle(sample_to_plot)
}



samples_to_plot <- unique(dt_plot$source)
sample_to_plot  <- samples_to_plot[1]

plot_list <- purrr::map(samples_to_plot[], plotfun, dt_plot)

plots     <- cowplot::plot_grid(plotlist = plot_list, align = 'hv', axis = 'tlbr', ncol = 3)

ggsave(paste0(out_dir, '/', out_dir, '_length_distribution_source.png'), plots, width = 20, height = 15, bg = 'white' )

plots


```


```{r, fig.width = 30, fig.height = 25}



xplot <- ggplot(dt_plot, aes(length)) +
  geom_density(aes(fill = primer), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  scale_x_continuous(labels = label_comma(), limits = c(0, 2500)) +
  theme_pubr() + #
  facet_nested(rows = vars(source), cols = vars(primer, rep))


ggsave(paste0(out_dir, '/', out_dir, '_length_distribution_sample.png'), xplot, width = 30, height = 25, bg = 'white' )

xplot


```



### Distribution of all reads
```{r}


sp <- ggplot(sample_summary, aes(n_reads, mean_Q)) +
  geom_point(aes(fill = source), alpha = 0.8, shape = 21, color = 'black', size = 3) +
  scale_fill_manual(values = my_pal) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    #title = "Read quality comparison between extraction methods",
    x = "Read count per sample",
    y = "Mean Q-score per sample"
  ) +
  theme_pub()


xplot <- ggplot(sample_summary, aes(n_reads)) +
  geom_density(aes(fill = source), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  theme_pubr() 


yplot <- ggplot(dt_plot, aes(mean_Q)) +
  geom_density(aes(fill = source), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  theme_pubr() +
  ggpubr::rotate()



# Cleaning the plots
sp <- sp + theme(legend.position = 'bottom') #+ rremove("legend")
yplot <- yplot + clean_theme() + rremove("legend")
xplot <- xplot + clean_theme() + rremove("legend")

# Arranging the plot using cowplot
p_count_q <- plot_grid(xplot, NULL, 
                       sp, yplot, 
                       ncol = 2, align = "hv",
                       rel_widths = c(2, 1), rel_heights = c(1, 2))


p_count_q

out_prefix <- "Primer_comparison_QC_Fig3A"

save_plot(p_count_q, out_dir, out_prefix, w = 300, h = 300, bg = 'white')

```


```{r}

sp <- ggplot(dt_plot, aes(length, mean_Q)) +
  geom_point_rast(aes(fill = source), alpha = 0.8, shape = 21, color = 'black', size = 1) +
  scale_fill_manual(values = my_pal) +
  scale_x_continuous(labels = label_comma(), limits = c(0,2000)) +
  labs(
    #title = "Read quality comparison between extraction methods",
    x = "Read Length (bp)",
    y = "Mean Q-score"
  ) +
  theme_pub()


xplot <- ggplot(dt_plot, aes(length)) +
  geom_density(aes(fill = source), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  scale_x_continuous(labels = label_comma(), limits = c(0,2000)) +
  theme_pubr() 


yplot <- ggplot(dt_plot, aes(mean_Q)) +
  geom_density(aes(fill = source), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  theme_pubr() +
  ggpubr::rotate()



# Cleaning the plots
sp <- sp + theme(legend.position = 'bottom') #+ rremove("legend")
yplot <- yplot + clean_theme() + rremove("legend")
xplot <- xplot + clean_theme() + rremove("legend")

# Arranging the plot using cowplot
p_count_q <- plot_grid(xplot, NULL, 
                       sp, yplot, 
                       ncol = 2, align = "hv",
                       rel_widths = c(2, 1), rel_heights = c(1, 2))


p_count_q

out_prefix <- "Primer_comparison_QC_Fig3B"

save_plot(p_count_q, out_dir, out_prefix, w = 900, h = 600, bg = 'white')

```


```{r}

sp <- ggplot(dt_plot, aes(length, mean_Q)) +
  geom_point_rast(aes(fill = source), alpha = 0.8, shape = 21, color = 'black', size = 1) +
  scale_fill_manual(values = my_pal) +
  #scale_x_continuous(labels = label_comma(), limits = c(0,2000)) +
  labs(
    #title = "Read quality comparison between extraction methods",
    x = "Read Length (bp)",
    y = "Mean Q-score"
  ) +
  theme_pub()


xplot <- ggplot(dt_plot, aes(length)) +
  geom_density(aes(fill = source), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  #scale_x_continuous(labels = label_comma(), limits = c(0,2000)) +
  theme_pubr() 


yplot <- ggplot(dt_plot, aes(mean_Q)) +
  geom_density(aes(fill = source), alpha = 0.5, n = 128) +
  #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
  scale_fill_manual(values = my_pal) +
  theme_pubr() +
  ggpubr::rotate()



# Cleaning the plots
sp <- sp + theme(legend.position = 'bottom') #+ rremove("legend")
yplot <- yplot + clean_theme() + rremove("legend")
xplot <- xplot + clean_theme() + rremove("legend")

# Arranging the plot using cowplot
p_count_q <- plot_grid(xplot, NULL, 
                       sp, yplot, 
                       ncol = 2, align = "hv",
                       rel_widths = c(2, 1), rel_heights = c(1, 2))


p_count_q

out_prefix <- "Primer_comparison_QC_Fig3D"

save_plot(p_count_q, out_dir, out_prefix, w = 900, h = 600, bg = 'white')

```



### Distribution of all reads faceted
```{r}
dir.create('3_Primer_comparison/Article Figures/suppfig')

plotfun <- function(sample_to_plot, dt_plot) {
    
  dt_sub <- dt_plot[source == sample_to_plot, ]
  
  sp <- ggplot(dt_sub, aes(length, mean_Q)) +
    geom_point_rast(aes(fill = primer), alpha = 0.8, shape = 21, color = 'black', size = 1) +
    scale_fill_manual(values = my_pal) +
    scale_x_continuous(labels = label_comma(), limits = c(0,2000)) +
    labs(
      #title = "Read quality comparison between extraction methods",
      x = "Read Length (bp)",
      y = "Mean Q-score"
    ) +
    theme_pub()
  
  
  xplot <- ggplot(dt_sub, aes(length)) +
    geom_density(aes(fill = primer), alpha = 0.5, n = 128) +
    #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
    scale_fill_manual(values = my_pal) +
    scale_x_continuous(labels = label_comma(), limits = c(0,2000)) +
    theme_pubr() 
  
  
  yplot <- ggplot(dt_sub, aes(mean_Q)) +
    geom_density(aes(fill = primer), alpha = 0.5, n = 128) +
    #geom_histogram(aes(fill = dna.isolation.methods), alpha = 0.5, bins = 100) +
    scale_fill_manual(values = my_pal) +
    theme_pubr() +
    ggpubr::rotate()
  
  
  
  # Cleaning the plots
  sp <- sp + theme(legend.position = 'bottom') #+ rremove("legend")
  yplot <- yplot + clean_theme() + rremove("legend")
  xplot <- xplot + clean_theme() + rremove("legend")
  
  # Arranging the plot using cowplot
  p_count_q <- plot_grid(xplot, NULL, 
                         sp, yplot, 
                         ncol = 2, align = "hv",
                         rel_widths = c(2, 1), rel_heights = c(1, 2))
  
  ggsave(paste0('3_Primer_comparison/Article Figures/suppfig/', sample_to_plot, '.png'),
         p_count_q, width = 9, height = 6, bg = 'white')
  
   return(p_count_q)

}

samples_to_plot <- unique(proj_meta$source)

plot_list <- purrr::map(samples_to_plot, plotfun, dt_plot)

plots     <- cowplot::plot_grid(plotlist = plot_list, align = 'hv', axis = 'tlbr', ncol = 5)

out_prefix <- "Primer_comparison_QC_Fig3C"

save_plot(plots, out_dir, out_prefix, w = 1000, h = 300, bg = 'white')

plots
```


# TO-DO:

## Find out wether the *WGS* peak's taxaonomic distribution is the same as the amplicons ( did the PCR caused bias in the taxonomy )



# END WORKFLOW



