---
title: "Homo sapiens : EMU on EMU database"
author: "Balázs Kakuk"
date: "2024-10-20"
output: html_document
---

```{css, echo=FALSE}
    body .main-container {
      max-width: 90% !important;
      width: 90% !important;
    }
    body {
      max-width: 90% !important;
      margin-left: auto;
      margin-right: auto;
    }
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, fig.align = 'center', fig.height = 10, fig.width = 30, warning = F, message = F, error = F)
library(tidyverse)
library(ggpubr)
library(factoextra)
library(FactoMineR)
library(ggrepel)
library(phyloseq, quietly = T)
library(DESeq2)
library(data.table)
library(ggsci)
library(hrbrthemes)
library(ggh4x)
library(RColorBrewer)
library(stringr); library(stringi)
library("ape")
library("ggtree")
library(Biostrings)
library(dplyr)
library(UpSetR)
library(ComplexHeatmap)
library(genefilter)
library(viridis)
library(vegan)

## Own functions
misc.dir    <- 'C:/GitHub/Rlyeh/R'
minitax.dir <- 'C:/GitHub/minitax/R'

if (.Platform$OS.type!="windows") {
  
  misc.dir  <- paste0('/mnt/', gsub(':', '/', tolower(gsub('/.*', '', misc.dir))),
                      stri_replace_first_regex(misc.dir, '.*:\\/', '')
  )
  
  minitax.dir  <- paste0('/mnt/', gsub(':', '/', tolower(gsub('/.*', '', minitax.dir))),
                         stri_replace_first_regex(minitax.dir, '.*:\\/', '')
  )
  
}

for(f in list.files(misc.dir,    '*.R', full.names = T)) { try({message(f); source(f)}) }
for(f in list.files(minitax.dir, '*.R', full.names = T)) { try({message(f); source(f)}) }

bam.flags           <- fread(paste0(misc.dir, '/bam.flags.tsv'))
gff_compare_classes <- fread(paste0(misc.dir, '/gff_compare.txt'))
nproc <- 32


### General settings

filt.version <- ''
filtering    <- ''
writetables  <- T
save.Figs    <- F
res.dir      <- 'Hsapiens_EMU_EMUdb'; dir.create(res.dir)
#platforms <- c('illumina', 'PacBIO')

## color palette
alpha   <- 0.9
pal.man <- c(brewer.pal(7, 'Set1'), brewer.pal(8, 'Accent')[c(1,2,6)], brewer.pal(8, 'Set2')[c(1,3,5,6,7)], brewer.pal(8, 'Dark2') )
pal.man <- c(colorvec[c(8,32,12:14,17,24,1:2,4,7,5,6,3,10,11,27:28,9,29)], 'grey', 'black')
pal.man <- alpha(c(pal.man, pal.man), alpha)

## taxonomic ranks
ranks <- c("superkingdom", "phylum", "class", "order", "family", "genus", "species")

### minitax settings
## Database
db <- 'EMUdb'
db.dir <- '../databases/EMUdb'
## Genome Size normalization?
NormToGenomeSize <- F
NormFunction     <- 'mean'
## MAPQ filtering ?
mapq.filt <- NA #  1:59 # 
## Filtering samples
filter.samples.pattern <- 'LCA|Undetermined|SpeciesEstimate|BestAln'  ### filter out LCA method


## Exclude unclassified reads?
exclude.unclass <- T

## Omit very low abundance samples
abund.threshold <- 10

## Other settings
consider.euk <- T
count_multiplier <- 100000
count_colname    <- c('16S Only', 'Gram Stain')


### Gold Standard Settings
cami.lineage <- 'db'
add.GS.taxa <- F ## add all GS taxa to the barplots and the correlation plots?
zymo.GS <- read_delim('../standards/zymo_6300_composition.txt')
GS_name <- 'Zymo_6300'
GS_nr   <- 1
GS_version <- 'D6300'
just_GS <- T
vars <- c('platform') # 'GS_name' ('project') # c('DNA_isolation_method', 'db', 'workflow', 'method', 
#vars <- c('Vregion', 'platform', 'db', 'workflow', 'method','DNA_isolation_method') # 'GS_version'
glom_spec <- T

## Gram Strain Info
phyla.gram <- read.delim('../databases/phyla.gram.tsv')[,1:2]
colnames(phyla.gram)[2] <- 'Gram.Stain'

## Exclude Bifidobacteria?
exclude.Bifido <- F
if(exclude.Bifido) {
  zymo.GS <- zymo.GS[!grepl('Bifodobac', zymo.GS[,1]) & !grepl('Bifodobac', zymo.GS[,2]), ]
}

## Keep and plot GS taxa only?
keep.GS.taxa.only <- F

mark.GS.Taxa <- F
merge.non.GS.Taxa <- F


## calculate correlation from top20 or from all taxa?
correl <- 'topN'

## Fix taxa
crop_taxa <- NULL # 'Veillonella rogosae|Prevotella corporis'

genera_to_glom <- data.frame(glom= c('Veillonella', 'Faecalibacterium', 'Prevotella', 'Roseburia'),
                           to  = c('Veillonella rogosae','Faecalibacterium prausnitzii', 'Prevotella corporis', 'Roseburia hominis')) ## NULL
species_to_glom <- data.frame(glom= c('Bacteroides fragilis', 'Bacteroides fragilis CAG:47', 'Bacteroides fragilis_A'),
                           to  = c('Bacteroides fragilis')) ## NULL



abundance_threshold <- 1
filter_func <- filterfun(function(x) sum(x) > abundance_threshold)

```



# Notes

- Unclassified reads *were* excluded
- Taxa with total reads number of 10 or lower were excluded 
- Eukaryotes *were* excluded from the analyis

```{r, Database import}
# altough *Lactobacillus fermentum* was changed to *Limosilactobacillus fermentum* 
  
#### Database import ####

if (db == 'proGcontigs_2') {
  prog.db <- fread(paste0(db.dir, '/proGenomes2.1_specI_lineageNCBI.tab'), header = F)
  colnames(prog.db) <- c("genome", "superkingdom", "phylum", "class", "order", "family", "genus", "species")
  prog.db <- data.frame(taxid=gsub('\\..*', '', prog.db$genome  ), prog.db)
  ranks <- c("superkingdom", "phylum", "class", "order", "family", "genus", "species")
  prog.db$taxid <- as.integer(prog.db$taxid)
  
  prog.db[,ranks]  <- as.data.frame(apply(prog.db[,ranks], 2, function(x) str_replace(as.character(unlist(x)),  pattern = "^\\d+\\s", replacement = "")))
  
  prog.gt <- gather(prog.db, rank, taxon, -c(1:2))
  prog.db.spec.uni <- unique.data.frame(prog.db[,ranks])
  
  db.data <- prog.db
  db.name <- db
  
  prog.db.uni     <- db.data %>% distinct(across(all_of(c('taxid', ranks)))) #unique.data.frame(db.data[,c('taxid', ranks)])
  db.uni.data     <- prog.db.uni

   if (NormToGenomeSize) {
    genome_sizes <- as.data.frame(fread(paste0(db.dir, '/proGenomes2.1_specI_lineageNCBI.genomesize.tab')))
    genome_sizes[genome_sizes == ''] <- NA
    genome_sizes[,ranks]  <- 
      as.data.frame(apply(genome_sizes[,ranks], 2, 
                          function(x) str_replace(as.character(unlist(x)),  pattern = "^\\d+\\s", replacement = "")))
  }
  
  db.uni.ranks <- unique.data.frame(db.uni.data[,ranks])
  
  #### Fix Lactobacillus in proGenomes_2
  db.uni.ranks$genus  [db.uni.ranks$species ==  'Lactobacillus fermentum' ] <- 'Limosilactobacillus'
  db.uni.ranks$species[db.uni.ranks$species ==  'Lactobacillus fermentum' ] <- 'Limosilactobacillus fermentum'

} else if (db=="all_NCBI_genomes") {
  
  #db.data     <- fread(paste0(db.dir, "/NCBI.db.tsv"), header = T)
  db.uni.data <- fread(paste0(db.dir, "/NCBI.db.uni.tsv"), header = T)
  db.uni.data <- unique(db.uni.data[,..ranks])
  db.uni.data[db.uni.data == ''] <- NA

  if (NormToGenomeSize) {
    genome_sizes <- as.data.frame(fread(paste0(db.dir, '/NCBI.db.genomesize.tsv')))
    genome_sizes[genome_sizes == ''] <- NA
  }
  
  
  #### Fix Akkermansia municiphila
  db.uni.data <- data.table(plyr::rbind.fill(db.uni.data,
    setNames(data.frame(
      rbind(c('Bacteria', 'Verrucomicrobia', 'Verrucomicrobiae', 'Verrucomicrobiales', 'Akkermansiaceae', 'Akkermansia', 'Akkermansia muciniphila'))),
             ranks)
  ))
} else if (db == 'EMUdb') {

   ranks <- c("superkingdom", "phylum", "class", "order", "family", "genus", "species")
  
  try({ db.uni.data <- fread(file.path(db.dir, 'db.uni.data.tsv')) })
  if(!exists('db.uni.data')) {
    emu.db    <- read.delim(paste0(db.dir, '/taxonomy.tsv'))
    colnames(emu.db)[1] <- 'taxid'
    emu.db <- emu.db[,c('taxid', ranks)]
    emu.db[emu.db == ''] <- NA
  
    emu.fasta <- seqinr::read.fasta(paste0(db.dir, '/species_taxid.fasta'), whole.header = T)
    emu.fasta <- data.frame(names=names(emu.fasta), taxid=gsub(':.*', '', names(emu.fasta)))
    emu.fasta$seq_id <- gsub(' \\[.*', '', emu.fasta$names)
  
    emu.idx    <- merge(emu.fasta, emu.db, by='taxid')
  
    db.data <- data.table(emu.db)
    db.name <- db #'EMUdb'
    db.uni.data <- db.data
  }


}


```


## Import Exp data
```{r, import Experimental Results}
#### Import Zymo D3600 results

ps.all <- readRDS("EMU/ps/unfiltered/primer.exp.all.PS.rds")

metadata <- data.frame(sample_data(ps.all))

## correct metadata
ps.all@sam_data$platform <- 'ONT'
ps.all@sam_data$workflow <- 'EMU'
ps.all@sam_data$db       <- 'EMUdb'
ps.all@sam_data$Vregion  <- 'V1_V9'

metadata <- data.frame(sample_data(ps.all))
otutab   <- data.frame(otu_table(ps.all))
taxtab   <- data.frame(tax_table(ps.all))

#taxtab[taxtab$species == 'unclassified',]
#colSums(otutab[taxtab$species == 'unclassified',]) / colSums(otutab[,]) *100


```

## Filter Exp data
```{r}
## Filter for samples
samples.to.keep <- c(
  sample_names(ps.all)[
    grepl('human',   metadata$source)]
)

ps.all <- prune_samples(samples.to.keep, ps.all)

##
ps.all@sam_data$sample <- sample_names(ps.all)  



## Filter out taxa that no longer have an abundance
abundance_threshold <- 1
filter_func <- filterfun(function(x) sum(x) > abundance_threshold)
ps.all <- filter_taxa(ps.all, filter_func, prune = TRUE)

metadata <- data.frame(sample_data(ps.all))
otutab   <- data.frame(otu_table(ps.all))
taxtab   <- data.frame(tax_table(ps.all))

```


## Filter samples based on total read count
```{r}
samples.to.keep <- colnames(
  as.data.frame(otutab[,colSums(otutab) >= abund.threshold])
  )

message("The following samples were filtered out, because they had a read count less than ", abund.threshold, ": \n", 
        paste0(setdiff(sample_names(ps.all),  samples.to.keep), collapse = '\n'))

ps.all <- prune_samples(samples.to.keep, ps.all)

abundance_threshold <- 1
filter_func <- filterfun(function(x) sum(x) > abundance_threshold)
ps.all <- filter_taxa(ps.all, filter_func, prune = TRUE)


metadata <- data.frame(sample_data(ps.all))
otutab   <- data.frame(otu_table(ps.all))
taxtab   <- data.frame(tax_table(ps.all))


```


## Fix sample names
```{r}
setDT(metadata)
metadata[,group        := .GRP, by=.(source, primer, Vregion, platform, workflow, db)] 
metadata[,sample_nr    := seq_along(sample), by=.(group)]
metadata[,sample_name  := paste0(primer, sample_nr)]

metadata <- data.frame(metadata, row.names = metadata$sample)

ps.all@sam_data <- sample_data(metadata)


metadata <- data.frame(sample_data(ps.all))
otutab   <- data.frame(otu_table(ps.all))
taxtab   <- data.frame(tax_table(ps.all))

```


## Add lineage and metadata to Gold Standard Data

*Limosilactobacillus fermentum* in the theoretical composition was changed to *Lactobacillus fermentum*
```{r, Import Theoretical Composition, eval=T}
###
#zymo.GS[,count_colname] <- as.numeric(zymo.GS[,count_colname])

colnames(zymo.GS)[1] <- 'species'
zymo.GS <- zymo.GS[,c('species', count_colname)]
colnames(zymo.GS)[2] <- 'count'

zymo.GS <- zymo.GS[zymo.GS$count > 0, ]

zymo.GS$count <- zymo.GS$count * count_multiplier

## summarise on species-level (mainly because of Eschericia strains)
if(glom_spec) {
  setDT(zymo.GS)
  zymo.GS$strain  <- zymo.GS$species
  zymo.GS$species <- gsub(' \\(.*', '', zymo.GS$species)
  cols_by         <- c('species') 
  zymo.GS[,count := sum(count), by=cols_by]
  zymo.GS[,strain := NULL]
  zymo.GS <- unique(zymo.GS)
}


if(!is.null(crop_taxa)) {
  zymo.GS <- zymo.GS[grep(crop_taxa, zymo.GS$species, invert = T),]
}

zymo.GS <- as.data.frame(merge(db.uni.data,  zymo.GS, by='species', all.y=T))
if(!consider.euk) { ## will be done later
  zymo.GS <- zymo.GS[
    #zymo.GS$superkingdom != 'Eukaryota'
    #grep('Saccharomyces|Candida|Cryptococcus', zymo.GS$species, invert = T)
    ,]
}

mtgs <- unique.data.frame(as.data.frame(metadata[,vars])); colnames(mtgs) <- vars
mtgs$sample_name <- GS_name
mtgs <- unite(mtgs, 'sample', c('sample_name', vars), remove = F)
rownames(mtgs) <- mtgs$sample

taxtab.gs <- zymo.GS[,ranks]
rownames(taxtab.gs) <- taxtab.gs[,'species']

otutab.gs <- setNames(data.frame(replicate(length(mtgs$sample), zymo.GS$count, simplify = FALSE), row.names = rownames(taxtab.gs)), mtgs$sample)

mtgs$GS_name   <- GS_name
mtgs$sample_nr <- GS_name

#mtgs <- merge(metadata.gs, mtgs, by=c(0,1,2,3))[,-1]
#rownames(mtgs) <- mtgs$sample


ps.gs <- phyloseq(otu_table(as.matrix(otutab.gs), taxa_are_rows = T), 
                   tax_table(as.matrix(taxtab.gs)), 
                   sample_data(mtgs))
```


```{r, Merge Theoretical Compositions}

ps.gsa <- ps.gs

```


```{r, eval=F}
### Add Gold Standard version to all sample names
sample_names(ps.all) <- paste0(sample_names(ps.all), '_', GS_version)
ps.all@sam_data$GS_version <- GS_version

### Fix sample names
ps.all@sam_data$sample <- sample_names(ps.all)

```



```{r, NA to unclassified}

#taxtab[is.na(taxtab)] <- 'unclassified'

#ps.all   <- phyloseq(otu_table(as.matrix(otutab), taxa_are_rows = T),
#                     tax_table(as.matrix(taxtab)),
#                     sample_data(metadata))

ps.all@tax_table[is.na(ps.all@tax_table)] <- 'unassigned'

otutab   <- data.frame(otu_table  (ps.all))
taxtab   <- data.frame(tax_table  (ps.all))
metadata <- data.frame(sample_data(ps.all))
```



```{r, Exclude unclassified reads}

if (exclude.unclass) {
  
  snames <- sample_names(ps.all)
  
  otutab <- otutab[rownames(otutab) != 'unassigned', ]
  taxtab <- taxtab[rownames(taxtab) != 'unassigned', ]
  
  otutab <- otutab[taxtab$species   != 'unassigned', ]
  taxtab <- taxtab[taxtab$species   != 'unassigned', ]
  
  otutab <- otutab[taxtab$phylum    != 'unassigned', ]
  taxtab <- taxtab[taxtab$phylum    != 'unassigned', ]
  
  colnames(otutab) <- snames
  
  ps.all   <- phyloseq(otu_table(as.matrix(otutab), taxa_are_rows = T),
                       tax_table(as.matrix(taxtab)),
                       sample_data(metadata))

}

metadata <- data.frame(sample_data(ps.all))
```


```{r}
## Filter for taxa below threshold, again ?
if (!is.na(abund.threshold)) {

  snames <- sample_names(ps.all)
  
  otutab   <- data.frame(otu_table(ps.all))
  taxtab   <- data.frame(tax_table(ps.all))
  metadata <- data.frame(sample_data(ps.all))

  otutab  <- otutab[rowSums(otutab) >= abund.threshold, ]
  taxtab  <- taxtab[rowSums(otutab) >= abund.threshold, ]

  colnames(otutab) <- snames
  
  ps.all   <- phyloseq(otu_table(as.matrix(otutab), taxa_are_rows = T),
                       tax_table(as.matrix(taxtab)),
                       sample_data(metadata))


}

```


```{r}
### Check if species names are unique
if( n_distinct(ps.all@tax_table[,'species']) == nrow(ps.all@tax_table[,]) ) {
  
  message('using species as taxon identifier')
  
  ### And get rid of tax.identity AND lineage from taxtab if so  
  
  ps <- ps.all
  
  metadata   <- data.frame(sample_data(ps))
  otutab     <- data.frame(otu_table(ps))
  taxtab     <- data.frame(tax_table(ps))
  samp_names <- sample_names(ps)

  rownames(taxtab) <- taxtab$species
  rownames(otutab) <- rownames(taxtab)
  
  taxtab           <- taxtab[,ranks]  
  
  colnames(otutab) <- samp_names
  
  ps <- phyloseq(tax_table(as.matrix(taxtab)),
                 otu_table(as.matrix(otutab), taxa_are_rows = T),
                 sample_data(metadata))
  
  ps.all <- ps
 
} else {
  
  ### Otherwise get rid of tax.identity OR lineage from taxtab if so
  
  
  rank_names(ps.all)
  
  ps <- ps.all
  
  metadata   <- data.frame(sample_data(ps))
  otutab     <- data.frame(otu_table(ps))
  taxtab     <- data.frame(tax_table(ps))
  samp_names <- sample_names(ps)
  colnames(otutab)   <- samp_names
  rownames(metadata) <- samp_names
  
  taxtabuni <- unique.data.frame(taxtab[,c(ranks, 'tax.identity')]) 
  stopifnot(nrow(taxtabuni) == nrow(taxtab))
  dups <- dup(taxtabuni$tax.identity)
  if(length(dups) == 0) {
    message('using tax.identity as taxon identifier')
    rownames(taxtabuni) <- taxtabuni$tax.identity  
  } else {
    message('using lineage as taxon identifier')
    #rownames(taxtabuni) <- taxtabuni$tax.identity  
  }
  
  taxtabuni <- unique.data.frame(taxtabuni[,c(ranks)])
  stopifnot(nrow(taxtabuni) == nrow(taxtab))
  
  rownames(otutab) <- rownames(taxtabuni)
  
  ps <- phyloseq(tax_table(as.matrix(taxtabuni)),
                 otu_table(as.matrix(otutab), taxa_are_rows = T),
                 sample_data(metadata))
  
  
  ## Glomerating on sepcies level for check
  ps.glom <- tax_glom_fast(ps, rank_level =  7, ignore_lineage = T)
  
  
  ##
  metadata   <- data.frame(sample_data(ps))
  otutab     <- data.frame(otu_table(ps))
  taxtab     <- data.frame(tax_table(ps))
  samp_names <- sample_names(ps)
  colnames(otutab)   <- samp_names
  rownames(metadata) <- samp_names

}




```

```{r, Fix genera}


if(!is.null(genera_to_glom)) {
    
  taxatofilt <- taxa_names(ps.all)[!grepl(paste0(genera_to_glom$glom, collapse='|'), taxtab$genus)]

  ps <- prune_taxa(taxatofilt, ps.all)

  for (i in 1:nrow(genera_to_glom)) {
   
    glom   <- genera_to_glom[i, 'glom']
    to     <- genera_to_glom[i, 'to']
    
    message('Glomerating all hits from: ', glom, ' to: ', to)
    
    taxatokeep <- taxa_names(ps.all)[grepl(glom, taxtab$genus)]
    
    try({
    ## Glomerate several taxa from the same genus
    ps.glom     <- prune_taxa(taxatokeep, ps.all)
    ps.glom     <- tax_glom_fast(ps.glom, rank_level = 6, ignore_lineage = F)
    taxtab.glom <- data.frame(tax_table(ps.glom))
    
    ## Change the species name of these to the desired
    taxtab.glom$species <- to
    ps.glom@tax_table <- tax_table(as.matrix(taxtab.glom))
    taxa_names(ps.glom) <- to
    taxtab.glom <- data.frame(tax_table(ps.glom))
    
    ##
    ps <- merge_phyloseq(ps.glom, ps)
    
    })
    
  }
  
  ps.all <- ps
  
}

otutab   <- data.frame(otu_table(ps))
taxtab   <- data.frame(tax_table(ps))
metadata <- data.frame(sample_data(ps))
##

```


```{r, Fix species}

if(!is.null(species_to_glom)) {
  taxtab <- data.frame(tax_table(ps.all))
  for (i in 1:nrow(species_to_glom)) {
   
    glom <- species_to_glom$glom[i]
    to   <- species_to_glom$to[i]
    message('Glomerating all hits from: ', paste(glom, collapse = ' ,'), ' to: ', to)
    
    taxtab[rownames(taxtab) == glom, 'species'] <- to
  }
  ps.all@tax_table <- tax_table(as.matrix(taxtab))
  taxtab <- data.frame(tax_table(ps.all))
}

##

```
## Fix sample names from '-'
```{r}

sample_names(ps.all) <- gsub('-', '.', sample_names(ps.all))

```



## Merge exp data with GS
```{r, eval=F}
metadata.gs <- data.frame(sample_data(ps.gsa))

## this caused non-sensical metadata changes
ps.all   <- merge_phyloseq(ps.gsa, ps.all)

## use own function to merge
#ps.all <- merge_phylo_fast(ps1 = ps.gsa, ps2 = ps.standard)

###
metadata <- data.frame(sample_data(ps.all))
otutab   <- data.frame(otu_table(ps.all))
taxtab   <- data.frame(tax_table(ps.all))
```



## Filter Taxa for Gold Standard

These are the taxa that were found in the experimental data
```{r, Gold Standard and Experimental Matches}
otutab.gsa     <- data.frame(otu_table(ps.gsa))
taxtab.gsa     <- data.frame(tax_table(ps.gsa))
metadata.gsa   <- data.frame(sample_data(ps.gsa))

otutab.sum     <- data.frame(EXP_sum=rowSums(otutab), row.names = rownames(otutab))

gs.exp.match   <- merge(zymo.GS, otutab.sum, all.x=T, by.x='species', by.y=0)

print(gs.exp.match)
```


```{r, keep taxa from Gold Standard}

taxa.gs        <- taxa_names(ps.gsa)
ps.standard    <- prune_taxa(taxa.gs, ps.all)

abundance_threshold <- abund.threshold
filter_func  <- filterfun(function(x) sum(x) > abundance_threshold)
   
ps.standard  <- filter_taxa(ps.standard, filter_func, prune = TRUE)


if (keep.GS.taxa.only) {

  ps.all <- ps.standard
  
} else {
  
  ps.all  <- filter_taxa(ps.all, filter_func, prune = TRUE) 
  
}


taxtab   <- data.frame(tax_table(ps.all))
otutab   <- data.frame(otu_table(ps.all))
metadata <- data.frame(sample_data(ps.all))

```


## Add Gram Stain Info
```{r, eval=T}

taxtab   <- data.frame(tax_table(ps.all))
otutab   <- data.frame(otu_table(ps.all))
metadata <- data.frame(sample_data(ps.all))

rownames <- rownames(taxtab)
taxtab$rownames <- rownames

stopifnot(all.equal(taxa_names(ps.all), rownames))

taxtab.gram <- merge(taxtab, phyla.gram, by='phylum', all.x=T)

taxtab      <- data.frame(taxtab.gram[,c(ranks, 'Gram.Stain')], row.names = taxtab.gram$rownames)

taxtab      <- taxtab[rownames,]
otutab      <- otutab[rownames,]

ps.all <- phyloseq(tax_table(as.matrix(taxtab)),
                   otu_table(as.matrix(otutab), taxa_are_rows = T),
                   sample_data(metadata))


taxtab   <- data.frame(tax_table(ps.all))
otutab   <- data.frame(otu_table(ps.all))
metadata <- data.frame(sample_data(ps.all))


```



## Omit Eukaryotes
```{r, Omit Eukaryotes?}

if(!consider.euk) {
  taxtab      <- data.frame(tax_table(ps.all))
  taxatoprune <- rownames(taxtab)[!is.element(taxtab$superkingdom, 'Eukaryota') & !is.na(taxtab$superkingdom)]
  ps.all      <- prune_taxa(taxatoprune, ps.all)
  
}

otutab   <- data.frame(otu_table(ps.all))
taxtab   <- data.frame(tax_table(ps.all))
metadata <- data.frame(sample_data(ps.all))

```



#  *Homo sapiens*: Relative abundance of taxa at each taxonomic level

```{r}

ps <- ps.all

```



```{r, Barplots}
scales <- 'fixed'
fig.height  <- 10
fig.width   <- 20
top <- 20
top.MCM <- 20

multiV <- c(F, T)

fig.height.vr <- fig.height
fig.width.vr  <- fig.width

## plotting function for barplots
plotfun <- function(ps, x='sample_name', t=t, fill=t) {
  plot_bar(ps, x=x, fill=fill
         ) + 
    scale_fill_manual(values=as.character(pal.man)) +
    guides(fill = guide_legend(direction = "vertical", ncol = 1)) +
    theme_ipsum() + 
    theme(legend.position="right", 
          legend.text = element_text(face = "italic"),
          panel.spacing.x = unit(1.5, 'mm'),
          axis.text.x = element_text(),
          axis.title.x = element_blank())
    # facet_grid(rows = vars(sequencing_date), cols = vars(DNA_isolation_method), scales = 'free')
    # facet_nested(cols   = vars(sample_name), rows   = vars(primer), scales = 'free_x')
}

```

## Gram Stain level

```{r, Glom taxa at Gram-level}
n <- 8

t <- rank_names(ps)[n]

## glomerate at n taxonomic level, regardless of lineage
ps.glom <- tax_glom_fast(ps, rank_level = n, ignore_lineage = T)
otutab <- data.frame(ps.glom@otu_table)

## normalize PS object per taxon sums
ps.glom.norm       <- norm.ps(ps.glom, 'sum')
#ps.glom.norm@otu_table

## top20 taxa from the glomerated at specific tax level
top.taxa           <- names(sort(taxa_sums(ps.glom.norm), decreasing=TRUE))[1:top]

## add all Gold standard taxa?
if (add.GS.taxa) {
  try({ top.taxa  <- unique(c(top.taxa, taxtab.gs[,t])) })
}

## prune normalized taxa
ps.glom.norm.top   <- prune_taxa(top.taxa, ps.glom.norm)
## prune taxa
ps.glom.top        <- prune_taxa(top.taxa, ps.glom)

## normalize PS object per taxon sums
ps.glom.norm       <- norm.ps(ps.glom, 'sum')

## top20 taxa from the glomerated at specific tax level
top.taxa           <- names(sort(taxa_sums(ps.glom.norm), decreasing=TRUE))[1:top]

## add all Gold standard taxa?
if (add.GS.taxa) {
  try({ top.taxa  <- unique(c(top.taxa, taxtab.gs[,t])) })
}

## prune normalized taxa
ps.glom.norm.top   <- prune_taxa(top.taxa, ps.glom.norm)
## prune taxa
ps.glom.top        <- prune_taxa(top.taxa, ps.glom)

## get data
ps.data.glom.top  <- df.from.ps(ps.glom.top,   comb=F)
ps.data.glom      <- df.from.ps(ps.glom,       comb=F)

## filenames for plots and tables
filenames <- c(
  paste0(res.dir, '_top' , top, '_', t, '_ratios'),
  paste0(res.dir, '_all_', t,           '_ratios') 
)
## write tables
if (writetables){
 write_tsv(ps.data.glom.top, paste0(res.dir, '/', filenames[1], '.tsv')) 
 write_tsv(ps.data.glom,     paste0(res.dir, '/', filenames[2], '.tsv')) 
}

```


```{r, Gram-level plot, fig.height=fig.height.vr, fig.width = fig.width.vr, eval=multiV[2]}

ggtop20    <- plotfun(
  ps.glom.norm.top,
                t=t, fill=t) + 
  theme(legend.position = 'right') +
  ggtitle(paste0('Top', top, ' ', t))

ggtop20

if (save.Figs){
  ggsave(paste0(res.dir, '/', filenames[1], '_Vregions.jpg'), width = fig.width.vr, height = fig.height.vr)
}



ps.glom.norm.top.gram <- ps.glom.norm.top
ps.glom.norm.gram     <- ps.glom.norm
ps.glom.top.gram      <- ps.glom.top
ps.glom.gram          <- ps.glom


```



## Phylum level

```{r, Glom taxa at Phylum-level}
n <- 2

t <- rank_names(ps)[n]

## glomerate at n taxonomic level, regardless of lineage
ps.glom <- tax_glom_fast(ps, rank_level = n, ignore_lineage = T)
otutab <- data.frame(ps.glom@otu_table)

## normalize PS object per taxon sums
ps.glom.norm       <- norm.ps(ps.glom, 'sum')
#ps.glom.norm@otu_table

## top20 taxa from the glomerated at specific tax level
top.taxa           <- names(sort(taxa_sums(ps.glom.norm), decreasing=TRUE))[1:top]

## add all Gold standard taxa?
if (add.GS.taxa) {
  top.taxa  <- unique(c(top.taxa, taxtab.gs[,t]))
}

## prune normalized taxa
ps.glom.norm.top   <- prune_taxa(top.taxa, ps.glom.norm)
## prune taxa
ps.glom.top        <- prune_taxa(top.taxa, ps.glom)

## normalize PS object per taxon sums
ps.glom.norm       <- norm.ps(ps.glom, 'sum')

## top20 taxa from the glomerated at specific tax level
top.taxa           <- names(sort(taxa_sums(ps.glom.norm), decreasing=TRUE))[1:top]

## add all Gold standard taxa?
if (add.GS.taxa) {
  top.taxa  <- unique(c(top.taxa, taxtab.gs[,t]))
}

## prune normalized taxa
ps.glom.norm.top   <- prune_taxa(top.taxa, ps.glom.norm)
## prune taxa
ps.glom.top        <- prune_taxa(top.taxa, ps.glom)

## get data
ps.data.glom.top  <- df.from.ps(ps.glom.top,   comb=F)
ps.data.glom      <- df.from.ps(ps.glom,       comb=F)

## filenames for plots and tables
filenames <- c(
  paste0(res.dir, '_top' , top, '_', t, '_ratios'),
  paste0(res.dir, '_all_', t,           '_ratios') 
)
## write tables
if (writetables){
 write_tsv(ps.data.glom.top, paste0(res.dir, '/', filenames[1], '.tsv')) 
 write_tsv(ps.data.glom,     paste0(res.dir, '/', filenames[2], '.tsv')) 
}

```

```{r}

## Mark GS Taxa with Asterisks
if (mark.GS.Taxa) {
  ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]] <- paste0(ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]], "*")
}

```

```{r}
if (merge.non.GS.Taxa) {
  gs.taxa          <- unique(taxtab.gs[,t])
  ps.glom.norm.top.merged <- merge.other.Taxa(ps.glom.norm.top, gs.taxa )
} else {
  ps.glom.norm.top.merged <- ps.glom.norm.top
}

```


```{r, Phylum-level plot, fig.height=fig.height.vr, fig.width = fig.width.vr, eval=multiV[2]}

ggtop20    <- plotfun(
  ps.glom.norm.top.merged,
                t=t, fill=t) + 
  theme(legend.position = 'right') +
  ggtitle(paste0('Top', top, ' ', t))

ggtop20

if (save.Figs){
  ggsave(paste0(res.dir, '/', filenames[1], '_Vregions.jpg'), width = fig.width.vr, height = fig.height.vr)
}



ps.glom.norm.top.phy <- ps.glom.norm.top
ps.glom.norm.phy     <- ps.glom.norm
ps.glom.top.phy      <- ps.glom.top
ps.glom.phy          <- ps.glom


```

## Class level

```{r, Glom taxa at Class-level}
n <- 3

t <- rank_names(ps)[n]

## glomerate at n taxonomic level, regardless of lineage
ps.glom <- tax_glom_fast(ps, rank_level = n, ignore_lineage = T)

## normalize PS object per taxon sums
ps.glom.norm       <- norm.ps(ps.glom, 'sum')

## top20 taxa from the glomerated at specific tax level
top.taxa           <- names(sort(taxa_sums(ps.glom.norm), decreasing=TRUE))[1:top]

## add all Gold standard taxa?
if (add.GS.taxa) {
 top.taxa  <- unique(c(top.taxa, taxtab.gs[,t]))
}

## prune normalized taxa
ps.glom.norm.top   <- prune_taxa(top.taxa, ps.glom.norm)
## prune taxa
ps.glom.top        <- prune_taxa(top.taxa, ps.glom)

## normalize PS object per taxon sums
ps.glom.norm       <- norm.ps(ps.glom, 'sum')

## top20 taxa from the glomerated at specific tax level
top.taxa           <- names(sort(taxa_sums(ps.glom.norm), decreasing=TRUE))[1:top]

## add all Gold standard taxa?
if (add.GS.taxa) {
  top.taxa  <- unique(c(top.taxa, taxtab.gs[,t]))
}

## prune normalized taxa
ps.glom.norm.top   <- prune_taxa(top.taxa, ps.glom.norm)
## prune taxa
ps.glom.top        <- prune_taxa(top.taxa, ps.glom)

## get data
ps.data.glom.top  <- df.from.ps(ps.glom.top,   comb=F)
ps.data.glom      <- df.from.ps(ps.glom,       comb=F)

## filenames for plots and tables
filenames <- c(
  paste0(res.dir, '_top' , top, '_', t, '_ratios'),
  paste0(res.dir, '_all_', t,           '_ratios') 
)
## write tables
if (writetables){
 write_tsv(ps.data.glom.top, paste0(res.dir, '/', filenames[1], '.tsv')) 
 write_tsv(ps.data.glom,     paste0(res.dir, '/', filenames[2], '.tsv')) 
}

```

```{r}

## Mark GS Taxa with Asterisks
if (mark.GS.Taxa) {
  ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]] <- paste0(ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]], "*")
}

```

```{r}
if (merge.non.GS.Taxa) {
  gs.taxa          <- unique(taxtab.gs[,t])
  ps.glom.norm.top.merged <- merge.other.Taxa(ps.glom.norm.top, gs.taxa )
} else {
  ps.glom.norm.top.merged <- ps.glom.norm.top
}

```


```{r, Class-level plot, fig.height = fig.height.vr, fig.width = fig.width.vr, eval=multiV[2]}


ggtop20    <- plotfun(
  ps.glom.norm.top.merged,
                t=t, fill=t) + 
  theme(legend.position = 'right') +
  ggtitle(paste0('Top', top, ' ', t))

ggtop20

if (save.Figs){
  ggsave(paste0(res.dir, '/', filenames[1], '_Vregions.jpg'), width = fig.width.vr, height = fig.height.vr)
}



ps.glom.norm.top.class <- ps.glom.norm.top
ps.glom.norm.class     <- ps.glom.norm
ps.glom.top.class      <- ps.glom.top
ps.glom.class          <- ps.glom

```


## Order level

```{r, Glom taxa at Order-level}
n <- 4

t <- rank_names(ps)[n]

## glomerate at n taxonomic level, regardless of lineage
ps.glom <- tax_glom_fast(ps, rank_level = n, ignore_lineage = T)

## normalize PS object per taxon sums
ps.glom.norm       <- norm.ps(ps.glom, 'sum')

## top20 taxa from the glomerated at specific tax level
top.taxa           <- names(sort(taxa_sums(ps.glom.norm), decreasing=TRUE))[1:top]

## add all Gold standard taxa?
if (add.GS.taxa) {
 top.taxa  <- unique(c(top.taxa, taxtab.gs[,t]))
}

## prune normalized taxa
ps.glom.norm.top   <- prune_taxa(top.taxa, ps.glom.norm)
## prune taxa
ps.glom.top        <- prune_taxa(top.taxa, ps.glom)

## normalize PS object per taxon sums
ps.glom.norm       <- norm.ps(ps.glom, 'sum')

## top20 taxa from the glomerated at specific tax level
top.taxa           <- names(sort(taxa_sums(ps.glom.norm), decreasing=TRUE))[1:top]

## add all Gold standard taxa?
if (add.GS.taxa) {
  top.taxa  <- unique(c(top.taxa, taxtab.gs[,t]))
}

## prune normalized taxa
ps.glom.norm.top   <- prune_taxa(top.taxa, ps.glom.norm)
## prune taxa
ps.glom.top        <- prune_taxa(top.taxa, ps.glom)

## get data
ps.data.glom.top  <- df.from.ps(ps.glom.top,   comb=F)
ps.data.glom      <- df.from.ps(ps.glom,       comb=F)

## filenames for plots and tables
filenames <- c(
  paste0(res.dir, '_top' , top, '_', t, '_ratios'),
  paste0(res.dir, '_all_', t,           '_ratios') 
)
## write tables
if (writetables){
 write_tsv(ps.data.glom.top, paste0(res.dir, '/', filenames[1], '.tsv')) 
 write_tsv(ps.data.glom,     paste0(res.dir, '/', filenames[2], '.tsv')) 
}

```

```{r}

## Mark GS Taxa with Asterisks
if (mark.GS.Taxa) {
  ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]] <- paste0(ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]], "*")
}

```

```{r}
if (merge.non.GS.Taxa) {
  gs.taxa          <- unique(taxtab.gs[,t])
  ps.glom.norm.top.merged <- merge.other.Taxa(ps.glom.norm.top, gs.taxa )
} else {
  ps.glom.norm.top.merged <- ps.glom.norm.top
}

```


```{r, Order-level plot, fig.height = fig.height.vr, fig.width = fig.width.vr, eval=multiV[2]}


ggtop20    <- plotfun(
  ps.glom.norm.top.merged,
                t=t, fill=t) + 
  theme(legend.position = 'right') +
  ggtitle(paste0('Top', top, ' ', t))

ggtop20

if (save.Figs){
  ggsave(paste0(res.dir, '/', filenames[1], '_Vregions.jpg'), width = fig.width.vr, height = fig.height.vr)
}



ps.glom.norm.top.ord <- ps.glom.norm.top
ps.glom.norm.ord     <- ps.glom.norm
ps.glom.top.ord      <- ps.glom.top
ps.glom.ord          <- ps.glom

```

## Family level

```{r, Glom taxa at Family-level}
n <- 5

t <- rank_names(ps)[n]

## glomerate at n taxonomic level, regardless of lineage
ps.glom <- tax_glom_fast(ps, rank_level = n, ignore_lineage = T)

## normalize PS object per taxon sums
ps.glom.norm       <- norm.ps(ps.glom, 'sum')

## top20 taxa from the glomerated at specific tax level
top.taxa           <- names(sort(taxa_sums(ps.glom.norm), decreasing=TRUE))[1:top]

## add all Gold standard taxa?
if (add.GS.taxa) {
  top.taxa  <- unique(c(top.taxa, taxtab.gs[,t]))
}

## prune normalized taxa
ps.glom.norm.top   <- prune_taxa(top.taxa, ps.glom.norm)
## prune taxa
ps.glom.top        <- prune_taxa(top.taxa, ps.glom)

## normalize PS object per taxon sums
ps.glom.norm       <- norm.ps(ps.glom, 'sum')

## top20 taxa from the glomerated at specific tax level
top.taxa           <- names(sort(taxa_sums(ps.glom.norm), decreasing=TRUE))[1:top]

## add all Gold standard taxa?
if (add.GS.taxa) {
  top.taxa  <- unique(c(top.taxa, taxtab.gs[,t]))
}

## prune normalized taxa
ps.glom.norm.top   <- prune_taxa(top.taxa, ps.glom.norm)
## prune taxa
ps.glom.top        <- prune_taxa(top.taxa, ps.glom)

## get data
ps.data.glom.top  <- df.from.ps(ps.glom.top,   comb=F)
ps.data.glom      <- df.from.ps(ps.glom,       comb=F)

## filenames for plots and tables
filenames <- c(
  paste0(res.dir, '_top' , top, '_', t, '_ratios'),
  paste0(res.dir, '_all_', t,           '_ratios') 
)
## write tables
if (writetables){
 write_tsv(ps.data.glom.top, paste0(res.dir, '/', filenames[1], '.tsv')) 
 write_tsv(ps.data.glom,     paste0(res.dir, '/', filenames[2], '.tsv')) 
}

```

```{r}

## Mark GS Taxa with Asterisks
if (mark.GS.Taxa) {
  ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]] <- paste0(ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]], "*")
}

```

```{r}
if (merge.non.GS.Taxa) {
  gs.taxa          <- unique(taxtab.gs[,t])
  ps.glom.norm.top.merged <- merge.other.Taxa(ps.glom.norm.top, gs.taxa )
} else {
  ps.glom.norm.top.merged <- ps.glom.norm.top
}

```


```{r, Family-level plot, fig.height = fig.height.vr, fig.width = fig.width.vr, eval=multiV[2]}

ggtop20    <- plotfun(
  ps.glom.norm.top.merged,
                t=t, fill=t) + 
  theme(legend.position = 'right') +
  ggtitle(paste0('Top', top, ' ', t))

ggtop20

if (save.Figs){
  ggsave(paste0(res.dir, '/', filenames[1], '_Vregions.jpg'), width = fig.width.vr, height = fig.height.vr)
}



ps.glom.norm.top.fam <- ps.glom.norm.top
ps.glom.norm.fam     <- ps.glom.norm
ps.glom.top.fam      <- ps.glom.top
ps.glom.fam          <- ps.glom

```


## Genus level

```{r, Glom taxa at Genus-level}
n <- 6

t <- rank_names(ps)[n]

## glomerate at n taxonomic level, regardless of lineage
ps.glom <- tax_glom_fast(ps, rank_level = n, ignore_lineage = T)

## normalize PS object per taxon sums
ps.glom.norm       <- norm.ps(ps.glom, 'sum')

## top20 taxa from the glomerated at specific tax level
top.taxa           <- names(sort(taxa_sums(ps.glom.norm), decreasing=TRUE))[1:top]

## add all Gold standard taxa?
if (add.GS.taxa) {
  top.taxa  <- unique(c(top.taxa, taxtab.gs[,t]))
}

## prune normalized taxa
ps.glom.norm.top   <- prune_taxa(top.taxa, ps.glom.norm)
## prune taxa
ps.glom.top        <- prune_taxa(top.taxa, ps.glom)

## normalize PS object per taxon sums
ps.glom.norm       <- norm.ps(ps.glom, 'sum')

## top20 taxa from the glomerated at specific tax level
top.taxa           <- names(sort(taxa_sums(ps.glom.norm), decreasing=TRUE))[1:top]

## add all Gold standard taxa?
if (add.GS.taxa) {
  top.taxa  <- unique(c(top.taxa, taxtab.gs[,t]))
}

## prune normalized taxa
ps.glom.norm.top   <- prune_taxa(top.taxa, ps.glom.norm)
## prune taxa
ps.glom.top        <- prune_taxa(top.taxa, ps.glom)

## get data
ps.data.glom.top  <- df.from.ps(ps.glom.top,   comb=F)
ps.data.glom      <- df.from.ps(ps.glom,       comb=F)

## filenames for plots and tables
filenames <- c(
  paste0(res.dir, '_top' , top, '_', t, '_ratios'),
  paste0(res.dir, '_all_', t,           '_ratios') 
)
## write tables
if (writetables){
 write_tsv(ps.data.glom.top, paste0(res.dir, '/', filenames[1], '.tsv')) 
 write_tsv(ps.data.glom,     paste0(res.dir, '/', filenames[2], '.tsv')) 
}

```

```{r}

## Mark GS Taxa with Asterisks
if (mark.GS.Taxa) {
  ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]] <- paste0(ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]], "*")
}

```

```{r}
if (merge.non.GS.Taxa) {
  gs.taxa          <- unique(taxtab.gs[,t])
  ps.glom.norm.top.merged <- merge.other.Taxa(ps.glom.norm.top, gs.taxa )
} else {
  ps.glom.norm.top.merged <- ps.glom.norm.top
}

```


```{r, Genus-level plot, fig.height = fig.height.vr, fig.width = fig.width.vr, eval=multiV[2]}

ggtop20    <- plotfun(
  ps.glom.norm.top.merged,
                t=t, fill=t) + 
  theme(legend.position = 'right') +
  ggtitle(paste0('Top', top, ' ', t))

ggtop20

if (save.Figs){
  ggsave(paste0(res.dir, '/', filenames[1], '_Vregions.jpg'), width = fig.width.vr, height = fig.height.vr)
}



ps.glom.norm.top.gen <- ps.glom.norm.top
ps.glom.norm.gen     <- ps.glom.norm
ps.glom.top.gen      <- ps.glom.top
ps.glom.gen          <- ps.glom

```


## Species level

```{r, Glom taxa at Species-level}
n <- 7

t <- rank_names(ps)[n]

## glomerate at n taxonomic level, regardless of lineage
#if (!all(taxa_names(ps) == data.frame(ps@tax_table$species))) {
  ps.glom <- tax_glom_fast(ps, rank_level = n, ignore_lineage = T)  
#} else {
#  ps.glom <- ps
#}


## normalize PS object per taxon sums
ps.glom.norm       <- norm.ps(ps.glom, 'sum')

## top20 taxa from the glomerated at specific tax level
top.taxa           <- names(sort(taxa_sums(ps.glom.norm), decreasing=TRUE))[1:top]

## add all Gold standard taxa?
if (add.GS.taxa) {
  top.taxa  <- unique(c(top.taxa, taxtab.gs[,t]))
}

## prune normalized taxa
ps.glom.norm.top   <- prune_taxa(top.taxa, ps.glom.norm)
## prune taxa
ps.glom.top        <- prune_taxa(top.taxa, ps.glom)

## normalize PS object per taxon sums
ps.glom.norm       <- norm.ps(ps.glom, 'sum')

## top20 taxa from the glomerated at specific tax level
top.taxa           <- names(sort(taxa_sums(ps.glom.norm), decreasing=TRUE))[1:top]

## add all Gold standard taxa?
if (add.GS.taxa) {
  top.taxa  <- unique(c(top.taxa, taxtab.gs[,t]))
}

## prune normalized taxa
ps.glom.norm.top   <- prune_taxa(top.taxa, ps.glom.norm)
## prune taxa
ps.glom.top        <- prune_taxa(top.taxa, ps.glom)

## get data
ps.data.glom.top  <- df.from.ps(ps.glom.top,   comb=F)
ps.data.glom      <- df.from.ps(ps.glom,       comb=F)

## filenames for plots and tables
filenames <- c(
  paste0(res.dir, '_top' , top, '_', t, '_ratios'),
  paste0(res.dir, '_all_', t,           '_ratios') 
)
## write tables
if (writetables){
 write_tsv(ps.data.glom.top, paste0(res.dir, '/', filenames[1], '.tsv')) 
 write_tsv(ps.data.glom,     paste0(res.dir, '/', filenames[2], '.tsv')) 
}

```

```{r}

## Mark GS Taxa with Asterisks
if (mark.GS.Taxa) {
  ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]] <- paste0(ps.glom.norm.top@tax_table[,t][ps.glom.norm.top@tax_table[,t] %in% taxtab.gs[,t]], "*")
}

```

```{r}
if (merge.non.GS.Taxa) {
  gs.taxa          <- unique(taxtab.gs[,t])
  ps.glom.norm.top.merged <- merge.other.Taxa(ps.glom.norm.top, gs.taxa )
} else {
  ps.glom.norm.top.merged <- ps.glom.norm.top
}

```


```{r, Species-level plot, fig.height = fig.height.vr, fig.width = fig.width.vr, eval=multiV[2]}

ggtop20    <- plotfun(
  ps.glom.norm.top.merged,
                t=t, fill=t) + 
  theme(legend.position = 'right') +
  ggtitle(paste0('Top', top, ' ', t))

ggtop20

if (save.Figs){
  ggsave(paste0(res.dir, '/', filenames[1], '_Vregions.jpg'), width = fig.width.vr, height = fig.height.vr)
}



ps.glom.norm.top.spec <- ps.glom.norm.top
ps.glom.norm.spec     <- ps.glom.norm
ps.glom.top.spec      <- ps.glom.top
ps.glom.spec          <- ps.glom

```
 
## Figure XB -- no eval
```{r, Figure XB, fig.height = fig.height.vr*1.5, fig.width = fig.width.vr, eval=F}

n <- 6
t <- rank_names(ps)[n]

## Mark GS Taxa with Asterisks
if (mark.GS.Taxa) {
  ps.glom.norm.top.gen@tax_table[,t][ps.glom.norm.top.gen@tax_table[,t] %in% taxtab.gs[,t]] <- paste0(ps.glom.norm.top.gen@tax_table[,t][ps.glom.norm.top.gen@tax_table[,t] %in% taxtab.gs[,t]], "*")
}

## Merge non-GS Taxa
if (merge.non.GS.Taxa) {
  gs.taxa          <- unique(taxtab.gs[,t])
  ps.glom.norm.top.merged <- merge.other.Taxa(ps.glom.norm.top.gen, gs.taxa )
} else {
  ps.glom.norm.top.merged <- ps.glom.norm.top.gen
}


fig4d    <- plotfun(
  ps.glom.norm.top.merged,
                t=t, fill=t) + 
  theme(legend.position = 'right') +
  ggtitle('b')

ggsave(paste0('FIGURES_REVISION', '/', 'Figure YB.jpg'), 
       width = fig.width.vr, height = fig.height.vr)

fig4d


saveRDS(ps.glom.norm.top.merged, paste0(res.dir, '/ps.glom.norm.top.gen.merged.rds'))
```



# Statistics
```{r}
group_vars <- c('platform', 'Vregion', 'workflow', 'db')
```


## Phylum level {.tabset}

```{r}
n <- 2
t <- rank_names(ps.all)[n]

if (correl == 'all') {
  ps <- ps.glom.norm.phy
} else if (correl == 'topN') {
  ps <- ps.glom.norm.top.phy
}
sampdat <- data.frame(sample_data(ps))
otutab  <- data.frame(otu_table  (ps))

group_sampdat <- unique.data.frame(sampdat[,group_vars])


```



```{r}
### Run DESeq

ps      <- ps.glom.phy
taxtab  <- data.frame(ps@tax_table)
otutab  <- data.frame(ps@otu_table)
sampdat <- data.frame(ps@sam_data )

taxtab.DT <- data.table(taxtab, taxon=rownames(taxtab))

# Run DESeq2 in Paired mode
de.seq <- phyloseq_to_deseq2(ps, ~primer)
de.seq <- DESeq(de.seq,  test = "Wald", fitType = "parametric", sfType = "poscounts" )

# Get the levels of the condition factor
conditions <- levels(colData(de.seq)$primer)

# Initialize an empty list to store results
results_list <- list()

# Perform pairwise comparisons between all levels
for (i in 1:(length(conditions) - 1)) {
  for (j in (i + 1):length(conditions)) {
    # Extract condition names
    condition1 <- conditions[i]
    condition2 <- conditions[j]
    
    # Get results for the comparison between condition1 and condition2
    res    <- results(de.seq, contrast = c("primer", condition1, condition2))
    res_dt <- as.data.table(as.data.frame(res), keep.rownames = "taxon")
    res_dt[, comparison := paste(condition1, "_vs_", condition2)]
    
    # Append to results list
    results_list[[paste(condition1, condition2, sep = "_vs_")]] <- res_dt
  }
}

# Combine all pairwise results into one data.table
res.deseq <- rbindlist(results_list, use.names = TRUE, fill = TRUE)


# merge with taxa
res.deseq <- merge(res.deseq, taxtab.DT, by='taxon', all.x=T)

res.deseq[,is.significant := fifelse(log2FoldChange >= 2 | log2FoldChange <= -2 & padj <= 0.05 , T, F)]

res.deseq$taxon_level <- t
res.deseq.spec <- res.deseq

significant_taxa <- res.deseq[is.significant == TRUE]
significant_taxa[,taxon := gsub('.*;', '', taxon)]
sigtaxa_N    <- length(unique(significant_taxa$taxon))


# figh   <- max(nrow(taxtab),  24)
# figw   <- 14

figh   <- min(max(sigtaxa_N, 8),  24)
figw   <- min(max(figh*2, 16), 32)

```


### Heatmap of VST-normalized abundance values

```{r, fig.height=figh, fig.width=figw}
# Get VST-normalized matrix from DESeq2 object
vst_normalized_data <- getVarianceStabilizedData(de.seq)

# Convert to data.table format for easier manipulation
vst_normalized_data <- data.table(as.data.frame(vst_normalized_data), keep.rownames = "taxon")

# keep just the actual taxon
vst_normalized_data[,taxon := gsub('.*;', '', taxon)]

# col names
colnames(vst_normalized_data)[-1] <- colData(de.seq)$name


# Convert to matrix format suitable for pheatmap
vst_matrix <- as.matrix(vst_normalized_data[, -1, with = FALSE])
rownames(vst_matrix) <- vst_normalized_data$taxon

# Extract sample information from colData
sample_info <- colData(de.seq)$name
names(sample_info) <- colnames(vst_matrix)

# Define color palette for the heatmap
heatmap_colors <- colorRampPalette(c("#00AFBB", "white", "#FC4E07"))(100)

# Define annotation colors
annotation_colors <- list(Sample = setNames(rainbow(40), unique(sample_info)))

# Define annotation colors using distinct divergent colors
distinct_colors <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b", "#e377c2", "#7f7f7f", "#bcbd22", "#17becf", "#393b79", "#637939", "#8c6d31", "#843c39", "#7b4173", "#5254a3", "#9c9ede", "#8ca252", "#bd9e39", "#ad494a", "#a55194", "#6b6ecf", "#b5cf6b", "#e7ba52", "#d6616b", "#ce6dbd", "#9c755f", "#d9d574", "#aec7e8", "#ffbb78", "#98df8a", "#ff9896", "#c5b0d5", "#c49c94", "#f7b6d2", "#c7c7c7", "#dbdb8d", "#9edae5", "#8c564b", "#ff7f0e")
annotation_colors <- list(Sample = setNames(distinct_colors[1:length(unique(sample_info))], unique(sample_info)))


# Create heatmap with annotation
pheatmap(vst_matrix, #[rownames(vst_matrix) %in% significant_taxa$taxon, ],
         color = heatmap_colors,
         cluster_rows = TRUE,       # Cluster rows (taxa)
         cluster_cols = TRUE,       # Cluster columns (samples)
         show_rownames = TRUE,      # Show row names (taxa)
         show_colnames = TRUE,      # Show column names (samples)
         fontsize_row = 8,          # Font size for row labels
         fontsize_col = 8,          # Font size for column labels
         main = "Heatmap of VST-normalized Taxon Abundances",
         annotation_col = data.frame(Sample = sample_info),
         annotation_colors = annotation_colors,
         fontface_row = "italic")


```



### Significant Differences - Statistics

```{r}
# 0. Number of significantly different taxa
sig.frq <- res.deseq[,.N,by=.(is.significant)]
sig.frq
ggsf <- ggplot(sig.frq, aes(is.significant, N)) +
  geom_col(aes(fill=is.significant)) +
  scale_fill_manual(values = c("FALSE" = "gray", "TRUE" = "red")) +
  labs(title = "Number of Significantly Different Taxa",
       #x = "Signifcantly Different",
       y = "Number of Taxa") +
  theme_minimal() +
  theme(legend.position = "none")

# 1. Volcano Plot: Visualizing log2FoldChange vs. -log10(p-value)
ggv <- ggplot(res.deseq, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point(aes(color = is.significant), alpha = 0.7) +
  scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "red")) +
  labs(title = "Volcano Plot of Differential Abundance",
       x = "Log2 Fold Change",
       y = "-Log10 P-value") +
  theme_minimal() +
  theme(legend.position = "bottom")

# 2. MA Plot: Visualizing average expression vs. log2FoldChange
ggma <- ggplot(res.deseq, aes(x = baseMean, y = log2FoldChange)) +
  geom_point(aes(color = is.significant), alpha = 0.7) +
  scale_x_log10() +
  scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "blue")) +
  labs(title = "MA Plot of Differential Abundance",
       x = "Average Expression (baseMean, log10 scale)",
       y = "Log2 Fold Change") +
  theme_minimal() +
  theme(legend.position = "bottom")

cowplot::plot_grid(ggsf, ggv, ggma, nrow=1, rel_widths = c(0.5,1,1))

```


### Significant Differences - Heatmap of log2FC values

```{r}
# 4. Heatmap of Significant Taxa

ggplot(significant_taxa, aes(x = comparison, y = taxon, fill = log2FoldChange)) +
  geom_tile() +
  scale_fill_gradient2(low = "#00AFBB", mid = "white", high = "#FC4E07", midpoint = 0) +
  scale_fill_viridis(option = "C", name = "log2 Fold Difference") +
  labs(title = "Heatmap of Significant Taxa",
       x = "Comparison",
       y = "Taxon",
       fill = "Log2 Fold Change") +
  theme_minimal() +
  theme(axis.text.y  = element_blank(), #element_text(face = "italic"),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y.left = element_text(angle = 0, face = "italic"),
        panel.spacing = unit(0.1, "lines"),
        strip.placement = "outside",
        ) + 
  facet_nested(rows = vars(phylum, taxon), scales = 'free_y', switch = "y")

```


### Significant Differences - Heatmap of VST-normalized Abundance Values
```{r}
# Get VST-normalized matrix from DESeq2 object
vst_normalized_data <- getVarianceStabilizedData(de.seq)

# Convert to data.table format for easier manipulation
vst_normalized_data <- data.table(as.data.frame(vst_normalized_data), keep.rownames = "taxon")


##
vst_normalized_melt <- melt(vst_normalized_data, variable.name = 'sample', value.name = 'abundance')
vst_normalized_melt <- merge(vst_normalized_melt, sampdat, by='sample')
vst_normalized_melt <- vst_normalized_melt[,.(name, primer, taxon, abundance)]
vst_normalized_melt <- merge(taxtab.DT, vst_normalized_melt, by='taxon')
#vst_normalized_melt <- merge(vst_normalized_melt, group_sampdat,  by='name')

# keep just the actual taxon
vst_normalized_melt[,taxon := gsub('.*;', '', taxon)]


plot_data           <- vst_normalized_melt[taxon %in% significant_taxa$taxon]

```

```{r, fig.width=figw, fig.height=figh}
##
p <- ggplot(plot_data, 
            aes(x = taxon, y = name, fill = abundance)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_gradient2(#low = "#d73027", mid = "white", high = "#4575b4",
                       low = "#FC4E07", mid = "white", high = "#00AFBB",
                       midpoint = 0, # median(melted_dists$distance, na.rm = TRUE), 
                       limit    =  c(-20,20), #range(res.deseq$log2FoldChange, na.rm = TRUE), # c(-3, 3),
                       name="VST-normalized Abundance") +
  scale_fill_viridis(option = "C", name = "VST-normalized Abundance") +
  theme_minimal() +
  coord_flip() +
  theme(axis.text.y  = element_blank(), #element_text(face = "italic"),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y.left = element_text(angle = 0, face = "italic"),
        panel.spacing = unit(0.1, "lines"),
        strip.placement = "outside",
      ) +
  facet_nested(rows=vars(phylum, taxon),
               cols=vars(primer), 
               scales = "free", space = "free", switch = 'y')


print(p)
```


### Significant Differences - Heatmap of VST-normalized Abundance Differences
```{r, fig.width=figw, fig.height=figh, eval=F}
##
vst_normalized_cast <- dcast(vst_normalized_melt, name + taxon ~ primer, value.var = 'abundance')
vst_normalized_cast <- merge(taxtab.DT, vst_normalized_cast, by='taxon')
vst_normalized_cast <- merge(vst_normalized_cast, group_sampdat,  by='name')
vst_normalized_cast <- unique(vst_normalized_cast)
vst_normalized_cast[,diff := ZHMW - ZMB]

plot_data           <- vst_normalized_cast[taxon %in% significant_taxa$taxon]

##
p <- ggplot(plot_data, 
            aes(x = taxon, y = name, fill = diff)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_gradient2(#low = "#d73027", mid = "white", high = "#4575b4",
                       low = "#FC4E07", mid = "white", high = "#00AFBB",
                       midpoint = 0, # median(melted_dists$distance, na.rm = TRUE), 
                       limit    =  c(-20,20), #range(res.deseq$log2FoldChange, na.rm = TRUE), # c(-3, 3),
                       name="VST-normalized Difference") +
  scale_fill_viridis(option = "C", name = "VST-normalized Difference") +
  theme_minimal() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(hjust = 1, face = 'italic'),
        #axis.title  = element_blank(),
        panel.background = element_rect(fill = "gray90", color = NA),
        #strip.text.y = element_blank(),
        #strip.text.y = element_text(face = 'italic'),
        panel.spacing = unit(0.2, 'mm')
        ) +
 facet_nested(cols=vars(sex, source),
              #rows=vars(grp2), 
              scales = "free", space = "free")

print(p)

```



### PERMANOVA

Full-model PERMANOVA: source + dna.isolation.method
```{r, fig.width=12, fig.height=6}
distance <- 'bray'

otutab_t <- data.frame(t(otu_table(ps)))
otutab   <- data.frame((otu_table(ps)))
sampdat  <- data.frame(sample_data(ps))


stopifnot(all(rownames(sampdat) == colnames(otutab)))
stopifnot(all(rownames(sampdat) == rownames(otutab_t)))

ps.dists <- phyloseq::distance(ps, distance, type='samples')

```

```{r}

permanova.full <- adonis2(otutab_t ~ primer, data=sampdat, method=distance, by = 'terms')
permanova.full
permanova.full <- data.frame(group=rownames(permanova.full), permanova.full)  

```

```{r, fig.width=12, fig.height=6}

ggplot(permanova.full, aes(x=group, y=R2)) + 
  geom_bar(stat="identity", fill="skyblue") +
  labs(title="R-squared values from PERMANOVA",
       y="R^2 value",
       x="library") +
  theme_ipsum() +
  theme(axis.text.x = element_text(angle=45, hjust=1)) # +facet_grid(cols=vars(group))
```



## Order level {.tabset}

```{r}
n <- 4
t <- rank_names(ps.all)[n]

if (correl == 'all') {
  ps <- ps.glom.norm.ord
} else if (correl == 'topN') {
  ps <- ps.glom.norm.top.ord
}
sampdat <- data.frame(sample_data(ps))
otutab  <- data.frame(otu_table  (ps))

group_sampdat <- unique.data.frame(sampdat[,group_vars])


```



```{r}
### Run DESeq

ps      <- ps.glom.ord
taxtab  <- data.frame(ps@tax_table)
otutab  <- data.frame(ps@otu_table)
sampdat <- data.frame(ps@sam_data)

taxtab.DT <- data.table(taxtab, taxon=rownames(taxtab))

# Run DESeq2 in Paired mode
de.seq <- phyloseq_to_deseq2(ps, ~primer)
de.seq <- DESeq(de.seq,  test = "Wald", fitType = "parametric", sfType = "poscounts" )

# Get the levels of the condition factor
conditions <- levels(colData(de.seq)$primer)

# Initialize an empty list to store results
results_list <- list()

# Perform pairwise comparisons between all levels
for (i in 1:(length(conditions) - 1)) {
  for (j in (i + 1):length(conditions)) {
    # Extract condition names
    condition1 <- conditions[i]
    condition2 <- conditions[j]
    
    # Get results for the comparison between condition1 and condition2
    res    <- results(de.seq, contrast = c("primer", condition1, condition2))
    res_dt <- as.data.table(as.data.frame(res), keep.rownames = "taxon")
    res_dt[, comparison := paste(condition1, "_vs_", condition2)]
    
    # Append to results list
    results_list[[paste(condition1, condition2, sep = "_vs_")]] <- res_dt
  }
}

# Combine all pairwise results into one data.table
res.deseq <- rbindlist(results_list, use.names = TRUE, fill = TRUE)


# merge with taxa
res.deseq <- merge(res.deseq, taxtab.DT, by='taxon', all.x=T)

res.deseq[,is.significant := fifelse(log2FoldChange >= 2 | log2FoldChange <= -2 & padj <= 0.05 , T, F)]

res.deseq$taxon_level <- t
res.deseq.spec <- res.deseq

significant_taxa <- res.deseq[is.significant == TRUE]
significant_taxa[,taxon := gsub('.*;', '', taxon)]
sigtaxa_N    <- length(unique(significant_taxa$taxon))



# figh   <- max(nrow(taxtab),  24)
# figw   <- 14

figh   <- min(max(sigtaxa_N, 8),  24)
figw   <- min(max(figh*2, 16), 32)

```


### Heatmap of VST-normalized abundance values

```{r, fig.height=figh, fig.width=figw}
# Get VST-normalized matrix from DESeq2 object
vst_normalized_data <- getVarianceStabilizedData(de.seq)

# Convert to data.table format for easier manipulation
vst_normalized_data <- data.table(as.data.frame(vst_normalized_data), keep.rownames = "taxon")

# keep just the actual taxon
vst_normalized_data[,taxon := gsub('.*;', '', taxon)]

# col names
colnames(vst_normalized_data)[-1] <- colData(de.seq)$name


# Convert to matrix format suitable for pheatmap
vst_matrix <- as.matrix(vst_normalized_data[, -1, with = FALSE])
rownames(vst_matrix) <- vst_normalized_data$taxon

# Extract sample information from colData
sample_info <- colData(de.seq)$name
names(sample_info) <- colnames(vst_matrix)

# Define color palette for the heatmap
heatmap_colors <- colorRampPalette(c("#00AFBB", "white", "#FC4E07"))(100)

# Define annotation colors
annotation_colors <- list(Sample = setNames(rainbow(40), unique(sample_info)))

# Define annotation colors using distinct divergent colors
distinct_colors <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b", "#e377c2", "#7f7f7f", "#bcbd22", "#17becf", "#393b79", "#637939", "#8c6d31", "#843c39", "#7b4173", "#5254a3", "#9c9ede", "#8ca252", "#bd9e39", "#ad494a", "#a55194", "#6b6ecf", "#b5cf6b", "#e7ba52", "#d6616b", "#ce6dbd", "#9c755f", "#d9d574", "#aec7e8", "#ffbb78", "#98df8a", "#ff9896", "#c5b0d5", "#c49c94", "#f7b6d2", "#c7c7c7", "#dbdb8d", "#9edae5", "#8c564b", "#ff7f0e")
annotation_colors <- list(Sample = setNames(distinct_colors[1:length(unique(sample_info))], unique(sample_info)))


# Create heatmap with annotation
pheatmap(vst_matrix, #[rownames(vst_matrix) %in% significant_taxa$taxon, ],
         color = heatmap_colors,
         cluster_rows = TRUE,       # Cluster rows (taxa)
         cluster_cols = TRUE,       # Cluster columns (samples)
         show_rownames = TRUE,      # Show row names (taxa)
         show_colnames = TRUE,      # Show column names (samples)
         fontsize_row = 8,          # Font size for row labels
         fontsize_col = 8,          # Font size for column labels
         main = "Heatmap of VST-normalized Taxon Abundances",
         annotation_col = data.frame(Sample = sample_info),
         annotation_colors = annotation_colors,
         fontface_row = "italic")


```



### Significant Differences - Statistics

```{r}
# 0. Number of significantly different taxa
sig.frq <- res.deseq[,.N,by=.(is.significant)]
sig.frq
ggsf <- ggplot(sig.frq, aes(is.significant, N)) +
  geom_col(aes(fill=is.significant)) +
  scale_fill_manual(values = c("FALSE" = "gray", "TRUE" = "red")) +
  labs(title = "Number of Significantly Different Taxa",
       #x = "Signifcantly Different",
       y = "Number of Taxa") +
  theme_minimal() +
  theme(legend.position = "none")

# 1. Volcano Plot: Visualizing log2FoldChange vs. -log10(p-value)
ggv <- ggplot(res.deseq, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point(aes(color = is.significant), alpha = 0.7) +
  scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "red")) +
  labs(title = "Volcano Plot of Differential Abundance",
       x = "Log2 Fold Change",
       y = "-Log10 P-value") +
  theme_minimal() +
  theme(legend.position = "bottom")

# 2. MA Plot: Visualizing average expression vs. log2FoldChange
ggma <- ggplot(res.deseq, aes(x = baseMean, y = log2FoldChange)) +
  geom_point(aes(color = is.significant), alpha = 0.7) +
  scale_x_log10() +
  scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "blue")) +
  labs(title = "MA Plot of Differential Abundance",
       x = "Average Expression (baseMean, log10 scale)",
       y = "Log2 Fold Change") +
  theme_minimal() +
  theme(legend.position = "bottom")

cowplot::plot_grid(ggsf, ggv, ggma, nrow=1, rel_widths = c(0.5,1,1))
```


### Significant Differences - Heatmap of log2FC values

```{r}
# 4. Heatmap of Significant Taxa

ggplot(significant_taxa, aes(x = comparison, y = taxon, fill = log2FoldChange)) +
  geom_tile() +
  scale_fill_gradient2(low = "#00AFBB", mid = "white", high = "#FC4E07", midpoint = 0) +
  scale_fill_viridis(option = "C", name = "log2 Fold Difference") +
  labs(title = "Heatmap of Significant Taxa",
       x = "Comparison",
       y = "Taxon",
       fill = "Log2 Fold Change") +
  theme_minimal() +
  theme(axis.text.y  = element_blank(), #element_text(face = "italic"),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y.left = element_text(angle = 0, face = "italic"),
        panel.spacing = unit(0.1, "lines"),
        strip.placement = "outside",
        ) + 
  facet_nested(rows = vars(phylum, taxon), scales = 'free_y', switch = "y")


```


### Significant Differences - Heatmap of VST-normalized Abundance Values
```{r}
# Get VST-normalized matrix from DESeq2 object
vst_normalized_data <- getVarianceStabilizedData(de.seq)

# Convert to data.table format for easier manipulation
vst_normalized_data <- data.table(as.data.frame(vst_normalized_data), keep.rownames = "taxon")


##
vst_normalized_melt <- melt(vst_normalized_data, variable.name = 'sample', value.name = 'abundance')
vst_normalized_melt <- merge(vst_normalized_melt, sampdat, by='sample')
vst_normalized_melt <- vst_normalized_melt[,.(name, primer, taxon, abundance)]
vst_normalized_melt <- merge(taxtab.DT, vst_normalized_melt, by='taxon')
#vst_normalized_melt <- merge(vst_normalized_melt, group_sampdat,  by='name')

# keep just the actual taxon
vst_normalized_melt[,taxon := gsub('.*;', '', taxon)]


plot_data           <- vst_normalized_melt[taxon %in% significant_taxa$taxon]

```

```{r, fig.width=figw, fig.height=figh}
##
p <- ggplot(plot_data, 
            aes(x = taxon, y = name, fill = abundance)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_gradient2(#low = "#d73027", mid = "white", high = "#4575b4",
                       low = "#FC4E07", mid = "white", high = "#00AFBB",
                       midpoint = 0, # median(melted_dists$distance, na.rm = TRUE), 
                       limit    =  c(-20,20), #range(res.deseq$log2FoldChange, na.rm = TRUE), # c(-3, 3),
                       name="VST-normalized Abundance") +
  scale_fill_viridis(option = "C", name = "VST-normalized Abundance") +
  theme_minimal() +
  coord_flip() +
  theme(axis.text.y  = element_blank(), #element_text(face = "italic"),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y.left = element_text(angle = 0, face = "italic"),
        panel.spacing = unit(0.1, "lines"),
        strip.placement = "outside",
      ) +
  facet_nested(rows=vars(phylum, taxon),
               cols=vars(primer), 
               scales = "free", space = "free", switch = 'y')


print(p)
```


### Significant Differences - Heatmap of VST-normalized Abundance Differences
```{r, fig.width=figw, fig.height=figh, eval=F}
##
vst_normalized_cast <- dcast(vst_normalized_melt, name + taxon ~ primer, value.var = 'abundance')
vst_normalized_cast <- merge(taxtab.DT, vst_normalized_cast, by='taxon')
vst_normalized_cast <- merge(vst_normalized_cast, group_sampdat,  by='name')
vst_normalized_cast <- unique(vst_normalized_cast)
vst_normalized_cast[,diff := ZHMW - ZMB]

plot_data           <- vst_normalized_cast[taxon %in% significant_taxa$taxon]

##
p <- ggplot(plot_data, 
            aes(x = taxon, y = name, fill = diff)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_gradient2(#low = "#d73027", mid = "white", high = "#4575b4",
                       low = "#FC4E07", mid = "white", high = "#00AFBB",
                       midpoint = 0, # median(melted_dists$distance, na.rm = TRUE), 
                       limit    =  c(-20,20), #range(res.deseq$log2FoldChange, na.rm = TRUE), # c(-3, 3),
                       name="VST-normalized Difference") +
  scale_fill_viridis(option = "C", name = "VST-normalized Difference") +
  theme_minimal() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(hjust = 1, face = 'italic'),
        #axis.title  = element_blank(),
        panel.background = element_rect(fill = "gray90", color = NA),
        #strip.text.y = element_blank(),
        #strip.text.y = element_text(face = 'italic'),
        panel.spacing = unit(0.2, 'mm')
        ) +
 facet_nested(cols=vars(sex, source),
              #rows=vars(grp2), 
              scales = "free", space = "free")

print(p)

```



### PERMANOVA

Full-model PERMANOVA: source + dna.isolation.method
```{r, fig.width=12, fig.height=6}
distance <- 'bray'

otutab_t <- data.frame(t(otu_table(ps)))
otutab   <- data.frame((otu_table(ps)))
sampdat  <- data.frame(sample_data(ps))


stopifnot(all(rownames(sampdat) == colnames(otutab)))
stopifnot(all(rownames(sampdat) == rownames(otutab_t)))

ps.dists <- phyloseq::distance(ps, distance, type='samples')

```

```{r}

permanova.full <- adonis2(otutab_t ~ primer, data=sampdat, method=distance, by = 'terms')
permanova.full
permanova.full <- data.frame(group=rownames(permanova.full), permanova.full)  

```

```{r, fig.width=12, fig.height=6}

ggplot(permanova.full, aes(x=group, y=R2)) + 
  geom_bar(stat="identity", fill="skyblue") +
  labs(title="R-squared values from PERMANOVA",
       y="R^2 value",
       x="library") +
  theme_ipsum() +
  theme(axis.text.x = element_text(angle=45, hjust=1)) # +facet_grid(cols=vars(group))
```



## Genus level {.tabset}

```{r}
n <- 6
t <- rank_names(ps.all)[n]

if (correl == 'all') {
  ps <- ps.glom.norm.gen
} else if (correl == 'topN') {
  ps <- ps.glom.norm.top.gen
}
sampdat <- data.frame(sample_data(ps))
otutab  <- data.frame(otu_table  (ps))

group_sampdat <- unique.data.frame(sampdat[,group_vars])


```



```{r}
### Run DESeq

ps      <- ps.glom.gen
taxtab  <- data.frame(ps@tax_table)
otutab  <- data.frame(ps@otu_table)
sampdat <- data.frame(ps@sam_data)

taxtab.DT <- data.table(taxtab, taxon=rownames(taxtab))

# Run DESeq2 in Paired mode
de.seq <- phyloseq_to_deseq2(ps, ~primer)
de.seq <- DESeq(de.seq,  test = "Wald", fitType = "parametric", sfType = "poscounts" )

# Get the levels of the condition factor
conditions <- levels(colData(de.seq)$primer)

# Initialize an empty list to store results
results_list <- list()

# Perform pairwise comparisons between all levels
for (i in 1:(length(conditions) - 1)) {
  for (j in (i + 1):length(conditions)) {
    # Extract condition names
    condition1 <- conditions[i]
    condition2 <- conditions[j]
    
    # Get results for the comparison between condition1 and condition2
    res    <- results(de.seq, contrast = c("primer", condition1, condition2))
    res_dt <- as.data.table(as.data.frame(res), keep.rownames = "taxon")
    res_dt[, comparison := paste(condition1, "_vs_", condition2)]
    
    # Append to results list
    results_list[[paste(condition1, condition2, sep = "_vs_")]] <- res_dt
  }
}

# Combine all pairwise results into one data.table
res.deseq <- rbindlist(results_list, use.names = TRUE, fill = TRUE)


# merge with taxa
res.deseq <- merge(res.deseq, taxtab.DT, by='taxon', all.x=T)

res.deseq[,is.significant := fifelse(log2FoldChange >= 2 | log2FoldChange <= -2 & padj <= 0.05 , T, F)]

res.deseq$taxon_level <- t
res.deseq.spec <- res.deseq

significant_taxa <- res.deseq[is.significant == TRUE]
significant_taxa[,taxon := gsub('.*;', '', taxon)]
sigtaxa_N    <- length(unique(significant_taxa$taxon))



# figh   <- max(nrow(taxtab),  24)
# figw   <- 14

figh   <- min(max(sigtaxa_N, 8),  24)
figw   <- min(max(figh*2, 16), 32)

```


### Heatmap of VST-normalized abundance values

```{r, fig.height=figh, fig.width=figw}
# Get VST-normalized matrix from DESeq2 object
vst_normalized_data <- getVarianceStabilizedData(de.seq)

# Convert to data.table format for easier manipulation
vst_normalized_data <- data.table(as.data.frame(vst_normalized_data), keep.rownames = "taxon")

# keep just the actual taxon
vst_normalized_data[,taxon := gsub('.*;', '', taxon)]

# col names
colnames(vst_normalized_data)[-1] <- colData(de.seq)$name


# Convert to matrix format suitable for pheatmap
vst_matrix <- as.matrix(vst_normalized_data[, -1, with = FALSE])
rownames(vst_matrix) <- vst_normalized_data$taxon

# Extract sample information from colData
sample_info <- colData(de.seq)$name
names(sample_info) <- colnames(vst_matrix)

# Define color palette for the heatmap
heatmap_colors <- colorRampPalette(c("#00AFBB", "white", "#FC4E07"))(100)

# Define annotation colors
annotation_colors <- list(Sample = setNames(rainbow(40), unique(sample_info)))

# Define annotation colors using distinct divergent colors
distinct_colors <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b", "#e377c2", "#7f7f7f", "#bcbd22", "#17becf", "#393b79", "#637939", "#8c6d31", "#843c39", "#7b4173", "#5254a3", "#9c9ede", "#8ca252", "#bd9e39", "#ad494a", "#a55194", "#6b6ecf", "#b5cf6b", "#e7ba52", "#d6616b", "#ce6dbd", "#9c755f", "#d9d574", "#aec7e8", "#ffbb78", "#98df8a", "#ff9896", "#c5b0d5", "#c49c94", "#f7b6d2", "#c7c7c7", "#dbdb8d", "#9edae5", "#8c564b", "#ff7f0e")
annotation_colors <- list(Sample = setNames(distinct_colors[1:length(unique(sample_info))], unique(sample_info)))


# Create heatmap with annotation
pheatmap(vst_matrix, #[rownames(vst_matrix) %in% significant_taxa$taxon, ],
         color = heatmap_colors,
         cluster_rows = TRUE,       # Cluster rows (taxa)
         cluster_cols = TRUE,       # Cluster columns (samples)
         show_rownames = TRUE,      # Show row names (taxa)
         show_colnames = TRUE,      # Show column names (samples)
         fontsize_row = 8,          # Font size for row labels
         fontsize_col = 8,          # Font size for column labels
         main = "Heatmap of VST-normalized Taxon Abundances",
         annotation_col = data.frame(Sample = sample_info),
         annotation_colors = annotation_colors,
         fontface_row = "italic")


```



### Significant Differences - Statistics

```{r}
# 0. Number of significantly different taxa
sig.frq <- res.deseq[,.N,by=.(is.significant)]
sig.frq
ggsf <- ggplot(sig.frq, aes(is.significant, N)) +
  geom_col(aes(fill=is.significant)) +
  scale_fill_manual(values = c("FALSE" = "gray", "TRUE" = "red")) +
  labs(title = "Number of Significantly Different Taxa",
       #x = "Signifcantly Different",
       y = "Number of Taxa") +
  theme_minimal() +
  theme(legend.position = "none")

# 1. Volcano Plot: Visualizing log2FoldChange vs. -log10(p-value)
ggv <- ggplot(res.deseq, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point(aes(color = is.significant), alpha = 0.7) +
  scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "red")) +
  labs(title = "Volcano Plot of Differential Abundance",
       x = "Log2 Fold Change",
       y = "-Log10 P-value") +
  theme_minimal() +
  theme(legend.position = "bottom")

# 2. MA Plot: Visualizing average expression vs. log2FoldChange
ggma <- ggplot(res.deseq, aes(x = baseMean, y = log2FoldChange)) +
  geom_point(aes(color = is.significant), alpha = 0.7) +
  scale_x_log10() +
  scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "blue")) +
  labs(title = "MA Plot of Differential Abundance",
       x = "Average Expression (baseMean, log10 scale)",
       y = "Log2 Fold Change") +
  theme_minimal() +
  theme(legend.position = "bottom")

cowplot::plot_grid(ggsf, ggv, ggma, nrow=1, rel_widths = c(0.5,1,1))
```


### Significant Differences - Heatmap of log2FC values

```{r}
# 4. Heatmap of Significant Taxa

ggplot(significant_taxa, aes(x = comparison, y = taxon, fill = log2FoldChange)) +
  geom_tile() +
  scale_fill_gradient2(low = "#00AFBB", mid = "white", high = "#FC4E07", midpoint = 0) +
  scale_fill_viridis(option = "C", name = "log2 Fold Difference") +
  labs(title = "Heatmap of Significant Taxa",
       x = "Comparison",
       y = "Taxon",
       fill = "Log2 Fold Change") +
  theme_minimal() +
  theme(axis.text.y  = element_blank(), #element_text(face = "italic"),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y.left = element_text(angle = 0, face = "italic"),
        panel.spacing = unit(0.1, "lines"),
        strip.placement = "outside",
        ) + 
  facet_nested(rows = vars(phylum, taxon), scales = 'free_y', switch = "y")


```


### Significant Differences - Heatmap of VST-normalized Abundance Values
```{r}
# Get VST-normalized matrix from DESeq2 object
vst_normalized_data <- getVarianceStabilizedData(de.seq)

# Convert to data.table format for easier manipulation
vst_normalized_data <- data.table(as.data.frame(vst_normalized_data), keep.rownames = "taxon")


##
vst_normalized_melt <- melt(vst_normalized_data, variable.name = 'sample', value.name = 'abundance')
vst_normalized_melt <- merge(vst_normalized_melt, sampdat, by='sample')
vst_normalized_melt <- vst_normalized_melt[,.(name, primer, taxon, abundance)]
vst_normalized_melt <- merge(taxtab.DT, vst_normalized_melt, by='taxon')
#vst_normalized_melt <- merge(vst_normalized_melt, group_sampdat,  by='name')

# keep just the actual taxon
vst_normalized_melt[,taxon := gsub('.*;', '', taxon)]


plot_data           <- vst_normalized_melt[taxon %in% significant_taxa$taxon]

```

```{r, fig.width=figw, fig.height=figh}
##
p <- ggplot(plot_data, 
            aes(x = taxon, y = name, fill = abundance)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_gradient2(#low = "#d73027", mid = "white", high = "#4575b4",
                       low = "#FC4E07", mid = "white", high = "#00AFBB",
                       midpoint = 0, # median(melted_dists$distance, na.rm = TRUE), 
                       limit    =  c(-20,20), #range(res.deseq$log2FoldChange, na.rm = TRUE), # c(-3, 3),
                       name="VST-normalized Abundance") +
  scale_fill_viridis(option = "C", name = "VST-normalized Abundance") +
  theme_minimal() +
  coord_flip() +
  theme(axis.text.y  = element_blank(), #element_text(face = "italic"),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y.left = element_text(angle = 0, face = "italic"),
        panel.spacing = unit(0.1, "lines"),
        strip.placement = "outside",
      ) +
  facet_nested(rows=vars(phylum, taxon),
               cols=vars(primer), 
               scales = "free", space = "free", switch = 'y')


print(p)
```


### Significant Differences - Heatmap of VST-normalized Abundance Differences
```{r, fig.width=figw, fig.height=figh, eval=F}
##
vst_normalized_cast <- dcast(vst_normalized_melt, name + taxon ~ primer, value.var = 'abundance')
vst_normalized_cast <- merge(taxtab.DT, vst_normalized_cast, by='taxon')
vst_normalized_cast <- merge(vst_normalized_cast, group_sampdat,  by='name')
vst_normalized_cast <- unique(vst_normalized_cast)
vst_normalized_cast[,diff := ZHMW - ZMB]

plot_data           <- vst_normalized_cast[taxon %in% significant_taxa$taxon]

##
p <- ggplot(plot_data, 
            aes(x = taxon, y = name, fill = diff)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_gradient2(#low = "#d73027", mid = "white", high = "#4575b4",
                       low = "#FC4E07", mid = "white", high = "#00AFBB",
                       midpoint = 0, # median(melted_dists$distance, na.rm = TRUE), 
                       limit    =  c(-20,20), #range(res.deseq$log2FoldChange, na.rm = TRUE), # c(-3, 3),
                       name="VST-normalized Difference") +
  scale_fill_viridis(option = "C", name = "VST-normalized Difference") +
  theme_minimal() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(hjust = 1, face = 'italic'),
        #axis.title  = element_blank(),
        panel.background = element_rect(fill = "gray90", color = NA),
        #strip.text.y = element_blank(),
        #strip.text.y = element_text(face = 'italic'),
        panel.spacing = unit(0.2, 'mm')
        ) +
 facet_nested(cols=vars(sex, source),
              #rows=vars(grp2), 
              scales = "free", space = "free")

print(p)

```



### PERMANOVA

Full-model PERMANOVA: source + dna.isolation.method
```{r, fig.width=12, fig.height=6}
distance <- 'bray'

otutab_t <- data.frame(t(otu_table(ps)))
otutab   <- data.frame((otu_table(ps)))
sampdat  <- data.frame(sample_data(ps))


stopifnot(all(rownames(sampdat) == colnames(otutab)))
stopifnot(all(rownames(sampdat) == rownames(otutab_t)))

ps.dists <- phyloseq::distance(ps, distance, type='samples')

```

```{r}

permanova.full <- adonis2(otutab_t ~ primer, data=sampdat, method=distance, by = 'terms')
permanova.full
permanova.full <- data.frame(group=rownames(permanova.full), permanova.full)  

```

```{r, fig.width=12, fig.height=6}

ggplot(permanova.full, aes(x=group, y=R2)) + 
  geom_bar(stat="identity", fill="skyblue") +
  labs(title="R-squared values from PERMANOVA",
       y="R^2 value",
       x="library") +
  theme_ipsum() +
  theme(axis.text.x = element_text(angle=45, hjust=1)) # +facet_grid(cols=vars(group))
```





## Species level {.tabset}

```{r}
n <- 7
t <- rank_names(ps.all)[n]

if (correl == 'all') {
  ps <- ps.glom.norm.spec
} else if (correl == 'topN') {
  ps <- ps.glom.norm.top.spec
}
sampdat <- data.frame(sample_data(ps))
otutab  <- data.frame(otu_table  (ps))

group_sampdat <- unique.data.frame(sampdat[,group_vars])


```



```{r}
### Run DESeq

ps      <- ps.glom.spec
taxtab  <- data.frame(ps@tax_table)
otutab  <- data.frame(ps@otu_table)
sampdat <- data.frame(ps@sam_data)

taxtab.DT <- data.table(taxtab, taxon=rownames(taxtab))

# Run DESeq2 in Paired mode
de.seq <- phyloseq_to_deseq2(ps, ~primer)
de.seq <- DESeq(de.seq,  test = "Wald", fitType = "parametric", sfType = "poscounts" )

# Get the levels of the condition factor
conditions <- levels(colData(de.seq)$primer)

# Initialize an empty list to store results
results_list <- list()

# Perform pairwise comparisons between all levels
for (i in 1:(length(conditions) - 1)) {
  for (j in (i + 1):length(conditions)) {
    # Extract condition names
    condition1 <- conditions[i]
    condition2 <- conditions[j]
    
    # Get results for the comparison between condition1 and condition2
    res    <- results(de.seq, contrast = c("primer", condition1, condition2))
    res_dt <- as.data.table(as.data.frame(res), keep.rownames = "taxon")
    res_dt[, comparison := paste(condition1, "_vs_", condition2)]
    
    # Append to results list
    results_list[[paste(condition1, condition2, sep = "_vs_")]] <- res_dt
  }
}

# Combine all pairwise results into one data.table
res.deseq <- rbindlist(results_list, use.names = TRUE, fill = TRUE)


# merge with taxa
res.deseq <- merge(res.deseq, taxtab.DT, by='taxon', all.x=T)

res.deseq[,is.significant := fifelse(log2FoldChange >= 2 | log2FoldChange <= -2 & padj <= 0.05 , T, F)]

res.deseq$taxon_level <- t
res.deseq.spec <- res.deseq

significant_taxa <- res.deseq[is.significant == TRUE]
significant_taxa[,taxon := gsub('.*;', '', taxon)]
sigtaxa_N    <- length(unique(significant_taxa$taxon))



# figh   <- max(nrow(taxtab),  24)
# figw   <- 14

figh   <- min(max(sigtaxa_N, 8),  24)
figw   <- min(max(figh*2, 16), 32)

```


### Heatmap of VST-normalized abundance values

```{r, fig.height=figh, fig.width=figw}
# Get VST-normalized matrix from DESeq2 object
vst_normalized_data <- getVarianceStabilizedData(de.seq)

# Convert to data.table format for easier manipulation
vst_normalized_data <- data.table(as.data.frame(vst_normalized_data), keep.rownames = "taxon")

# keep just the actual taxon
vst_normalized_data[,taxon := gsub('.*;', '', taxon)]

# col names
colnames(vst_normalized_data)[-1] <- colData(de.seq)$name


# Convert to matrix format suitable for pheatmap
vst_matrix <- as.matrix(vst_normalized_data[, -1, with = FALSE])
rownames(vst_matrix) <- vst_normalized_data$taxon

# Extract sample information from colData
sample_info <- colData(de.seq)$name
names(sample_info) <- colnames(vst_matrix)

# Define color palette for the heatmap
heatmap_colors <- colorRampPalette(c("#00AFBB", "white", "#FC4E07"))(100)

# Define annotation colors
annotation_colors <- list(Sample = setNames(rainbow(40), unique(sample_info)))

# Define annotation colors using distinct divergent colors
distinct_colors <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b", "#e377c2", "#7f7f7f", "#bcbd22", "#17becf", "#393b79", "#637939", "#8c6d31", "#843c39", "#7b4173", "#5254a3", "#9c9ede", "#8ca252", "#bd9e39", "#ad494a", "#a55194", "#6b6ecf", "#b5cf6b", "#e7ba52", "#d6616b", "#ce6dbd", "#9c755f", "#d9d574", "#aec7e8", "#ffbb78", "#98df8a", "#ff9896", "#c5b0d5", "#c49c94", "#f7b6d2", "#c7c7c7", "#dbdb8d", "#9edae5", "#8c564b", "#ff7f0e")
annotation_colors <- list(Sample = setNames(distinct_colors[1:length(unique(sample_info))], unique(sample_info)))


# Create heatmap with annotation
pheatmap(vst_matrix, #[rownames(vst_matrix) %in% significant_taxa$taxon, ],
         color = heatmap_colors,
         cluster_rows = TRUE,       # Cluster rows (taxa)
         cluster_cols = TRUE,       # Cluster columns (samples)
         show_rownames = TRUE,      # Show row names (taxa)
         show_colnames = TRUE,      # Show column names (samples)
         fontsize_row = 8,          # Font size for row labels
         fontsize_col = 8,          # Font size for column labels
         main = "Heatmap of VST-normalized Taxon Abundances",
         annotation_col = data.frame(Sample = sample_info),
         annotation_colors = annotation_colors,
         fontface_row = "italic")


```



### Significant Differences - Statistics

```{r}
# 0. Number of significantly different taxa
sig.frq <- res.deseq[,.N,by=.(is.significant)]
sig.frq
ggsf <- ggplot(sig.frq, aes(is.significant, N)) +
  geom_col(aes(fill=is.significant)) +
  scale_fill_manual(values = c("FALSE" = "gray", "TRUE" = "red")) +
  labs(title = "Number of Significantly Different Taxa",
       #x = "Signifcantly Different",
       y = "Number of Taxa") +
  theme_minimal() +
  theme(legend.position = "none")

# 1. Volcano Plot: Visualizing log2FoldChange vs. -log10(p-value)
ggv <- ggplot(res.deseq, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point(aes(color = is.significant), alpha = 0.7) +
  scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "red")) +
  labs(title = "Volcano Plot of Differential Abundance",
       x = "Log2 Fold Change",
       y = "-Log10 P-value") +
  theme_minimal() +
  theme(legend.position = "bottom")

# 2. MA Plot: Visualizing average expression vs. log2FoldChange
ggma <- ggplot(res.deseq, aes(x = baseMean, y = log2FoldChange)) +
  geom_point(aes(color = is.significant), alpha = 0.7) +
  scale_x_log10() +
  scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "blue")) +
  labs(title = "MA Plot of Differential Abundance",
       x = "Average Expression (baseMean, log10 scale)",
       y = "Log2 Fold Change") +
  theme_minimal() +
  theme(legend.position = "bottom")

cowplot::plot_grid(ggsf, ggv, ggma, nrow=1, rel_widths = c(0.5,1,1))
```


### Significant Differences - Heatmap of log2FC values

```{r}
# 4. Heatmap of Significant Taxa

ggplot(significant_taxa, aes(x = comparison, y = taxon, fill = log2FoldChange)) +
  geom_tile() +
  scale_fill_gradient2(low = "#00AFBB", mid = "white", high = "#FC4E07", midpoint = 0) +
  scale_fill_viridis(option = "C", name = "log2 Fold Difference") +
  labs(title = "Heatmap of Significant Taxa",
       x = "Comparison",
       y = "Taxon",
       fill = "Log2 Fold Change") +
  theme_minimal() +
  theme(axis.text.y  = element_blank(), #element_text(face = "italic"),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y.left = element_text(angle = 0, face = "italic"),
        panel.spacing = unit(0.1, "lines"),
        strip.placement = "outside",
        ) + 
  facet_nested(rows = vars(phylum, taxon), scales = 'free_y', switch = "y")


```


### Significant Differences - Heatmap of VST-normalized Abundance Values
```{r}
# Get VST-normalized matrix from DESeq2 object
vst_normalized_data <- getVarianceStabilizedData(de.seq)

# Convert to data.table format for easier manipulation
vst_normalized_data <- data.table(as.data.frame(vst_normalized_data), keep.rownames = "taxon")


##
vst_normalized_melt <- melt(vst_normalized_data, variable.name = 'sample', value.name = 'abundance')
vst_normalized_melt <- merge(vst_normalized_melt, sampdat, by='sample')
vst_normalized_melt <- vst_normalized_melt[,.(name, primer, taxon, abundance)]
vst_normalized_melt <- merge(taxtab.DT, vst_normalized_melt, by='taxon')
#vst_normalized_melt <- merge(vst_normalized_melt, group_sampdat,  by='name')

# keep just the actual taxon
vst_normalized_melt[,taxon := gsub('.*;', '', taxon)]


plot_data           <- vst_normalized_melt[taxon %in% significant_taxa$taxon]

```

```{r, fig.width=figw, fig.height=figh}
##
p <- ggplot(plot_data, 
            aes(x = taxon, y = name, fill = abundance)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_gradient2(#low = "#d73027", mid = "white", high = "#4575b4",
                       low = "#FC4E07", mid = "white", high = "#00AFBB",
                       midpoint = 0, # median(melted_dists$distance, na.rm = TRUE), 
                       limit    =  c(-20,20), #range(res.deseq$log2FoldChange, na.rm = TRUE), # c(-3, 3),
                       name="VST-normalized Abundance") +
  scale_fill_viridis(option = "C", name = "VST-normalized Abundance") +
  theme_minimal() +
  coord_flip() +
  theme(axis.text.y  = element_blank(), #element_text(face = "italic"),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y.left = element_text(angle = 0, face = "italic"),
        panel.spacing = unit(0.1, "lines"),
        strip.placement = "outside",
      ) +
  facet_nested(rows=vars(phylum, taxon),
               cols=vars(primer), 
               scales = "free", space = "free", switch = 'y')


print(p)
```


### Significant Differences - Heatmap of VST-normalized Abundance Differences
```{r, fig.width=figw, fig.height=figh, eval=F}
##
vst_normalized_cast <- dcast(vst_normalized_melt, name + taxon ~ primer, value.var = 'abundance')
vst_normalized_cast <- merge(taxtab.DT, vst_normalized_cast, by='taxon')
vst_normalized_cast <- merge(vst_normalized_cast, group_sampdat,  by='name')
vst_normalized_cast <- unique(vst_normalized_cast)
vst_normalized_cast[,diff := ZHMW - ZMB]

plot_data           <- vst_normalized_cast[taxon %in% significant_taxa$taxon]

##
p <- ggplot(plot_data, 
            aes(x = taxon, y = name, fill = diff)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_gradient2(#low = "#d73027", mid = "white", high = "#4575b4",
                       low = "#FC4E07", mid = "white", high = "#00AFBB",
                       midpoint = 0, # median(melted_dists$distance, na.rm = TRUE), 
                       limit    =  c(-20,20), #range(res.deseq$log2FoldChange, na.rm = TRUE), # c(-3, 3),
                       name="VST-normalized Difference") +
  scale_fill_viridis(option = "C", name = "VST-normalized Difference") +
  theme_minimal() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(hjust = 1, face = 'italic'),
        #axis.title  = element_blank(),
        panel.background = element_rect(fill = "gray90", color = NA),
        #strip.text.y = element_blank(),
        #strip.text.y = element_text(face = 'italic'),
        panel.spacing = unit(0.2, 'mm')
        ) +
 facet_nested(cols=vars(sex, source),
              #rows=vars(grp2), 
              scales = "free", space = "free")

print(p)

```



### PERMANOVA

Full-model PERMANOVA: source + dna.isolation.method
```{r, fig.width=12, fig.height=6}
distance <- 'bray'

otutab_t <- data.frame(t(otu_table(ps)))
otutab   <- data.frame((otu_table(ps)))
sampdat  <- data.frame(sample_data(ps))


stopifnot(all(rownames(sampdat) == colnames(otutab)))
stopifnot(all(rownames(sampdat) == rownames(otutab_t)))

ps.dists <- phyloseq::distance(ps, distance, type='samples')

```

```{r}

permanova.full <- adonis2(otutab_t ~ primer, data=sampdat, method=distance, by = 'terms')
permanova.full
permanova.full <- data.frame(group=rownames(permanova.full), permanova.full)  

```

```{r, fig.width=12, fig.height=6}

ggplot(permanova.full, aes(x=group, y=R2)) + 
  geom_bar(stat="identity", fill="skyblue") +
  labs(title="R-squared values from PERMANOVA",
       y="R^2 value",
       x="library") +
  theme_ipsum() +
  theme(axis.text.x = element_text(angle=45, hjust=1)) # +facet_grid(cols=vars(group))
```





```{r}
## save phyloseq object
saveRDS(ps.all, paste0(res.dir, '.rds'))

## save teh whole image
save.image(paste0(res.dir, '.RData'))
```


